### 2014-02-19 reanalysis/ComBat_original/combat_paper_dataset2_reproduced.rmd
start

Reproduction and alternative analysis of data set 2 from Johnson et al
========================================================

Description how to obtain the data were found [here](https://groups.google.com/forum/#!msg/combat-user-forum/S9vBcXw8RGk/QIoD6oRBM9IJ) which had links to [dataExample2.txt](http://www.bu.edu/jlab/wp-assets/ComBat/data/dataExample2.txt)
and [sampleInfoExample2.txt](http://www.bu.edu/jlab/wp-assets/ComBat/data/sampleInfoExample2.txt)

The data consist of 35 samples of which 5 are annotated as "WT" and are not
refered to in the text or the plots. These are taken out before ComBat adjustment.

Read data and sample annotation
-------------------------------------------------------

```{r}
library(pheatmap)
library(RColorBrewer)
library(sva)
library(qvalue)
```
```{r, cache=TRUE}

orgdatamatrix = as.matrix(read.table("data/dataExample2.txt", sep="\t", header=TRUE))
datamatrix=orgdatamatrix[1:3000,]# for dev.
datamatrix=orgdatamatrix# for dev.
sampleannotation = read.table("data/sampleInfoExample2.txt", 
                              sep="\t", header=TRUE, stringsAsFactors=FALSE)
rownames(sampleannotation)=sampleannotation$ArrayName
sampleannotation$Batch=factor(as.character(sampleannotation$Batch)) 
sampleannotation$Cell=factor(as.character(sampleannotation$Cell)) 
# must be discrete for the pheatmap
#table(sampleannotation$ArrayName==dimnames(datamatrix)[[2]]) # ordercheck
datamatrix=datamatrix[,sampleannotation$Type!="WT"]
sampleannotation=sampleannotation[sampleannotation$Type!="WT",]
sampleannotation$Type=factor(sampleannotation$Type)
print(dim(datamatrix))
#dev/debug
useparprior=TRUE
```

Reproduce results
-------------------------------------------------------
First we try to reproduce the heatmap in Figure A.1, which uses 2698 genes with large variation.
```{r, tidy=FALSE}
variatonmeasure = apply(datamatrix, 1, FUN= function(x){var(x)})
#variatonmeasure = apply(datamatrix, 1, FUN= function(x){sd(x)/mean(x)}) #CV
#clustermatrix = datamatrix[order(variatonmeasure, decreasing=TRUE),][1:50,]#[1:2698,]
clustermatrix = datamatrix[order(variatonmeasure, decreasing=TRUE),][1:2698,]
Batchcol = brewer.pal(8,"Set2")[1:3]
names(Batchcol) = levels(sampleannotation$Batch)
Typecol =brewer.pal(8,"Set2")[4:5]
names(Typecol) = levels(sampleannotation$Type)
ann_colors = list(Batch = Batchcol, Type = Typecol)
```

```{r reproduced_figA1, dev='png', fig.width=8, fig.height=15}
pheatmap(clustermatrix, scale = "row", cluster_rows=T, cluster_cols=T, 
         color = colorRampPalette(c("red", "black", "green"))(n = 299),
         annotation = sampleannotation[, c("Batch", "Type")],  annotation_colors = ann_colors,
         cellheight=NA, cellwidth=13,fontsize=12,  drop_levels=TRUE, show_rownames=F,
         main=paste("Reproduced Fig A1.",sep=""),border_color=NA,treeheight_row=0)

```

The Heatmap is not exactly as Fig. A1 in the paper. This could be due to different ways of calculating variation for a probe and many other ways the clustring could have been performed. But the purpose of the figure was to show that a batch effect is present in the data and espeically for batch 3. This is also achived in the reproduced figure.

Next 4 datasets were made,
- EB2: batches 1 and 2 adjusted with nonparametric use of ComBat
- EB3: All thre batches adjusted with nonparametric use of ComBat
- Batch12: batches 1 and 2 unadjusted
- Batch3: Only batch 3 (no adjustments)
```{r,results='hide', tidy=FALSE}
mat = datamatrix[,sampleannotation$Batch %in% c("1","2")]
EB2 = as.matrix( sva::ComBat(
            dat=mat, 
            batch=sampleannotation[colnames(mat),"Batch"], 
            mod=model.matrix( ~as.factor(sampleannotation[colnames(mat),"Type"])  ), 
            numCovs=NULL, 
            par.prior=useparprior, 
            prior.plots=FALSE))

mat = datamatrix[, ]
EB3 = as.matrix( sva::ComBat(
            dat=mat, 
            batch=sampleannotation[colnames(mat),"Batch"], 
            mod=model.matrix(~as.factor(sampleannotation[colnames(mat),"Type"])), 
            numCovs=NULL, 
            par.prior=useparprior, 
            prior.plots=FALSE))
Batch12 = datamatrix[,sampleannotation$Batch %in% c("1","2")]
Batch3 = datamatrix[,sampleannotation$Batch=="3"]
```

Next we try to re-create Figure A.2

> A heatmap diagram of 770 genes from data set 2 after applying the EB batch adjustments.<cite> Johnson et al

There is no description in how the 770 genes were selected so we use the same approach as in the reprodution of Figure A.1

```{r reproduced_figA2, dev='png', fig.width=8, fig.height=15}
clustermatrix = EB3
variatonmeasure = apply(clustermatrix, 1, FUN= function(x){var(x)})
#variatonmeasure = apply(datamatrix, 1, FUN= function(x){sd(x)/mean(x)}) #CV
clustermatrix = clustermatrix[order(variatonmeasure, decreasing=TRUE),][1:770,]
Batchcol = brewer.pal(8,"Set2")[1:3]
names(Batchcol) = levels(sampleannotation$Batch)
Typecol =brewer.pal(8,"Set2")[4:6]
names(Typecol) = levels(sampleannotation$Type)
Cellcol =brewer.pal(8,"Set1")[1:5]
names(Cellcol) = levels(sampleannotation$Cell)
ann_colors = list(Batch = Batchcol, Type = Typecol, Cell=Cellcol)
pheatmap(clustermatrix, scale = "row", cluster_rows=T, cluster_cols=T, 
         color = colorRampPalette(c("red", "black", "green"))(n = 299),
         annotation = sampleannotation[, c("Batch", "Type", "Cell")],  
         cellheight=NA, cellwidth=13,fontsize=12,  drop_levels=TRUE, show_rownames=F,
         main=paste("Reproduced Fig A2.",sep=""),border_color=NA,treeheight_row=0)

```

> Differential expression was assessed using Welch’s t-test to determine the differential expression of RNAi versus control samples. EB2 produced at list of 86 significant genes at a false discovery (q-value) threshold of 0.05 (Storey and Tibshirani, 2003).<cite> Johnson et al

T-test for EB2 and count probes q<0.05.
```{r, tidy=FALSE}
EB2_pvals = apply(EB2, 1 , 
                  FUN=function(x){t.test(
                                    x[sampleannotation[colnames(EB2), "Type"]=="C"],
                                    x[sampleannotation[colnames(EB2), "Type"]=="R"]
                                    )$p.value})
print(table(qvalue(EB2_pvals)$qvalue<0.05))
```

>The third batch alone produced a list of 37 significant genes using the same threshold. Crossing the significant gene lists, we observed 13 genes common in both lists (Fisher’s exact p-value < 1e-15).

```{r, tidy=FALSE}
Batch3_pvals = apply(Batch3, 1 , 
                  FUN=function(x){t.test(
                                    x[sampleannotation[colnames(Batch3), "Type"]=="C"],
                                    x[sampleannotation[colnames(Batch3), "Type"]=="R"]
                                    )$p.value})
print(table(qvalue(Batch3_pvals)$qvalue<0.05))
```

>Without any adjustment, combining these two batches produced a list of only 9 genes a q-value cutoff of 0.05

```{r, tidy=FALSE}
Batch12_pvals = apply(Batch12, 1 , 
                  FUN=function(x){t.test(
                                    x[sampleannotation[colnames(Batch12), "Type"]=="C"],
                                    x[sampleannotation[colnames(Batch12), "Type"]=="R"]
                                    )$p.value})
print(table(qvalue(Batch12_pvals)$qvalue<0.05))
```

>Welch’s t-test was also applied to EB3 to find differential expressed genes; yielding 1599 genes significant at a q-value cutoff of 0.05

```{r, tidy=FALSE}
EB3_pvals = apply(EB3, 1 , 
                  FUN=function(x){t.test(
                                    x[sampleannotation[colnames(EB3), "Type"]=="C"],
                                    x[sampleannotation[colnames(EB3), "Type"]=="R"]
                                    )$p.value})
print(table(qvalue(EB3_pvals)$qvalue<0.05))
```

>Reducing the q-value threshold to 0.01 yielded 488 significant genes
```{r}
print(table(qvalue(EB3_pvals)$qvalue<0.01))
```

>..decreasing the threshold further to 0.001 yielded 161 significant genes.
```{r}
print(table(qvalue(EB3_pvals)$qvalue<0.001))
```
The reproduced number of significant probes are not the same as the reported ones, but they are not completly off.

```{r}
require(limma)
Type = as.factor(sampleannotation$Type)
design = model.matrix(~0 + Type)
colnames(design)=levels(Type)
fit <- lmFit(EB3, design)
cont.matrix = makeContrasts ( contrasts="R-C", levels=design)	
fit2 = contrasts.fit(fit, cont.matrix)
fit2 <- eBayes(fit2)
EB3_limma_pvalues = fit2$p.value[,1]
print(table(qvalue(EB3_limma_pvalues)$qvalue<0.05))
```

Result when including batch in the statistical test
--------------

```{r}
Type = as.factor(sampleannotation$Type)
Block = as.factor(sampleannotation$Batch)
design <- model.matrix(~0+Type+Block)
fit <- lmFit(datamatrix, design)
cont.matrix = makeContrasts ( contrasts="TypeR-TypeC", levels=design)	
fit2 = contrasts.fit(fit, cont.matrix)
fit2 <- eBayes(fit2)
datamatrix_limma_pvalues = fit2$p.value[,1]
print(table(qvalue(datamatrix_limma_pvalues)$qvalue<0.05))
```

```{r, dev='svg', fig.width=8, fig.height=8}
hist(EB3_limma_pvalues, border="blue", main="P-values, ComBat vs Limma", breaks=100, xlab="p-value")
hist(datamatrix_limma_pvalues, border="red", add=T, breaks=100)
legend("topright", legend=c("ComBat adjusted", "Limma adjusted"), text.col=c("blue", "red"))
```
This plot shows that the p-values calculated from the ComBat adjusted data is skewed towards the low end compared to when batch is considered inside the statistical test in limma. However, the difference is not large and from this plot alone it could be argued that ComBat is just better.

Additional sanity checks
---------

```{r, tidy=FALSE}
set.seed(100)
randommatrix = datamatrix
randommatrix[,] =rnorm(length(datamatrix), mean=0, sd=1)
EBrand = as.matrix( sva::ComBat(
            dat=randommatrix, 
            batch=sampleannotation[colnames(randommatrix),"Batch"], 
            mod=model.matrix(~as.factor(sampleannotation[colnames(randommatrix),"Type"])), 
            numCovs=NULL, 
            par.prior=useparprior, 
            prior.plots=FALSE))

#source("../../helper_functions.r")
#randommatrixbatch = addbatcheffect(randommatrix, sampleannotation$Batch, thismean=0, thissd=0)# introduce batch effect

```
```{r, tidy=FALSE}

Type = as.factor(sampleannotation$Type)
design = model.matrix(~0 + Type)
cont.matrix = makeContrasts ( contrasts="TypeR-TypeC", levels=design)

fit <- lmFit(EBrand, design)
fit2 = contrasts.fit(fit, cont.matrix)
fit2 <- eBayes(fit2)
EBrand_limma_pvalues = fit2$p.value[,1]

fit <- lmFit(randommatrix, design)
fit2 = contrasts.fit(fit, cont.matrix)
fit2 <- eBayes(fit2)
randommatrix_limma_pvalues = fit2$p.value[,1]

Block = as.factor(sampleannotation$Batch)
design <- model.matrix(~0+Type+Block)

fit <- lmFit(randommatrix, design)
cont.matrix = makeContrasts ( contrasts="TypeR-TypeC", levels=design) 
fit2 = contrasts.fit(fit, cont.matrix)
fit2 <- eBayes(fit2)
randommatrix_limma_batch_pvalues = fit2$p.value[,1]

```

```{r, dev='svg', fig.width=8, fig.height=8}
hist(EBrand_limma_pvalues, border="blue",  main="P-values, Random numbers", breaks=100, xlab="p-value")
hist(randommatrix_limma_pvalues, border="black", add=T, breaks=100)
hist(randommatrix_limma_batch_pvalues, border="red", add=T, breaks=100)
legend("topright", legend=c("ComBat adjusted", "Limma adjusted", "No adjustment"), text.col=c("blue", "black", "red"))
```
The p-values for ComBat adjusted random numbers are enriched slighlty for low p-values. Indicating that the same enrichment seen for the real data is also false.

Another sanity check is to subset the real data into fictive batches. The EB2 data set contains the same number of samples with the same "Type" balance as the third batch alone. 
>The third batch was used for comparison against the EB2 analysis results because it was an identical experiment to EB2 other than the fact that it was conducted in a single batch.
Thus it is interesting to inspect the p-value from these two indentical experiments, EB2 (batch 1 and 2 adjusted with ComBat) and batch 3 (no ComBat adjustment). These p-values are computed above.
```{r, dev='svg', fig.width=8, fig.height=8}
hist(EB2_pvals, border="blue",  main="P-values, real data", breaks=100, xlab="p-value")
hist(Batch3_pvals, border="red", add=T, breaks=100)
legend("topright", legend=c("ComBat adjusted batch 1 and 2", "Batch 3"), text.col=c("blue", "red"))
```
The 2-batch experiment is able to retrive more significant genes than running all in one batch. A further test of this is to split batch 3 into two imaginary batches in the same design as for batch 1 and 2, and then look at the p-value distribution compared with the real batch 3.

```{r, dev='svg', fig.width=8, fig.height=8, tidy=FALSE}
Batch45 = Batch3
Batch45_annot = sampleannotation[dimnames(Batch45)[[2]],]
Batch45_annot$Batch="4"
Batch45_annot$Batch[Batch45_annot$Type=="C"][1:4] = "5"
Batch45_annot$Batch[Batch45_annot$Type=="R"][1:3] = "5"

table(sampleannotation[sampleannotation$Batch %in% c("1", "2"), c("Batch", "Type")])
table(Batch45_annot[, c("Batch", "Type")])

EB45 = as.matrix( sva::ComBat(
            dat=Batch45, 
            batch=Batch45_annot[colnames(Batch45),"Batch"], 
            mod=model.matrix( ~as.factor(Batch45_annot[colnames(Batch45),"Type"])  ), 
            numCovs=NULL, 
            par.prior=useparprior, 
            prior.plots=FALSE))

EB45_pvals = apply(EB45, 1 , 
                  FUN=function(x){t.test(
                                    x[Batch45_annot[colnames(EB45), "Type"]=="C"],
                                    x[Batch45_annot[colnames(EB45), "Type"]=="R"]
                                    )$p.value})

hist(EB45_pvals, border="blue",  main="P-values, real data", breaks=100, xlab="p-value")
hist(Batch3_pvals, border="red", add=T, breaks=100)
legend("topright", legend=c("ComBat adjusted pseudo-batches of batch 3", "Batch 3 not adjusted"), text.col=c("blue", "red"))
```

Done!

### 2014-02-19 reanalysis/ComBat_original/scripts/ComBat_vegardsmod.R

@@ -1,0 +1,286 @@
+# 'expression_xls' is the expression index file (e.g. outputted by dChip); 'sample_info_file' is a tab-delimited text file containing the colums: Array  name, sample name, Batch, and any other covariates to be included in the modeling; 'type' currently supports two data file types 'txt' for a tab-delimited text file and 'csv' for an Excel .csv file (sometimes R handles the .csv file better, so use this if you have problems with a .txt file!); 'write' if 'T' ComBat writes adjusted data to a file, and if 'F' and ComBat outputs the adjusted data matrix if 'F' (so assign it to an object! i.e. NewData <- ComBat('my expression.xls','Sample info file.txt', write=F)); 'covariates=all' will use all of the columns in your sample info file in the modeling (except array/sample name), if you only want use a some of the columns in your sample info file, specify these columns here as a vector (you must include the Batch column in this list); 'par.prior' if 'T' uses the parametric adjustments, if 'F' uses the nonparametric adjustments--if you are unsure what to use, try the parametric adjustments (they run faster) and check the plots to see if these priors are reasonable; 'filter=value' filters the genes with absent calls in > 1-value of the samples. The defaut here (as well as in dchip) is .8. Filter if you can as the EB adjustments work better after filtering. Filter must be numeric if your expression index file contains presence/absence calls (but you can set it >1 if you don't want to filter any genes) and must be 'F' if your data doesn't have presence/absence calls; 'skip' is the number of columns that contain probe names and gene information, so 'skip=5' implies the first expression values are in column 6; 'prior.plots' if true will give prior plots with black as a kernal estimate of the empirical b
[…]

### 2014-02-19 reanalysis/ComBat_original/combat_paper_dataset2_reproduced.rmd
h

@@
-Reproduction and alternative analysis of data set 2 from Johnson et al
+Reproduction and alternative analysis of "Data Set 2" from "Adjusting batch effects in microarray data using Empirical Bayes methods."
 ========================================================
 
-Description how to obtain the data were found [here](https://groups.google.com/forum/#!msg/combat-user-forum/S9vBcXw8RGk/QIoD6oRBM9IJ) which had links to [dataExample2.txt](http://www.bu.edu/jlab/wp-assets/ComBat/data/dataExample2.txt)
+Overview
+-----
+
+This report aims to show to what extent the use of ComBat led to false results in the second "proof of concept" analysis performed in Johnson et al.
+The example named "Data Set 2" and their analysis are described in the supplementary materials for Johnson et al.
+Description on how to obtain the data were found [here](https://groups.google.com/forum/#!msg/combat-user-forum/S9vBcXw8RGk/QIoD6oRBM9IJ) which had links to [dataExample2.txt](http://www.bu.edu/jlab/wp-assets/ComBat/data/dataExample2.txt)
 and [sampleInfoExample2.txt](http://www.bu.edu/jlab/wp-assets/ComBat/data/sampleInfoExample2.txt)
 
-The data consist of 35 samples of which 5 are annotated as "WT" and are not
-refered to in the text or the plots. These are taken out before ComBat adjustment.
+This document has three main parts
+- Reproduce some of the results to show that we are working on the same data and analysis workflow.
+- Remove the use of ComBat and perform the same analysis with alternative established tools and estimate the error introduced by ComBat.
+- Perform a few more sanity checks to substantiate that the difference in results for the two above analysis is an error introduced by ComBat.
 
@@
 
-```{r}
+The data files consist of 35 samples of which 5 are annotated as "WT" and were not
+refered to in Johnson et al. These are taken out in the beginning and not used in this report.
+
+```{r, results='hide'}
 library(pheatmap)
@@
 library(qvalue)
+library(limma)
 ```
@@
 datamatrix=orgdatamatrix[1:3000,]# for dev.
-datamatrix=orgdatamatrix# for dev.
+#datamatrix=orgdatamatrix# for dev.
 sampleannotation = read.table("data/sampleInfoExample2.txt", 
@@
 ```
+The covariate/batch balance;
+```{r}
+print(table(sampleannotation$Type, sampleannotation$Batch))
+```
 
@@
 -------------------------------------------------------
-First we try to reproduce the heatmap in Figure A.1, which uses 2698 genes with large variation.
+Following the description in section A.1 and A.2 and figure texts in the supplementary materials for Johnson et al. we try to reproduce some of their results, first the heatmap in Figure A.1.
+> A heatmap clustering of data set 2. 698 genes with large variation across all the samples are clustered.<cite> Johnson et al.
+
 ```{r, tidy=FALSE}
 variatonmeasure = apply(datamatrix, 1, FUN= function(x){var(x)})
-#variatonmeasure = apply(datamatrix, 1, FUN= function(x){sd(x)/mean(x)}) #CV
-#clustermatrix = datamatrix[order(variatonmeasure, decreasing=TRUE),][1:50,]#[1:2698,]
 clustermatrix = datamatrix[order(variatonmeasure, decreasing=TRUE),][1:2698,]
@@
 ```
-
-```{r reproduced_figA1, dev='png', fig.width=8, fig.height=15}
+```{r reproduced_figA1, dev='png', fig.width=8, fig.height=15, fig.cap="fig. 1"}
 pheatmap(clustermatrix, scale = "row", cluster_rows=T, cluster_cols=T, 
@@
 ```
-
-The Heatmap is not exactly as Fig.
+The Heatmap does not look exactly as Fig.
 A1 in the paper.
-This could be due to different ways of calculating variation for a probe and many other ways the clustring could have been performed.
+This could be due to undocumented transformations preformed in Johnson et al., for example log-transformation or standarization.
+And there are different ways of calculating variation for a probe.
+What parameters that were used for the clustring could also be different.
 But the purpose of the figure was to show that a batch effect is present in the data and espeically for batch 3.
@@
 
-Next 4 datasets were made,
+Next 3 datasets were made,
 - EB2: batches 1 and 2 adjusted with nonparametric use of ComBat
 - EB3: All thre batches adjusted with nonparametric use of ComBat
-- Batch12: batches 1 and 2 unadjusted
 - Batch3: Only batch 3 (no adjustments)
@@
             prior.plots=FALSE))
-Batch12 = datamatrix[,sampleannotation$Batch %in% c("1","2")]
+
 Batch3 = datamatrix[,sampleannotation$Batch=="3"]
@@
 
-> A heatmap diagram of 770 genes from data set 2 after applying the EB batch adjustments.<cite> Johnson et al
+> A heatmap diagram of 770 genes from data set 2 after applying the EB batch adjustments.<cite> Johnson et al.
 
-There is no description in how the 770 genes were selected so we use the same approach as in the reprodution of Figure A.1
-
+There is no description in how the 770 genes were selected so we use the same approach as in the reprodution of Figure A.1.
 ```{r reproduced_figA2, dev='png', fig.width=8, fig.height=15}
@@
 variatonmeasure = apply(clustermatrix, 1, FUN= function(x){var(x)})
-#variatonmeasure = apply(datamatrix, 1, FUN= function(x){sd(x)/mean(x)}) #CV
 clustermatrix = clustermatrix[order(variatonmeasure, decreasing=TRUE),][1:770,]
@@
 ```
+Again the heatmap is not exactly as in Johnson et al, but the batch clustring is broken and the cell clones with the same number seem to cluster.
 
+Now follows a string of differentially expression test for different data sets.
 > Differential expression was assessed using Welch’s t-test to determine the differential expression of RNAi versus control samples.
-EB2 produced at list of 86 significant genes at a false discovery (q-value) threshold of 0.05 (Storey and Tibshirani, 2003).<cite> Johnson et al
+EB2 produced at list of 86 significant genes at a false discovery (q-value) threshold of 0.05 (Storey and Tibshirani, 2003).<cite> Johnson et al.
 
-T-test for EB2 and count probes q<0.05.
 ```{r, tidy=FALSE}
@@
 
->The third batch alone produced a list of 37 significant genes using the same threshold.
+> The third batch alone produced a list of 37 significant genes using the same threshold.
 Crossing the significant gene lists, we observed 13 genes common in both lists (Fisher’s exact p-value < 1e-15).
+<cite> Johnson et al.
 
@@
 
->Without any adjustment, combining these two batches produced a list of only 9 genes a q-value cutoff of 0.05
+> Without any adjustment, combining these two batches produced a list of only 9 genes a q-value cutoff of 0.05<cite> Johnson et al.
 
 ```{r, tidy=FALSE}
+Batch12 = datamatrix[,sampleannotation$Batch %in% c("1","2")]
 Batch12_pvals = apply(Batch12, 1 , 
@@
 
->Welch’s t-test was also applied to EB3 to find differential expressed genes; yielding 1599 genes significant at a q-value cutoff of 0.05
+> Welch’s t-test was also applied to EB3 to find differential expressed genes; yielding 1599 genes significant at a q-value cutoff of 0.05. <cite> Johnson et al.
 
@@
 
->Reducing the q-value threshold to 0.01 yielded 488 significant genes
+> Reducing the q-value threshold to 0.01 yielded 488 significant genes. <cite> Johnson et al.
+
 ```{r}
@@
 
->..decreasing the threshold further to 0.001 yielded 161 significant genes.
+> ..decreasing the threshold further to 0.001 yielded 161 significant genes.<cite> Johnson et al.
+
 ```{r}
@@
 ```
-The reproduced number of significant probes are not the same as the reported ones, but they are not completly off.
 
+The reproduced number of significant probes are not the same as the reported ones, but they are not completly off. Again this is likly to be due to undocumented steps performed by Johnson et al.
+
+Result when including batch in the statistical test
+--------------
+
+An alternative (and better?) way of handling batch effect is to include it in the model for the statistical test.
+This is possible in several tools and we choose to use the popular limma package (Smyth et al).
+Since this is not the tool used above we will first create the corresponding analysis in Limma for the whole ComBat adjusted data set in order to make a fair comparison.
 ```{r}
-require(limma)
 Type = as.factor(sampleannotation$Type)
 design = model.matrix(~0 + Type)
-colnames(design)=levels(Type)
 fit <- lmFit(EB3, design)
-cont.matrix = makeContrasts ( contrasts="R-C", levels=design)	
+cont.matrix = makeContrasts ( contrasts="TypeR-TypeC", levels=design)  
 fit2 = contrasts.fit(fit, cont.matrix)
@@
 ```
+The number of genes below the significance threshold is fairly similar to the t-test.
 
-Result when including batch in the statistical test
---------------
-
+Next the same test is applied to the data whithout ComBat adjustment, but with the batch included as a blocking factor.
 ```{r}
@@
 ```
+Which gives a somwhat lower number.
 
+Now we can compare the P-values for the two alternative procedures.
 ```{r, dev='svg', fig.width=8, fig.height=8}
@@
 
+Lets assume the aim of this study was to obtain a list of genes that differed between control and treatment with a significant cut-off of q<0.05.
+Then handling the batch effect with ComBat adjustment will yield a list of `r sum(qvalue(EB3_limma_pvalues)$qvalue<0.05)` wheras a list of `r sum(qvalue(datamatrix_limma_pvalues)$qvalue<0.05)` probes would have been closer to the truth.
+
+table( qvalue(datamatrix_limma_pvalues)$qvalue<0.05 & qvalue(EB3_limma_pvalues)$qvalue<0.05)
+
 Additional sanity checks
@@
 
-```{r, tidy=FALSE}
+```{r, results='hide', tidy=FALSE}
 set.seed(100)
@@
             prior.plots=FALSE))
-
-#source("../../helper_functions.r")
-#randommatrixbatch = addbatcheffect(randommatrix, sampleannotation$Batch, thismean=0, thissd=0)# introduce batch effect
-
 ```
@@
 design <- model.matrix(~0+Type+Block)
-
 fit <- lmFit(randommatrix, design)
@@
 
-```{r, dev='svg', fig.width=8, fig.height=8, tidy=FALSE}
+```{r, results='hide', tidy=FALSE}
 Batch45 = Batch3
@@
                                     )$p.value})
-
+```
+```{r, dev='svg', fig.width=8, fig.height=8, tidy=FALSE}
 hist(EB45_pvals, border="blue",  main="P-values, real data", breaks=100, xlab="p-value")

### 2014-02-20 reanalysis/ComBat_original/combat_paper_dataset2_reproduced.rmd
h

@@
 
-This document has three main parts
+This document has four main parts
 - Reproduce some of the results to show that we are working on the same data and analysis workflow.
-- Remove the use of ComBat and perform the same analysis with alternative established tools and estimate the error introduced by ComBat.
+- Remove the use of ComBat and perform the same analysis with alternative established tools
+- Estimate the error introduced by ComBat and its consequences for the conclusion of the study
 - Perform a few more sanity checks to substantiate that the difference in results for the two above analysis is an error introduced by ComBat.
@@
 ```{r}
-print(table(sampleannotation$Type, sampleannotation$Batch))
+print(table(sampleannotation[ , c("Batch", "Type")])  )
 ```
@@
 
-Result when including batch in the statistical test
+Analysis without ComBat
 --------------
@@
 This is possible in several tools and we choose to use the popular limma package (Smyth et al).
-Since this is not the tool used above we will first create the corresponding analysis in Limma for the whole ComBat adjusted data set in order to make a fair comparison.
+
+We start with the data before ComBat adjustment.
+Limma works best with log transformed and between array normalized values, so first we have to do something with the negative signals.
+Then the test is run with the batch included as a blocking factor.
 ```{r}
+thismat = datamatrix
+thismat[thismat<1]=1
+thismat = log2(thismat)
+thismat=normalizeBetweenArrays(thismat)
+Type = as.factor(sampleannotation$Type)
+Block = as.factor(sampleannotation$Batch)
+design <- model.matrix(~0+Type+Block)
+fit <- lmFit(thismat, design)
+cont.matrix = makeContrasts ( contrasts="TypeR-TypeC", levels=design)	
+fit2 = contrasts.fit(fit, cont.matrix)
+datamatrix_limma_res <- eBayes(fit2)
+print(table(qvalue(datamatrix_limma_res$p.value[,1])$qvalue<0.05))
+```
+
+Consequences of ComBat use for the end result
+--------------
+
+Since limma was not the tool used in Johnson et al.s workflow, we will first create the corresponding analysis in Limma for the whole ComBat adjusted data set in order to make a fair comparison.
+```{r}
+thismat = EB3
+thismat[thismat<1]=1
+thismat = log2(thismat)
+thismat=normalizeBetweenArrays(thismat)
 Type = as.factor(sampleannotation$Type)
 design = model.matrix(~0 + Type)
-fit <- lmFit(EB3, design)
+fit <- lmFit(thismat, design)
 cont.matrix = makeContrasts ( contrasts="TypeR-TypeC", levels=design)  
 fit2 = contrasts.fit(fit, cont.matrix)
-fit2 <- eBayes(fit2)
-EB3_limma_pvalues = fit2$p.value[,1]
-print(table(qvalue(EB3_limma_pvalues)$qvalue<0.05))
+EB3_limma_res <- eBayes(fit2)
+print(table(qvalue(EB3_limma_res$p.value[,1])$qvalue<0.05))
 ```
@@
 
-Next the same test is applied to the data whithout ComBat adjustment, but with the batch included as a blocking factor.
-```{r}
-Type = as.factor(sampleannotation$Type)
-Block = as.factor(sampleannotation$Batch)
-design <- model.matrix(~0+Type+Block)
-fit <- lmFit(datamatrix, design)
-cont.matrix = makeContrasts ( contrasts="TypeR-TypeC", levels=design)	
-fit2 = contrasts.fit(fit, cont.matrix)
-fit2 <- eBayes(fit2)
-datamatrix_limma_pvalues = fit2$p.value[,1]
-print(table(qvalue(datamatrix_limma_pvalues)$qvalue<0.05))
-```
-Which gives a somwhat lower number.
-
 Now we can compare the P-values for the two alternative procedures.
 ```{r, dev='svg', fig.width=8, fig.height=8}
-hist(EB3_limma_pvalues, border="blue", main="P-values, ComBat vs Limma", breaks=100, xlab="p-value")
-hist(datamatrix_limma_pvalues, border="red", add=T, breaks=100)
+hist(EB3_limma_res$p.value[,1], border="blue", main="P-values, ComBat vs Limma", breaks=100, xlab="p-value")
+hist(datamatrix_limma_res$p.value[,1], border="red", add=T, breaks=100)
 legend("topright", legend=c("ComBat adjusted", "Limma adjusted"), text.col=c("blue", "red"))
@@
 
-Lets assume the aim of this study was to obtain a list of genes that differed between control and treatment with a significant cut-off of q<0.05.
-Then handling the batch effect with ComBat adjustment will yield a list of `r sum(qvalue(EB3_limma_pvalues)$qvalue<0.05)` wheras a list of `r sum(qvalue(datamatrix_limma_pvalues)$qvalue<0.05)` probes would have been closer to the truth.
-
-table( qvalue(datamatrix_limma_pvalues)$qvalue<0.05 & qvalue(EB3_limma_pvalues)$qvalue<0.05)
+Lets assume one aim of this study was to obtain a list of genes that differed between control and treatment with a significant cut-off of q<0.05.
+Then handling the batch effect with ComBat adjustment will yield a list of `r sum(qvalue(EB3_limma_res$p.value[,1])$qvalue<0.05)` wheras a list of `r sum(qvalue(datamatrix_limma_res$p.value[,1])$qvalue<0.05)` probes would have been closer to the truth.
+Also it is of interest to note that it is mostly the same probes that are found.
+```{r, tidy=FALSE}
+table( qvalue(datamatrix_limma_res$p.value[,1])$qvalue<0.05 ,
+         qvalue(EB3_limma_res$p.value[,1])$qvalue<0.05,
+       dnn=c("limma", "ComBat"))
+```
+And if they were to use the top 1000 probes in a gene set test they found have used many of the same genes regardless of the use of ComBat.
+```{r, tidy=FALSE}
+table( rank(datamatrix_limma_res$p.value[,1])  <=1000 & 
+         rank(EB3_limma_res$p.value[,1])<=1000   )
+```
+Our conclusion is that for this study the error introduced by the use of ComBat would only have a modest effect on the final result.
 
@@
 ---------
+For this data set ComBat's handling of batch-effects yields only modestly different results compared to the alternative.
+Explanation for the difference could be that ComBat is better.
+To substantiate that the result from the use of ComBat is less trustworthy we provide a few additional sanity checks.
 
+First we use random numbers drawn from the same distribution regardless of batch or covariate but retain the batch/covariate design.
 ```{r, results='hide', tidy=FALSE}
@@
 ```
+Limma is then used in 3 ways
+- On the ComBat adjusted random numbers
+- On the random numbers including batch as a blocking factor
+- On the random numbers ignorant to batch information
 ```{r, tidy=FALSE}
@@
 ```
-
+Plotting the p-values
 ```{r, dev='svg', fig.width=8, fig.height=8}

### 2014-02-22 reanalysis/ComBat_original/combat_paper_dataset2_reproduced.rmd

@@
 
-This report aims to show to what extent the use of ComBat led to false results in the second "proof of concept" analysis performed in Johnson et al.
+This <b>report</b> aims to show to what extent the use of ComBat led to false results in the second "proof of concept" analysis performed in Johnson et al.
 The example named "Data Set 2" and their analysis are described in the supplementary materials for Johnson et al.
@@
 ```{r, results='hide'}
-library(pheatmap)
-library(RColorBrewer)
-library(sva)
-library(qvalue)
-library(limma)
+includelibs = c("pheatmap", "sva", "qvalue", "limma")
+lapply(includelibs, require, character.only=T)
 ```
@@
 datamatrix=orgdatamatrix[1:3000,]# for dev.
-#datamatrix=orgdatamatrix# for dev.
 sampleannotation = read.table("data/sampleInfoExample2.txt", 
@@
 
-```{r, tidy=FALSE}
+```{r reproduced_figA1, dev='png', fig.width=8, fig.height=15, fig.cap="fig. 1", tidy=FALSE, results='hold', fig.show='hold'}
 variatonmeasure = apply(datamatrix, 1, FUN= function(x){var(x)})
 clustermatrix = datamatrix[order(variatonmeasure, decreasing=TRUE),][1:2698,]
-Batchcol = brewer.pal(8,"Set2")[1:3]
-names(Batchcol) = levels(sampleannotation$Batch)
-Typecol =brewer.pal(8,"Set2")[4:5]
-names(Typecol) = levels(sampleannotation$Type)
-ann_colors = list(Batch = Batchcol, Type = Typecol)
-```
-```{r reproduced_figA1, dev='png', fig.width=8, fig.height=15, fig.cap="fig. 1"}
 pheatmap(clustermatrix, scale = "row", cluster_rows=T, cluster_cols=T, 
          color = colorRampPalette(c("red", "black", "green"))(n = 299),
-         annotation = sampleannotation[, c("Batch", "Type")],  annotation_colors = ann_colors,
+         annotation = sampleannotation[, c("Batch", "Type")], 
          cellheight=NA, cellwidth=13,fontsize=12,  drop_levels=TRUE, show_rownames=F,
          main=paste("Reproduced Fig A1.",sep=""),border_color=NA,treeheight_row=0)
-
+rm(variatonmeasure, clustermatrix)
 ```
@@
 Batch3 = datamatrix[,sampleannotation$Batch=="3"]
+rm(mat)
 ```
@@
 There is no description in how the 770 genes were selected so we use the same approach as in the reprodution of Figure A.1.
-```{r reproduced_figA2, dev='png', fig.width=8, fig.height=15}
-clustermatrix = EB3
-variatonmeasure = apply(clustermatrix, 1, FUN= function(x){var(x)})
-clustermatrix = clustermatrix[order(variatonmeasure, decreasing=TRUE),][1:770,]
-Batchcol = brewer.pal(8,"Set2")[1:3]
-names(Batchcol) = levels(sampleannotation$Batch)
-Typecol =brewer.pal(8,"Set2")[4:6]
-names(Typecol) = levels(sampleannotation$Type)
-Cellcol =brewer.pal(8,"Set1")[1:5]
-names(Cellcol) = levels(sampleannotation$Cell)
-ann_colors = list(Batch = Batchcol, Type = Typecol, Cell=Cellcol)
+```{r reproduced_figA2, dev='png', fig.width=8, fig.height=15, fig.show='hold'}
+variatonmeasure = apply(EB3, 1, FUN= function(x){var(x)})
+clustermatrix = EB3[order(variatonmeasure, decreasing=TRUE),][1:770,]
 pheatmap(clustermatrix, scale = "row", cluster_rows=T, cluster_cols=T, 
@@
          main=paste("Reproduced Fig A2.",sep=""),border_color=NA,treeheight_row=0)
-
+rm(variatonmeasure, clustermatrix)
 ```
@@
 We start with the data before ComBat adjustment.
-Limma works best with log transformed and between array normalized values, so first we have to do something with the negative signals.
+Limma works best with log transformed and between array normalized values, so first we set negative values to 1 and filter out probes that has this in more than half the samples.
+To ease later comparisons the same filter will also be applied to the EB3 data set.
 Then the test is run with the batch included as a blocking factor.
 ```{r}
-thismat = datamatrix
+negativeprobesfilter =( rowSums(datamatrix>=1) >= (0.5*ncol(datamatrix)) ) & 
+                      ( rowSums(EB3>=1) >= (0.5*ncol(EB3)) )
+thismat = datamatrix[negativeprobesfilter, ]
 thismat[thismat<1]=1
-thismat = log2(thismat)
-thismat=normalizeBetweenArrays(thismat)
+thismat=normalizeBetweenArrays(log2(thismat))
 Type = as.factor(sampleannotation$Type)
 Block = as.factor(sampleannotation$Batch)
-design <- model.matrix(~0+Type+Block)
-fit <- lmFit(thismat, design)
+design = model.matrix(~0+Type+Block)
+fit = lmFit(thismat, design)
 cont.matrix = makeContrasts ( contrasts="TypeR-TypeC", levels=design)	
 fit2 = contrasts.fit(fit, cont.matrix)
-datamatrix_limma_res <- eBayes(fit2)
+datamatrix_limma_res = eBayes(fit2)
+rm(thismat, Type, Block, design, fit, cont.matrix, fit2)
 print(table(qvalue(datamatrix_limma_res$p.value[,1])$qvalue<0.05))
@@
 
-Since limma was not the tool used in Johnson et al.s workflow, we will first create the corresponding analysis in Limma for the whole ComBat adjusted data set in order to make a fair comparison.
+Since limma was not the tool used in Johnson et al.s workflow, we will first need to create the corresponding analysis in Limma for the whole ComBat adjusted data set in order to make a fair comparison.
 ```{r}
-thismat = EB3
+thismat = EB3[negativeprobesfilter, ]
 thismat[thismat<1]=1
-thismat = log2(thismat)
-thismat=normalizeBetweenArrays(thismat)
+thismat=normalizeBetweenArrays(log2(thismat))
 Type = as.factor(sampleannotation$Type)
+Block = as.factor(sampleannotation$Batch)
 design = model.matrix(~0 + Type)
-fit <- lmFit(thismat, design)
+fit = lmFit(thismat, design)
 cont.matrix = makeContrasts ( contrasts="TypeR-TypeC", levels=design)  
 fit2 = contrasts.fit(fit, cont.matrix)
-EB3_limma_res <- eBayes(fit2)
+EB3_limma_res = eBayes(fit2)
+rm(thismat, Type, Block, design, fit, cont.matrix, fit2)
 print(table(qvalue(EB3_limma_res$p.value[,1])$qvalue<0.05))
+
 ```
@@
 ```
-And if they were to use the top 1000 probes in a gene set test they found have used many of the same genes regardless of the use of ComBat.
+And if they were to use the top 1000 probes in a gene set test they found many of the same genes regardless of the use of ComBat.
 ```{r, tidy=FALSE}
@@
 ---------
-For this data set ComBat's handling of batch-effects yields only modestly different results compared to the alternative.
-Explanation for the difference could be that ComBat is better.
-To substantiate that the result from the use of ComBat is less trustworthy we provide a few additional sanity checks.
+To substantiate that the result from the use of ComBat is less trustworthy than the alternative analysis we provide a few additional sanity checks.
 
@@
 
-fit <- lmFit(EBrand, design)
+fit = lmFit(EBrand, design)
 fit2 = contrasts.fit(fit, cont.matrix)
-fit2 <- eBayes(fit2)
-EBrand_limma_pvalues = fit2$p.value[,1]
+EBrand_limma_pvalues = eBayes(fit2)$p.value[,1]
 
-fit <- lmFit(randommatrix, design)
+fit = lmFit(randommatrix, design)
 fit2 = contrasts.fit(fit, cont.matrix)
-fit2 <- eBayes(fit2)
-randommatrix_limma_pvalues = fit2$p.value[,1]
+randommatrix_limma_pvalues = eBayes(fit2)$p.value[,1]
 
 Block = as.factor(sampleannotation$Batch)
-design <- model.matrix(~0+Type+Block)
-fit <- lmFit(randommatrix, design)
+design = model.matrix(~0+Type+Block)
+fit = lmFit(randommatrix, design)
 cont.matrix = makeContrasts ( contrasts="TypeR-TypeC", levels=design) 
 fit2 = contrasts.fit(fit, cont.matrix)
-fit2 <- eBayes(fit2)
+fit2 = eBayes(fit2)
 randommatrix_limma_batch_pvalues = fit2$p.value[,1]
@@
 ```
-The p-values for ComBat adjusted random numbers are enriched slighlty for low p-values. Indicating that the same enrichment seen for the real data is also false.
+The p-values for ComBat adjusted random numbers are slighlty enriched for low p-values. Indicating that the same enrichment seen for the real data is also false.
 
@@
 
-Done!
+References
+-------------
+
+```{r, echo=FALSE, tidy=FALSE,  warning=FALSE, comment="", prompt=FALSE}
+#x = lapply(includelibs, FUN=citation)
+for(x in lapply(includelibs, FUN=citation))
+{
+  print(x, style="textVersion")
+  print("", quote = FALSE,include.rownames=FALSE)
+}
+
+```
+
+```{r sessionInfo, echo=FALSE, comment=""}
+sessionInfo()
+```

### 2014-02-22 reanalysis/ComBat_original/combat_paper_dataset2_reproduced.rmd
g

@@
 This <b>report</b> aims to show to what extent the use of ComBat led to false results in the second "proof of concept" analysis performed in Johnson et al.
-The example named "Data Set 2" and their analysis are described in the supplementary materials for Johnson et al.
+The example named "Data Set 2" and their analysis are described in the [supplementary material](http://biostatistics.oxfordjournals.org/content/suppl/2006/04/21/kxj037.DC1/kxj037supp.pdf) for [Johnson et al.](http://biostatistics.oxfordjournals.org/content/8/1/118.abstract)
 Description on how to obtain the data were found [here](https://groups.google.com/forum/#!msg/combat-user-forum/S9vBcXw8RGk/QIoD6oRBM9IJ) which had links to [dataExample2.txt](http://www.bu.edu/jlab/wp-assets/ComBat/data/dataExample2.txt)
@@
 
+Johnson, WE, Rabinovic, A, and Li, C (2007). Adjusting batch effects in microarray expression data using Empirical Bayes methods. Biostatistics 8(1):118-127.
+
 ```{r, echo=FALSE, tidy=FALSE,  warning=FALSE, comment="", prompt=FALSE}

### 2014-02-24 reanalysis/ComBat_original/combat_paper_dataset2_reproduced.rmd
tt

--- 2014-02-22 reanalysis/ComBat_original/combat_paper_dataset2_reproduced.rmd
@@
 
-Overview
------
+```{r, results='hide', echo=FALSE}
+starttime = Sys.time()
+```
+`r as.character(starttime)`
 
-This <b>report</b> aims to show to what extent the use of ComBat led to false results in the second "proof of concept" analysis performed in Johnson et al.
-The example named "Data Set 2" and their analysis are described in the supplementary materials for Johnson et al.
+### Overview
+This report aims to show to what extent the use of ComBat led to false results in the second analysis example given in [Johnson et al.](http://biostatistics.oxfordjournals.org/content/8/1/118.abstract) The example named "Data Set 2" and the analysis are described in the [supplementary material](http://biostatistics.oxfordjournals.org/content/suppl/2006/04/21/kxj037.DC1/kxj037supp.pdf) for Johnson et al.
 Description on how to obtain the data were found [here](https://groups.google.com/forum/#!msg/combat-user-forum/S9vBcXw8RGk/QIoD6oRBM9IJ) which had links to [dataExample2.txt](http://www.bu.edu/jlab/wp-assets/ComBat/data/dataExample2.txt)
@@
 This document has four main parts
-- Reproduce some of the results to show that we are working on the same data and analysis workflow.
-- Remove the use of ComBat and perform the same analysis with alternative established tools
-- Estimate the error introduced by ComBat and its consequences for the conclusion of the study
-- Perform a few more sanity checks to substantiate that the difference in results for the two above analysis is an error introduced by ComBat.
+- Reproduce some of the results to show that we are working on the same data and analysis workflow
+- Remove the use of ComBat and perform the same analysis with an alternative established tool
+- Estimate the error introduced by ComBat and the consequences for the conclusion of the study
+- Perform a few more sanity checks to substantiate that the difference in results for the two above analysis is mainly due to error introduced by ComBat
 
-Read data and sample annotation
--------------------------------------------------------
+### Read data and sample annotation
+The data files consist of 35 samples of which 5 are annotated as "WT" and were not
+referred to in Johnson et al. These are taken out in the beginning and are not used in this report.
 
-The data files consist of 35 samples of which 5 are annotated as "WT" and were not
-refered to in Johnson et al. These are taken out in the beginning and not used in this report.
-
-```{r, results='hide'}
+```{r, results='hide',  message=FALSE}
 includelibs = c("pheatmap", "sva", "qvalue", "limma")
@@
 ```{r, cache=TRUE}
-
-orgdatamatrix = as.matrix(read.table("data/dataExample2.txt", sep="\t", header=TRUE))
-datamatrix=orgdatamatrix[1:3000,]# for dev.
+datamatrix = as.matrix(read.table("data/dataExample2.txt", sep="\t", header=TRUE))#[1:3000,]# for dev.
 sampleannotation = read.table("data/sampleInfoExample2.txt", 
@@
 sampleannotation$Type=factor(sampleannotation$Type)
-print(dim(datamatrix))
 #dev/debug
 useparprior=TRUE
+print(dim(datamatrix))
+
 ```
-The covariate/batch balance;
+Inspection of the covariate/batch balance;
 ```{r}
@@
 
-Reproduce results
--------------------------------------------------------
-Following the description in section A.1 and A.2 and figure texts in the supplementary materials for Johnson et al. we try to reproduce some of their results, first the heatmap in Figure A.1.
+### Reproduce the original results
+Following the description in section A.1 and A.2 and figure texts in the supplementary materials for Johnson et al., we try to reproduce some of their results. First, the heatmap in Figure A.1.
 > A heatmap clustering of data set 2. 698 genes with large variation across all the samples are clustered.<cite> Johnson et al.
@@
 ```{r reproduced_figA1, dev='png', fig.width=8, fig.height=15, fig.cap="fig. 1", tidy=FALSE, results='hold', fig.show='hold'}
-variatonmeasure = apply(datamatrix, 1, FUN= function(x){var(x)})
-clustermatrix = datamatrix[order(variatonmeasure, decreasing=TRUE),][1:2698,]
+variationmeasure = apply(datamatrix, 1, FUN= function(x){var(x)})
+clustermatrix = datamatrix[order(variationmeasure, decreasing=TRUE),][1:2698,]
 pheatmap(clustermatrix, scale = "row", cluster_rows=T, cluster_cols=T, 
@@
          main=paste("Reproduced Fig A1.",sep=""),border_color=NA,treeheight_row=0)
-rm(variatonmeasure, clustermatrix)
+rm(variationmeasure, clustermatrix)
 ```
-The Heatmap does not look exactly as Fig.
+The heatmap does not look exactly as Fig.
 A1 in the paper.
-This could be due to undocumented transformations preformed in Johnson et al., for example log-transformation or standarization.
+This could be due to undocumented transformations preformed in Johnson et al., for example log-transformation or standardization.
 And there are different ways of calculating variation for a probe.
-What parameters that were used for the clustring could also be different.
-But the purpose of the figure was to show that a batch effect is present in the data and espeically for batch 3.
-This is also achived in the reproduced figure.
+What parameters that were used for the clustering could also be different.
+But the purpose of the figure was to show that a batch effect is present in the data and especially for batch 3.
+This is also achieved in the reproduced figure.
 
-Next 3 datasets were made,
-- EB2: batches 1 and 2 adjusted with nonparametric use of ComBat
-- EB3: All thre batches adjusted with nonparametric use of ComBat
-- Batch3: Only batch 3 (no adjustments)
+Next 3 data sets were made,
+- **EB2**: Batches 1 and 2 adjusted with non-parametric use of ComBat
+- **EB3**: All thee batches adjusted with non-parametric use of ComBat
+- **Batch3**: Only batch 3 (no adjustments)
 ```{r,results='hide', tidy=FALSE}
@@
             prior.plots=FALSE))
-
-mat = datamatrix[, ]
+rm(mat)
 EB3 = as.matrix( sva::ComBat(
-            dat=mat, 
-            batch=sampleannotation[colnames(mat),"Batch"], 
-            mod=model.matrix(~as.factor(sampleannotation[colnames(mat),"Type"])), 
+            dat=datamatrix, 
+            batch=sampleannotation[colnames(datamatrix),"Batch"], 
+            mod=model.matrix(~as.factor(sampleannotation[colnames(datamatrix),"Type"])), 
             numCovs=NULL, 
@@
             prior.plots=FALSE))
+Batch3 = datamatrix[,sampleannotation$Batch=="3"]
 
-Batch3 = datamatrix[,sampleannotation$Batch=="3"]
-rm(mat)
 ```
@@
 
-There is no description in how the 770 genes were selected so we use the same approach as in the reprodution of Figure A.1.
-```{r reproduced_figA2, dev='png', fig.width=8, fig.height=15, fig.show='hold'}
-variatonmeasure = apply(EB3, 1, FUN= function(x){var(x)})
-clustermatrix = EB3[order(variatonmeasure, decreasing=TRUE),][1:770,]
+There is no description in how the 770 genes were selected so we use the same approach as in the reproduction of Figure A.1.
+```{r reproduced_figA2, dev='png', fig.width=8, fig.height=15, fig.show='hold', tidy=FALSE}
+variationmeasure = apply(EB3, 1, FUN= function(x){var(x)})
+clustermatrix = EB3[order(variationmeasure, decreasing=TRUE),][1:770,]
 pheatmap(clustermatrix, scale = "row", cluster_rows=T, cluster_cols=T, 
@@
          main=paste("Reproduced Fig A2.",sep=""),border_color=NA,treeheight_row=0)
-rm(variatonmeasure, clustermatrix)
+rm(variationmeasure, clustermatrix)
 ```
-Again the heatmap is not exactly as in Johnson et al, but the batch clustring is broken and the cell clones with the same number seem to cluster.
+Again the heatmap is not exactly as in Johnson et al, but the batch clustering is broken("Batch, bottom annotation row") and the samples cluster more by cell type("Cell", top annotation row) and treatment type("Type", middle annotation row).
 
-Now follows a string of differentially expression test for different data sets.
+Now follows a few tests for differentially expressed probes.
 > Differential expression was assessed using Welch’s t-test to determine the differential expression of RNAi versus control samples.
@@
 ```
+Original number was **86**, reproduced number is **`r sum(qvalue(EB2_pvals)$qvalue<0.05)`**.
 
-> The third batch alone produced a list of 37 significant genes using the same threshold.
-Crossing the significant gene lists, we observed 13 genes common in both lists (Fisher’s exact p-value < 1e-15).
-<cite> Johnson et al.
+> The third batch alone produced a list of 37 significant genes using the same threshold. <cite> Johnson et al.
 
@@
 ```
+Original number was **37**, reproduced number is **`r sum(qvalue(Batch3_pvals)$qvalue<0.05)`**.
 
@@
 ```
+Original number was **9**, reproduced number is **`r sum(qvalue(Batch12_pvals)$qvalue<0.05)`**.
 
@@
 ```
+Original number was **1599**, reproduced number is **`r sum(qvalue(EB3_pvals)$qvalue<0.05)`**.
 
@@
 ```
+Original number was **488**, reproduced number is **`r sum(qvalue(EB3_pvals)$qvalue<0.01)`**.
 
@@
 ```
+Original number was **161**, reproduced number is **`r sum(qvalue(EB3_pvals)$qvalue<0.001)`**.
 
-The reproduced number of significant probes are not the same as the reported ones, but they are not completly off. Again this is likly to be due to undocumented steps performed by Johnson et al.
+The reproduced numbers are not the same as the reported ones, but they are not completely off. Again this is likely to be due to undocumented steps performed by Johnson et al.
 
-Analysis without ComBat
---------------
-
+### Analysis without ComBat
 An alternative (and better?) way of handling batch effect is to include it in the model for the statistical test.
@@
 ```
+Number of differentially expressed probes found with the alternative anaysis is **`r sum(qvalue(datamatrix_limma_res$p.value[,1])$qvalue<0.05)`**.
 
-Consequences of ComBat use for the end result
---------------
+### Consequences of ComBat use for the end result
 
-Since limma was not the tool used in Johnson et al.s workflow, we will first need to create the corresponding analysis in Limma for the whole ComBat adjusted data set in order to make a fair comparison.
+Since limma was not the tool used in Johnson et al.'s workflow, we will first need to create the corresponding analysis in Limma for the whole ComBat adjusted data set in order to make a fair comparison.
 ```{r}
@@
 Lets assume one aim of this study was to obtain a list of genes that differed between control and treatment with a significant cut-off of q<0.05.
-Then handling the batch effect with ComBat adjustment will yield a list of `r sum(qvalue(EB3_limma_res$p.value[,1])$qvalue<0.05)` wheras a list of `r sum(qvalue(datamatrix_limma_res$p.value[,1])$qvalue<0.05)` probes would have been closer to the truth.
+Then handling the batch effect with ComBat adjustment will yield a list of **`r sum(qvalue(EB3_limma_res$p.value[,1])$qvalue<0.05)`** whereas a list of **`r sum(qvalue(datamatrix_limma_res$p.value[,1])$qvalue<0.05)`** probes would have been closer to the truth.
 Also it is of interest to note that it is mostly the same probes that are found.
@@
 ```
-And if they were to use the top 1000 probes in a gene set test they found many of the same genes regardless of the use of ComBat.
+And if they were to use the top 1000 probes in a gene set test, they would have found many of the same genes regardless of hadling batch effects by ComBat or limma.
 ```{r, tidy=FALSE}
@@
 ```
-Our conclusion is that for this study the error introduced by the use of ComBat would only have a modest effect on the final result.
+Our conclusion is that for this study the error introduced by the use of ComBat would probably have a modest effect on the final result.
 
@@
 
-First we use random numbers drawn from the same distribution regardless of batch or covariate but retain the batch/covariate design.
+First we use random numbers drawn from the same distribution regardless of batch or covariate but retaining the batch/covariate design.
 ```{r, results='hide', tidy=FALSE}
@@
 - On the random numbers including batch as a blocking factor
-- On the random numbers ignorant to batch information
+- On the random numbers ignoring batch information
 ```{r, tidy=FALSE}
@@
 ```
-The p-values for ComBat adjusted random numbers are slighlty enriched for low p-values. Indicating that the same enrichment seen for the real data is also false.
+The p-values for ComBat adjusted random numbers are slightly enriched for low p-values. Indicating that the same enrichment seen in the real data is also false.
 
-Another sanity check is to subset the real data into fictive batches. The EB2 data set contains the same number of samples with the same "Type" balance as the third batch alone. 
->The third batch was used for comparison against the EB2 analysis results because it was an identical experiment to EB2 other than the fact that it was conducted in a single batch.
-Thus it is interesting to inspect the p-value from these two indentical experiments, EB2 (batch 1 and 2 adjusted with ComBat) and batch 3 (no ComBat adjustment). These p-values are computed above.
+Another sanity check is to subset the real data into fictive batches.
+The idea behind this check stems from the observation that the EB2 data set contains the same number of samples with the same batch/covariate balance as the third batch alone.
+> The third batch was used for comparison against the EB2 analysis results because it was an identical experiment to EB2 other than the fact that it was conducted in a single batch.
+<cite> Johnson et al.
+
+Thus it is interesting to inspect the p-value from these two identical experiments, EB2 (batch 1 and 2 adjusted with ComBat) and batch 3 (no ComBat adjustment).
+These p-values are already computed above.
 ```{r, dev='svg', fig.width=8, fig.height=8}
@@
 ```
-The 2-batch experiment is able to retrive more significant genes than running all in one batch.
-A further test of this is to split batch 3 into two imaginary batches in the same design as for batch 1 and 2, and then look at the p-value distribution compared with the real batch 3.
+The 2-batch experiment is able to retrieve more genes as significant than running all in one batch for the same sample size.
+This is also observed by Johnson et al., but not commented.
+We claim that the apparent better result for the two-batch experiment is likely a consequence of the use of ComBat.
+A test of this is to split batch 3 into two imaginary batches with the same design as for batch 1 and 2, and then look at the p-value distribution compared with the real batch 3.
 
-```{r, results='hide', tidy=FALSE}
+```{r, tidy=FALSE}
 Batch45 = Batch3
@@
 Batch45_annot$Batch[Batch45_annot$Type=="R"][1:3] = "5"
-
-table(sampleannotation[sampleannotation$Batch %in% c("1", "2"), c("Batch", "Type")])
-table(Batch45_annot[, c("Batch", "Type")])
-
+print(table(sampleannotation[sampleannotation$Batch %in% c("1", "2"), c("Batch", "Type")]))
+print(table(Batch45_annot[, c("Batch", "Type")]))
+```
+Then adjust the data with fictive batches and calculate p-values as done for the unadjusted batch 3.
+```{r, results='hide', tidy=FALSE}
 EB45 = as.matrix( sva::ComBat(
@@
 hist(Batch3_pvals, border="red", add=T, breaks=100)
-legend("topright", legend=c("ComBat adjusted pseudo-batches of batch 3", "Batch 3 not adjusted"), text.col=c("blue", "red"))
+legend("topright", legend=c("ComBat adjusted fictive-batches of batch 3", "Batch 3 not adjusted"), text.col=c("blue", "red"))
+```
+The fictive-batches version with ComBat seemingly outperforms the original.
+This plot looks very similar to the EB2 vs.
+Batch3 p-values, and for both, the p-values for the ComBat adjusted data are likely deflated due to ComBat use.
+
+The fictive batch asssignment above was just one of several possible. Repeating the selection 10 times shows that the lower p-values was not a coincidence.
+
+```{r, tidy=FALSE, results='hide'}
+fictivebatches_pvalcounts=list()
+runs = 10
+for(i in 1:runs)
+{
+Batch45 = Batch3
+Batch45_annot = sampleannotation[dimnames(Batch45)[[2]],]
+Batch45_annot$Batch="4"
+Batch45_annot$Batch[Batch45_annot$Type=="C"][sample(1:sum(Batch45_annot$Type=="C"), 4)] = "5"
+Batch45_annot$Batch[Batch45_annot$Type=="R"][sample(1:sum(Batch45_annot$Type=="R"), 3)] = "5"
+print(table(sampleannotation[sampleannotation$Batch %in% c("1", "2"), c("Batch", "Type")]))
+print(table(Batch45_annot[, c("Batch", "Type")]))
+
+EB45 = as.matrix( sva::ComBat(
+            dat=Batch45, 
+            batch=Batch45_annot[colnames(Batch45),"Batch"], 
+            mod=model.matrix( ~as.factor(Batch45_annot[colnames(Batch45),"Type"])  ), 
+            numCovs=NULL, 
+            par.prior=TRUE, 
+            prior.plots=FALSE))
+
+pvals = apply(EB45, 1 , 
+                  FUN=function(x){t.test(
+                                    x[Batch45_annot[colnames(EB45), "Type"]=="C"],
+                                    x[Batch45_annot[colnames(EB45), "Type"]=="R"]
+                                    )$p.value})
+
+fictivebatches_pvalcounts[[i]] = hist(pvals, plot=FALSE, breaks=100)$counts
+}
+```
+In order to save some computational time the parameteric version of ComBat was used for this.
+```{r, dev='svg', fig.width=8, fig.height=8, tidy=FALSE}
+plot((1:20)/100, hist(Batch3_pvals, plot=FALSE, breaks=100)$counts[1:20], col="red",  main="P-values, fictive batches", 
+      xlab="p-value", ylab="Frequency", type="l", lwd=2, 
+      ylim=c(0, max(unlist(fictivebatches_pvalcounts))))
+for(i in 1:length(fictivebatches_pvalcounts))
+{
+  lines((1:20)/100,fictivebatches_pvalcounts[[i]][1:20], col="blue")
+}
+legend("topright", legend=c("ComBat adjusted fictive-batches of batch 3", "Batch 3 not adjusted"), text.col=c("blue", "red"))
 ```
 
-References
--------------
+### References
 
-```{r, echo=FALSE, tidy=FALSE,  warning=FALSE, comment="", prompt=FALSE}
-#x = lapply(includelibs, FUN=citation)
-for(x in lapply(includelibs, FUN=citation))
-{
-  print(x, style="textVersion")
-  print("", quote = FALSE,include.rownames=FALSE)
-}
+Johnson, WE, Rabinovic, A, and Li, C (2007). Adjusting batch effects in microarray expression data using Empirical Bayes methods. Biostatistics 8(1):118-127.
 
+Storey, J. D. and Tibshirani, R. (2003) Proc Natl Acad Sci U S A, 100, 9440-5.
+
+Raivo Kolde (2013). pheatmap: Pretty Heatmaps. R package version 0.7.7. http://CRAN.R-project.org/package=pheatmap
+
+Leek JT, Johnson WE, Parker HS, Jaffe AE, Storey JD.(2012) The sva package for removing batch effects and other unwanted variation in high-throughput experiments.
+Bioinformatics.
+2012 Mar 15;28(6):882-3.
+
+Smyth, GK (2005). Limma: linear models for microarray data. In: 'Bioinformatics and Computational Biology Solutions
+  using R and Bioconductor'. R. Gentleman, V. Carey, S. Dudoit, R. Irizarry, W. Huber (eds), Springer, New York, pages
+  397-420.
+
+  R Core Team (2013). R: A language and environment for statistical computing. R Foundation for Statistical Computing,
+  Vienna, Austria. URL http://www.R-project.org/
+
+  Yihui Xie (2013). knitr: A general-purpose package for dynamic report generation in R. R package version 1.5.
+
+  Yihui Xie (2013) Dynamic Documents with R and knitr. Chapman and Hall/CRC. ISBN 978-1482203530
+
+  Yihui Xie (2013) knitr: A Comprehensive Tool for Reproducible Research in R. In Victoria Stodden, Friedrich Leisch and
+  Roger D. Peng, editors, Implementing Reproducible Computational Research. Chapman and Hall/CRC. ISBN 978-1466561595
+
+  RStudio Team (2012). RStudio: Integrated Development for R. RStudio, Inc., Boston, MA URL http://www.rstudio.com/.
+
+```{r sessionInfo, comment=""}
+sessionInfo()
 ```
 
-```{r sessionInfo, echo=FALSE, comment=""}
-sessionInfo()
-```
+generation ended `r as.character(Sys.time())`. Time spent `r  as.integer(round(difftime(Sys.time(),starttime, units="mins")))` minutes .
+
+table(sampleannotation[ sampleannotation$Batch!="3", c("Batch", "Cell")]  )
+table(Batch45_annot[, c("Batch", "Cell")])
+table(sampleannotation[, c("Batch", "Cell")])

### 2014-02-24 reanalysis/Fadi_Towfic_generic_medicine/towfic_et_at.rmd
tt

Reproduction and alternative analysis of "Comparing the Biological Impact of Glatiramer Acetate with the Biological Impact of a Generic"
========================================================

```{r, results='hide', echo=FALSE}
starttime = Sys.time()
```
`r as.character(starttime)`

### Overview
This report aims to show that the use of the statistical tool ComBat from [Johnson et al.](http://biostatistics.oxfordjournals.org/content/8/1/118.abstract) led to false results in the analysis performed in [Towfic et al.'s ](http://www.plosone.org/article/info%3Adoi%2F10.1371%2Fjournal.pone.0083757) "Comparing the Biological Impact of Glatiramer Acetate with the Biological Impact of a Generic".

This document has five main parts
- Getting and formatting the data
- Reproduce some of the results to show that we are working on the same data and analysis workflow
- Remove the use of ComBat and perform an similar analysis with alternative established tools
- Estimate the error introduced by ComBat and the consequences for the conclusion of the study
- Perform a few more sanity checks to substantiate that the difference in results for the two above analysis is mainly due to error introduced by ComBat

### Getting the data

```{r, results='hide', message=FALSE}
includelibs = c("Biobase", "GEOquery", "sva", "limma",
                 "preprocessCore")
lapply(includelibs, require, character.only=T)
```
The important sample information is described in Table S1 from Towfic et al. the its usage in ComBat is described briefly in the methods;
>  Each microarray’s chip designation was supplied a batch label; there were 18 batches in all. The labels for the treatments (i.e. drug product, reference standard…) were added as covariates.<cite> Towfic et al.

Table S1 does have a "Chip"- coloumn, unfortunatly there is no dedicated "treatment"-coloumn.
Communication with the corresponding author yielded this explanation
> ..and the only thing we used as a covariate was the unique treatment names themselves (e.g. ³Medium² or ³RS").

Based on these descriptions and the annotation for the GEO deposit, we compiled a more complete sample annotation file.

```{r}
sampleannotation = read.table("data/sampleannotation.csv", sep="\t", header=TRUE,  stringsAsFactors=FALSE)
sampleannotation$code = make.names(sampleannotation$code)
sampleannotation$chip = as.character(sampleannotation$chip)
dimnames(sampleannotation)[[1]] = sampleannotation$code
head(sampleannotation)
```
The "covariate"-coloumn is made based on the "code"-coloumn, but might not match  to what were actually used. The last two coloumns are from the GEO deposit and reveal that 3 of the samples do not have data and are taken out.

```{r}
# take out 3 samples that are not assign to a geoaccession. Failed QC?
sampleannotation = sampleannotation[!is.na(sampleannotation$geoaccession),] 
```
The batch/covariate design shows many batches and covariate groups.
```{r}
# the batch/covariate design
table(sampleannotation[, c("chip", "covariate")])
```
A look at the most interesting groups, "DP"(drug product) and "N"(generic) reveals a lack of balance.
```{r}
table(sampleannotation[sampleannotation$covariate %in% c("DP", "N"), c("chip", "covariate")])
```

> The microarray data have been deposited in the Gene Expression Omnibus, under accession number [GSE40566](http://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE40566)

The processed data from the above GEO deposit was deposited for another publication which did not processes the data as described in Tawfic et al. (personal communication). The raw data linked to as [GSE40566_RAW.tar](http://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE40566&format=file) and [GSE40566_non_normalized.txt.gz](http://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE40566&format=file&file=GSE40566%5Fnon%5Fnormalized%2Etxt%2Egz) were thus used, although it is not confirmed that this is the version of the data actually used in Towfic et al.
```{r}
# the output from Illumina software.
rawdata = read.table("data/GSE40566_non_normalized.txt", sep="\t", header=TRUE, stringsAsFactors=FALSE)

# the probe annotation file found inside GSE40566_RAW.tar
rawannotation = read.table("data/GPL6887_MouseWG-6_V2_0_R3_11278593_A.txt", sep="\t", header=TRUE, stringsAsFactors=FALSE, skip=8, comment.char="", quote="", fill=TRUE)
```

Then some formatting is needed to match the probe information to the measurments and the measurments to the sample information.
```{r}
# annotation file has two separate tables, one for experimental and one at the end for controls.
# splitting and fixing the tables:
tmp =  rawannotation$Species=="Mus musculus"
experimentalannot = rawannotation[tmp,]
experimentalannot$Array_Address_Id = as.numeric(experimentalannot$Array_Address_Id)
controlannot = rawannotation[!tmp,]
dimnames(controlannot)[[2]] = rawannotation[rawannotation[,2]=="Array_Address_Id",]
controlannot$Array_Address_Id = suppressWarnings(as.numeric(controlannot$Array_Address_Id))
controlannot = controlannot[!is.na(controlannot$Array_Address_Id),]
controlannot=controlannot[,1:6]
probeannotation = merge(experimentalannot, controlannot, all=TRUE )
#dim(probeannotation)
rm(tmp, experimentalannot, controlannot)

probeannotation = probeannotation[!duplicated(probeannotation$Array_Address_Id),]
probeannotation = probeannotation[probeannotation$Array_Address_Id %in% rawdata$ID_REF, ] # 
probeannotation$Symbol=tolower(probeannotation$Symbol)
dimnames(probeannotation)[[1]] = probeannotation$Probe_Id
#dim(probeannotation)

#sort and filter probe and data similar.
datamatrix_raw = as.matrix(rawdata[,-1])
datamatrix_raw = datamatrix_raw[match( probeannotation$Array_Address_Id , rawdata$ID_REF), ]
dimnames(datamatrix_raw)[[1]] = probeannotation$Probe_Id
#dim(datamatrix_raw)
#dim(probeannotation)

#and match data to samples.
#table(sampleannotation$code %in% dimnames(datamatrix_raw)[[2]])# check
#table(dimnames(datamatrix_raw)[[2]] %in% sampleannotation$code)# check
datamatrix_raw = datamatrix_raw[, match(sampleannotation$code , dimnames(datamatrix_raw)[[2]])]
```

Several tables with results are presented in the Supporting Information of in Tawfic et al. We are aiming to reproduce these. Unfortunatly these were only presented in a pdf-format not easely parsed by a computer. We had to resort to a ad-hoc method of cut-and-paste from pdf into text files which were somewhat polluted by pdf-formatting code. For some tables about 10% of the rows will be lost, but the rest will suffice.

```{r}

table_s2 = read.table("data/table_s2.csv", sep="\t", header=TRUE, stringsAsFactors=FALSE, fill=TRUE)
table_s2 = as.matrix(table_s2)
table_s2[,3:6] = as.numeric(table_s2[,3:6])
# not able to paste the pdf whitout a lot of gibberish clutter the data. Some probes are lost!
a =rowSums(is.na(table_s2[,3:6])) > 0
print(paste("Lost rows table s2: ", sum(a)))
table_s2=data.frame(table_s2[!a,], stringsAsFactors=FALSE)
# data.frame makes the columns characters again.
for(n in 3:6)
{
  table_s2[,n]=as.numeric(table_s2[,n])
}

table_s3 = read.table("data/table_s3.csv", sep="\t", header=TRUE, stringsAsFactors=FALSE, fill=TRUE) 
#one Probe, ("1770446" or "ILM_1770446") was not found in the data and was thus removed from this textfile

table_s4 = read.table("data/table_s4.csv", sep="\t", header=TRUE, stringsAsFactors=FALSE, fill=TRUE)
#not able to copy-paste the pdf whitout a lot of gibberish clutter the data. Some probes are lost!
a = is.na(as.numeric(table_s4$Var_generic_vs_Var_GA))
print(paste("Lost rows table s4: ", sum(a)))
table_s4=table_s4[!a,]
table_s4$Var_generic_vs_Var_GA = as.numeric(table_s4$Var_generic_vs_Var_GA)

table_s5 = read.table("data/table_s5.csv", sep="\t", header=TRUE, stringsAsFactors=FALSE, fill=TRUE)
tmp = as.matrix(table_s5)
tmp[,3:14] = as.numeric(tmp[,3:14])
# not able to paste the pdf whitout a lot of gibberish clutter the data. Some probes are lost!
a =rowSums(is.na(tmp[,3:14])) > 0
print(paste("Lost rows table s5: ", sum(a)))
table_s5=data.frame(table_s5[!a,], stringsAsFactors=FALSE)
table_s5$Fold_Change = as.numeric(table_s5$Fold_Change)
```

### Reproduce the original results

```{r}
set.seed(100)
datamatrices = list()
datamatrices[["real_raw"]] = datamatrix_raw[1:5000,]

# used in paper, but seems to do the same as the limma version. Except it loses dimnames
datamatrices[["real_qnorm"]] = normalize.quantiles(datamatrices[["real_raw"]]) 
dimnames(datamatrices[["real_qnorm"]]) = dimnames(datamatrices[["real_raw"]])

mod = model.matrix(~as.factor(sampleannotation$covariate))
mod0 = model.matrix(~1,data=sampleannotation)

datamatrices[["real_combat_covariates"]]= as.matrix(ComBat(dat=datamatrices[["real_qnorm"]], batch=sampleannotation$chip, mod=mod, numCovs=NULL, par.prior=TRUE, prior.plots=FALSE))

```
The nameing convention seeme to differ slightly between the Table S1 and the text. This is the covariate labels with its corresponging name in the text.
- **DP** refferd to as GA (but not GA as in table S1)
- **N** refferd to as "generic"
- **M** reffered to as "medium"
- **RS** reffered to as "reference standard""

First we will try to reproduce Table S2.

> Genes utilized for the tolerance method illustrated in Figure 1B. (figure text)

> This standard of comparison was constructed by first identifying the top 1000 probes by absolute fold change of reference standard compared to the medium (Table S2). The list includes both upregulated and downregulated probes compared to medium. Probes were filtered such that ones upregulated by reference standard needed to have an average reference standard expression of 6.00 or higher and ones downregulated by reference standard needed to have an average medium expression of 6.00 or higher.

This is how the table looks
```{r}
head(table_s2)
```

**Table S2**
```{r}
thiscovariatename = "covariate"
thisdata = datamatrices[["real_combat_covariates"]]

# the mean uses log2 numbers and will be a geometric mean.
meanM = rowMeans(thisdata[, sampleannotation[, "covariate"]=="M"]) 
meanRS = rowMeans(thisdata[, sampleannotation[, "covariate"]=="RS"]) 
meanDP = rowMeans(thisdata[, sampleannotation[, "covariate"]=="DP"])
meanN = rowMeans(thisdata[, sampleannotation[, "covariate"] =="N"])

foldchange_RS_vs_M = meanRS - meanM
foldchange_RS_vs_M = foldchange_RS_vs_M[meanM>6 | meanRS>6]
top1000_RS_vs_M = names(foldchange_RS_vs_M[order(abs(foldchange_RS_vs_M), decreasing=TRUE)])[1:1000]
table(table_s2$ID %in% top1000_RS_vs_M) #  851 out of 909. For some reason it is not a 100% match.

#notin =!(table_s2$ID %in% top1000_RS_vs_M)
#notinnames = table_s2$ID[notin]
#cbind(table_s2[notin,], meanRS[notinnames], meanM[notinnames], FC=meanRS[notinnames]-meanM[notinnames])
```
```{r reproduced_tables2, dev='png', fig.width=8, fig.height=8, fig.show='hold', tidy=FALSE}
# scatter plot correlation of the 4 columns in table_S2 vs. the reproduced numbers. String correlation.
par(mfrow=c(2, 2))
plot(table_s2$AVG_Medium, meanM[table_s2$ID], pch=".")
plot(table_s2$AVG_Reference_Standard, meanRS[table_s2$ID], pch=".")
plot(table_s2$AVG_GA, meanDP[table_s2$ID], pch=".")
plot(table_s2$AVG_generic, meanN[table_s2$ID], pch=".")
```

**Table S3**

```{r}
#Table S3.
#The highly variable probes that were significant by F-test in either GA or generic (see methods section) and are depicted in Figure 1A.   
# MEthods: In order to identify probes with variability induced specifically by activation (as opposed to experimental noise), we sought to identify probes that were significantly more variable when activated with either GA or generic than medium. Using an F-test, we compared GA against Medium for each probe and compared generic against Medium. We then took the set of probes where either treatment comes up to be more variable than medium (union, passes in at least at least one). For those set of probes only, we compared the variability of GA across 34 samples representing 30 batches, to the variability of generic across 11 samples representing 5 batches, utilizing an F-test to measure significance of the differences between the probes.

# do a f-test between  DP and medium, and N and medium. And see what is "significant" (significance cut-off not noted in the text)?
pvalsN=apply( datamatrices[["real_combat_covariates"]], 1, 
              function(x) { var.test(x[sampleannotation$covariate=="N"], x[sampleannotation$covariate=="M"],  
              alternative="greater")$p.value } )
pvalsDP=apply( datamatrices[["real_combat_covariates"]], 1, 
              function(x) { var.test(x[sampleannotation$covariate=="DP"], x[sampleannotation$covariate=="M"],  
              alternative="greater")$p.value } ) 
padjustedN = p.adjust(pvalsN,method="BH")
padjustedDP = p.adjust(pvalsDP,method="BH")
print(table(padjustedN < 0.05)) 
print(table(padjustedDP < 0.05))
# nothing significant acording to 5%FDR that was used in the Differentially expressed genes tests.
```

```{r reproduced_tables3, dev='png', fig.width=8, fig.height=8, fig.show='hold', tidy=FALSE}
# look at the pvalues for the probes found in table_s3
thiscolors = c("black", "brown", "red", "blue")
plot(density(pvalsN[table_s3$Probe[table_s3$Type=="generic"]]), col=thiscolors[3], main="P-values for F-test, N or DP vs. Medium")
lines(density(pvalsN), col=thiscolors[1])
lines(density(pvalsDP), col=thiscolors[2])
lines(density(pvalsDP[table_s3$Probe[table_s3$Type=="GA"]]), col=thiscolors[4])
legend("topright", legend=c("All probes, N vs. M", "All probes, DP vs. M", "Table S3 probes, N vs. M", "Table S3 probes, DP vs. M"), text.col=thiscolors)
# The PRobes from table_s3 are at least enriched for low p-values.
# But table_s3 is not fully reproduced. More detailed explanation is needed
```

**Table S4**

```{r}
  # Table s4
# "Ranked list of probes by ratio of the variance in generic-activated samples to the variance in GA-activated samples."
# "Thus, we calculated for each probe the ratio of the variance in generic to the variance in GA"

# do a f-test between  N and DP. Correlate with table s4.
# f-test statistic is the same as ratio of variance?
fstatistics=apply( datamatrices[["real_combat_covariates"]], 1, 
              function(x) { var.test(x[sampleannotation$covariate=="N"], x[sampleannotation$covariate=="DP"],  
                                     alternative="greater")$statistic } )
fstatistics = fstatistics[order(fstatistics, decreasing=TRUE)]
table(names(fstatistics)[1:100] %in% table_s4$Probe) # not a complete reproduction.
# look at the f-statistic for the probes found in table_s3 vs all f-statistics
```

```{r reproduced_table_s4, dev='png', fig.width=8, fig.height=8, fig.show='hold', tidy=FALSE}
par(mfrow=c(2, 1))
plot(table_s4[,3], fstatistics[table_s4$Probe], main="F-statitics from table S4 vs. reproduced")
thiscolors = c("black", "red")
plot(density(fstatistics), col=thiscolors[1], main="Reproduced F-statistics for N vs. DP")
lines(density(fstatistics[table_s4$Probe]), col=thiscolors[2])
legend("topright", legend=c("All probes", "Table S4 probes"), text.col=thiscolors)
```

**Table S5**

```{r dev='png', fig.width=8, fig.height=8, fig.show='hold', tidy=FALSE}
# Table s5
#"Comparison of expression in GA to expression in GA for each probe, including fold change, ANOVA, LIMMA with background subtraction, comparative marker selection by signal-to-noise ratio, comparative marker selection by t-test, and the Wilcoxon non-parametric method."
#"Finally, genes that have significantly higher expression in samples activated by generic than in samples activated by GA by 4 different parametric methods (Table S5)"
#"Probes with Fold Change <  1.10 notdisplayed"

# Table s5 seems to be a comparison of expression in  GA (drug product) vs. generic (N) , listing the probes with greatest fold change (>abs(1.1)??) and the p-value calculated by several different tests. There seems to be no description of how the fold-change was calculated and after an inspection they look weird.

meanDP = rowMeans(thisdata[, sampleannotation$covariate=="DP"]) # DP is drug product also called GA
meanN = rowMeans(thisdata[, sampleannotation$covariate =="N"]) # N is Natco the generic,  (or G for alternative covariates)

# fold change between N and DP, calculated as geometric mean.
x = table_s5$Fold_Change
y = (meanDP - meanN)[table_s5$Probe]
lim =c( min(   c(min(x), min(y))  ), max( c(max(x), max(y)) ))
plot(x, y , ylim=lim, xlim=lim, pch=".", xlab="Table_S5", ylab="reproduced", main="FC comparison") # Not a perfect match. And the table_s5 numbers seems offset.
```
```{r dev='png', fig.width=8, fig.height=8, fig.show='hold', tidy=FALSE}
# table S2 has the AVG for both DP and N listed
fc_table_s2 = table_s2$AVG_GA - table_s2$AVG_generic
names(fc_table_s2) = table_s2$ID
x = table_s5$Fold_Change
y = fc_table_s2[table_s5$Probe]
lim =c( min(   c(min(x, na.rm=T), min(y, na.rm=T))  ), max( c(max(x, na.rm=T), max(y, na.rm=T)) ))
plot(x, y , ylim=lim, xlim=lim, pch="." ,  xlab="Table_S5", ylab="Table_S2", main="FC comparison") # perfect match.
```
# But this is odd, we were able to reproduce table s2 and table s2 matches table s5, but table s5 fold changes does not match our reproduced values as good as the match for table s2.

# The reason that the line is not as straight for the reproduce FC's for table s5 as for the mean expressions in table S2 is because of different scale in the plot and the fact that the small difference will be higher for a ratio (combining to values with a individual deviation from the tables.)
```{r dev='png', fig.width=8, fig.height=8, fig.show='hold', tidy=FALSE}
x = (meanDP - meanN)[table_s2$ID]
y = fc_table_s2
plot(x, y , ylim=lim, xlim=lim, pch="." ,  xlab="reproduced", ylab="Table_S2", main="FC comparison") # perfect match.
```

```{r}
# Still there is a offset in the numbers.
commonprobes = table_s5$Probe[table_s5$Probe %in% table_s2$ID]
fc_table_s5 = table_s5$Fold_Change[match(commonprobes ,table_s5$Probe)]
head(cbind(fc_table_s2[commonprobes], fc_table_s5, fc_table_s5-fc_table_s2[commonprobes], fc_table_s5/fc_table_s2[commonprobes]))
# seems that some unacounted transformation has occured on the fold changes.
```
```{r}
source("../../commonscripts/helper_functions.r")
getdifftab_limma
```
```{r}
# reproduce the Nom P-value columns
limma_ret= getdifftab_limma(edata=datamatrices[["real_combat_covariates"]], condition=sampleannotation[,"covariate"], contrast="DP-N" )
limma_p = limma_ret$p
names(limma_p) = limma_ret$gene
DP = sampleannotation$covariate=="DP"
N = sampleannotation$covariate=="N"
ttest_p = apply(datamatrices[["real_combat_covariates"]], 1, function(x) { t.test(x[DP], x[N])$p.value } )
anova_p = apply(datamatrices[["real_combat_covariates"]], 1, function(x){ summary(aov(x[DP|N]~sampleannotation$covariate[DP|N]))[[1]][1,5]}  )  # aov er kanskje feil for dette, Er ikke live mange pr?ver i guppene.
wilcoxon_p = apply(datamatrices[["real_combat_covariates"]], 1, function(x) { wilcox.test(x[sampleannotation$covariate=="DP"], x[sampleannotation$covariate=="N"])$p.value } )
```
```{r dev='png', fig.width=8, fig.height=8, fig.show='hold', tidy=FALSE}
par(mfrow=c(2, 2))
#plot(table_s5$Fold_Change, (meanDP - meanN)[table_s5$Probe] ,  ylim=c(-1.5,1.5), xlim=c(-1.5,1.5), pch=".", xlab="Table_S5", ylab="reproduced", main="FC comparison")
plot(log10(table_s5$t_test_p), log10(ttest_p[table_s5$Probe]), pch=".", main="t-test p")
plot(log10(table_s5$limma_p), log10(limma_p[table_s5$Probe]), pch=".", main="limma p")
plot(log10(table_s5$anova_p), log10(anova_p[table_s5$Probe]), pch=".", main="anova p")
plot(log10(table_s5$wilcoxon_p), log10(wilcoxon_p[table_s5$Probe]), pch=".", main="wilcoxon p")
```
# The p-values does not make a straight line, but they are low. But maybe all are low?

```{r dev='png', fig.width=8, fig.height=8, fig.show='hold', tidy=FALSE}
par(mfrow=c(2, 2))
thiscolors=c("black", "red")
hist(ttest_p, border=thiscolors[1], main="Reproduced t-test p-value distribution.", breaks=100)
hist(ttest_p[table_s5$Probe], border=thiscolors[2], add=T, breaks=100)
legend("topright", legend=c("all probes", "S5 probes"), text.col=thiscolors)
hist(limma_p, border=thiscolors[1], main="Reproduced limma p-value distribution.", breaks=100)
hist(limma_p[table_s5$Probe], border=thiscolors[2], add=T, breaks=100)
legend("topright", legend=c("all probes", "S5 probes"), text.col=thiscolors)
hist(anova_p, border=thiscolors[1], main="Reproduced anova p-value distribution.", breaks=100)
hist(anova_p[table_s5$Probe], border=thiscolors[2], add=T, breaks=100)
legend("topright", legend=c("all probes", "S5 probes"), text.col=thiscolors)
hist(wilcoxon_p, border=thiscolors[1], main="Reproduced wilcoxon p-value distribution.", breaks=100)
hist(wilcoxon_p[table_s5$Probe], border=thiscolors[2], add=T, breaks=100)
legend("topright", legend=c("all probes", "S5 probes"), text.col=thiscolors)
```

### 2014-02-24 commonscripts/helper_functions.r

@@ -1,0 +1,290 @@
+
+createsampleannotation = function( treatment_in_batches)
+{
+	batches = vector()
+	treatments = vector()
+	for(i in 1:length(treatment_in_batches))
+	{
+		#thisbatch = paste("batch", i, sep="")
+		thisbatch = i
+		for(s in 1:length(treatment_in_batches[[i]]))
+		{
+			batches = c(batches, rep(thisbatch, treatment_in_batches[[i]][s]))
+			treatments = c(treatments, rep(paste("treatment", s, sep=""), thisbatch, treatment_in_batches[[i]][s]))
+		}
+	}
+	sampleannotation = data.frame(id=paste("sample", 1:length(treatments), sep=""),    treatment = treatments, batch=batches)
+	return(sampleannotation)
+}
+
+
+
+
+createrandomdata = function(ngenes, sa, mean=0, sd=1)
+{
+	ret = matrix( rnorm(dim(sa)[1]*ngenes, mean=mean, sd=sd), ncol=dim(sa)[1], nrow=ngenes)
+	dimnames(ret) = list( paste("gene", 1:ngenes, sep=""),  sa$id)
+	return(ret)
+}
+
+addbatcheffect = function(df, batches, thismean=1, thissd=1)
+{
+	ret = df
+	for(i in 1:length(unique(batches)))
+	{
+		for(s in 1:dim(df)[1])
+		{
+			thisgenesbatcheffect = rnorm(1,mean=thismean, sd=thissd)
+			a=batches==i
+			#ret[s, a] = ret[s, a] + rnorm(length(ret[s, a]), mean=thisgenesbatcheffect)	
+			ret[s, a] = ret[s, a] + thisgenesbatcheffect
+		}
+	}
+	return(ret)
+}
+
+# 
+addconditioneffect = function(df, conditions, ndiffgenes, thismean=1, betweensd=1, insidesd=1)
+{
+	ret = df
+	if(ndiffgenes>0)
+	{
+		for(i in 1:length(unique(conditions)))
+		{
+			for(s in 1:ndiffgenes)
+			{
+				thisgeneseffect = rnorm(thismean, sd=betweensd)
+				a=conditions==conditions[i]
+				#ret[s, a] = ret[s, a]+ rnorm(length(ret[s, a]), mean=thisgeneseffect)		
+				ret[s, a] = rnorm(length(ret[s, a]), mean=thisgeneseffect, sd=insidesd)		
+			}
+		}
+	}
+	return(ret)
+}
+
+
[…]

### 2014-02-24 reanalysis/Fadi_Towfic_generic_medicine/rscript.r

@@ -1,0 +1,1112 @@
+
+
+
+
+#setwd("C:\\Users\\vegard2013\\Dropbox\\div\\jobb\\combat_test\\papers\\Fadi_Towfic_generic_medicine")
+setwd("C:\\Users\\vegard2013\\Dropbox\\div\\jobb\\combat_test\\papers\\Fadi_Towfic_generic_medicine")
+#    setwd("/Users/vegardnygaard/Dropbox/div/jobb/combat_test/papers/Fadi_Towfic_generic_medicine")
+
+#  source("http://bioconductor.org/biocLite.R")
+#      biocLite("preprocessCore")
+#      biocLite("GEOquery")
+source("../../helper_functions.r")
+
+library(Biobase)
+library(GEOquery)
+library(limma)
+library(sva)
+library(RColorBrewer)
+library(preprocessCore)
+library(car)
+
+########################################################################
+##   READ SAMPLEANNOTATION		
+########################################################################
+sampleannotation = read.table("data/sampleannotation.csv", sep="\t", header=TRUE,  stringsAsFactors=FALSE)
+sampleannotation$code = make.names(sampleannotation$code)
+sampleannotation$chip = as.character(sampleannotation$chip)
+dimnames(sampleannotation)[[1]] = sampleannotation$code
+
+#table(sampleannotation$group)
+#table(sampleannotation$chip, sampleannotation$group)
+sampleannotation = sampleannotation[!is.na(sampleannotation$group),] # take out 3 samples that are not assign to a group. Failed QC?
+#sampleannotation = sampleannotation[!is.na(sampleannotation$covariate1),]
+table(sampleannotation$chip, sampleannotation$covariate1) # the batch-covariate design.
+
+
+########################################################################
+##   READ Raw data file	and probe annotation file. Filter and match
+########################################################################
+
+rawdata = read.table("data/GSE40566_non_normalized.txt", sep="\t", header=TRUE, stringsAsFactors=FALSE)
+
+rawannotation = read.table("data/GPL6887_MouseWG-6_V2_0_R3_11278593_A.txt", sep="\t", header=TRUE, stringsAsFactors=FALSE, skip=8, comment.char="", quote="", fill=TRUE)
+
+# annotation file has two separate tables, one for experimental and one at the end for controls.
+# splitting and fixing the tables:
+a =  rawannotation$Species=="Mus musculus"
[…]

### 2014-02-24 reanalysis/ComBat_original/combat_paper_dataset2_reproduced.rmd
fra iMAc

@@
 
-```{r, results='hide',  message=FALSE}
+```{r, results='hide',  message=FALSE, error=FALSE}
 includelibs = c("pheatmap", "sva", "qvalue", "limma")
-lapply(includelibs, require, character.only=T)
+tmp = lapply(includelibs, require, character.only=T)
+print(tmp)
+if(any(!unlist(tmp)))
+{
+  stop( paste("Not able to find all packages. Please install ",
+              paste(includelibs[!unlist(tmp)], collapse=", ") )
+              )
+}
+rm(tmp)
 ```
-```{r, cache=TRUE}
-datamatrix = as.matrix(read.table("data/dataExample2.txt", sep="\t", header=TRUE))#[1:3000,]# for dev.
+```{r, cache=TRUE, tidy=FALSE}
+datamatrix = as.matrix(read.table("data/dataExample2.txt", sep="\t", header=TRUE))#[1:3000,]#dev.
 sampleannotation = read.table("data/sampleInfoExample2.txt", 
@@
 # must be discrete for the pheatmap
-#table(sampleannotation$ArrayName==dimnames(datamatrix)[[2]]) # ordercheck
+#table(sampleannotation$ArrayName==dimnames(datamatrix)[[2]])#ordercheck
 datamatrix=datamatrix[,sampleannotation$Type!="WT"]

### 2014-02-25 reanalysis/Fadi_Towfic_generic_medicine/towfic_et_at.rmd
update towfic

@@
 
-The processed data from the above GEO deposit was deposited for another publication which did not processes the data as described in Tawfic et al.
+The processed data from the above GEO deposit was deposited for another publication which did not processes the data as described in Towfic et al.
 (personal communication).
@@
 # the output from Illumina software.
-rawdata = read.table("data/GSE40566_non_normalized.txt", sep="\t", header=TRUE, stringsAsFactors=FALSE)
+rawdata = read.table("not_in_github/GSE40566_non_normalized.txt", sep="\t", header=TRUE, stringsAsFactors=FALSE)
 
 # the probe annotation file found inside GSE40566_RAW.tar
-rawannotation = read.table("data/GPL6887_MouseWG-6_V2_0_R3_11278593_A.txt", sep="\t", header=TRUE, stringsAsFactors=FALSE, skip=8, comment.char="", quote="", fill=TRUE)
+rawannotation = read.table("not_in_github/GPL6887_MouseWG-6_V2_0_R3_11278593_A.txt", sep="\t", header=TRUE, stringsAsFactors=FALSE, skip=8, comment.char="", quote="", fill=TRUE)
 ```
@@
 
-Several tables with results are presented in the Supporting Information of in Tawfic et al.
+Several tables with results are presented in the Supporting Information of in Towfic et al.
 We are aiming to reproduce these.
@@
 datamatrices = list()
-datamatrices[["real_raw"]] = datamatrix_raw[1:5000,]
+datamatrices[["real_raw"]] = datamatrix_raw#[1:5000,]
 
@@
 ```
+The numbers in Table S2 is almost reproduced.
 
@@
 
-```{r}
-#Table S3.
-#The highly variable probes that were significant by F-test in either GA or generic (see methods section) and are depicted in Figure 1A.   
-# MEthods: In order to identify probes with variability induced specifically by activation (as opposed to experimental noise), we sought to identify probes that were significantly more variable when activated with either GA or generic than medium.
+> The highly variable probes that were significant by F-test in either GA or generic (see methods section) and are depicted in Figure 1A.<cite> Figure text, Towfic et al.
+
+> In order to identify probes with variability induced specifically by activation (as opposed to experimental noise), we sought to identify probes that were significantly more variable when activated with either GA or generic than medium.
 Using an F-test, we compared GA against Medium for each probe and compared generic against Medium.
 We then took the set of probes where either treatment comes up to be more variable than medium (union, passes in at least at least one).
-For those set of probes only, we compared the variability of GA across 34 samples representing 30 batches, to the variability of generic across 11 samples representing 5 batches, utilizing an F-test to measure significance of the differences between the probes.
+For those set of probes only, we compared the variability of GA across 34 samples representing 30 batches, to the variability of generic across 11 samples representing 5 batches, utilizing an F-test to measure significance of the differences between the probes.<cite> Methods, Towfic et al.
 
-# do a f-test between  DP and medium, and N and medium. And see what is "significant" (significance cut-off not noted in the text)?
+There is not description of what was used as a significance threshold here, but for the Identification of differentially expressed genes (TAble S5) a 0.05 FDR cut-off.
+
+```{r}
 pvalsN=apply( datamatrices[["real_combat_covariates"]], 1, 
@@
 ```
+We are not able to reproduce this list. An alternative is to inspect the p-values for the probes in the list and see how they compare to the overall p-values.
 
 ```{r reproduced_tables3, dev='png', fig.width=8, fig.height=8, fig.show='hold', tidy=FALSE}
-# look at the pvalues for the probes found in table_s3
-thiscolors = c("black", "brown", "red", "blue")
-plot(density(pvalsN[table_s3$Probe[table_s3$Type=="generic"]]), col=thiscolors[3], main="P-values for F-test, N or DP vs. Medium")
-lines(density(pvalsN), col=thiscolors[1])
-lines(density(pvalsDP), col=thiscolors[2])
-lines(density(pvalsDP[table_s3$Probe[table_s3$Type=="GA"]]), col=thiscolors[4])
-legend("topright", legend=c("All probes, N vs. M", "All probes, DP vs. M", "Table S3 probes, N vs. M", "Table S3 probes, DP vs. M"), text.col=thiscolors)
+par(mfrow=c(2, 1))
+thiscolors = c("red", "blue")
+hist( pvalsN[table_s3$Probe[table_s3$Type=="generic"]], 
+      breaks=100, border="red", 
+      main="P-values for F-test, Table S3 probes",
+      xlab="P-values")
+hist(pvalsDP[table_s3$Probe[table_s3$Type=="GA"]], 
+     breaks=100, border="blue", add=T)
+legend("topright", legend=c("N vs. M", "DP vs. M"), text.col=c("red", "blue"))
+hist( pvalsN, 
+      breaks=100, border="red", 
+      main="All probes", xlab="P-values")
+hist(pvalsDP, 
+     breaks=100, border="blue", add=T)
+
+```
 # The PRobes from table_s3 are at least enriched for low p-values.
 # But table_s3 is not fully reproduced. More detailed explanation is needed
-```
 
@@
 ```
-# But this is odd, we were able to reproduce table s2 and table s2 matches table s5, but table s5 fold changes does not match our reproduced values as good as the match for table s2.
+But this is odd, we were able to reproduce table s2 and table s2 matches table s5, but table s5 fold changes does not match our reproduced values as good as the match for table s2.
 
-# The reason that the line is not as straight for the reproduce FC's for table s5 as for the mean expressions in table S2 is because of different scale in the plot and the fact that the small difference will be higher for a ratio (combining to values with a individual deviation from the tables.)
+The reason that the line is not as straight for the reproduce FC's for table s5 as for the mean expressions in table S2 is because of different scale in the plot and the fact that the small difference will be higher for a ratio (combining to values with a individual deviation from the tables.)
 ```{r dev='png', fig.width=8, fig.height=8, fig.show='hold', tidy=FALSE}

### 2014-02-26 reanalysis/Fadi_Towfic_generic_medicine/towfic_et_at.rmd
more on towfic

@@
 - Estimate the error introduced by ComBat and the consequences for the conclusion of the study
-- Perform a few more sanity checks to substantiate that the difference in results for the two above analysis is mainly due to error introduced by ComBat
+- Perform a few more sanity checks to substantiate that the difference in results for the two above analysis is mainly introduced by ComBat
 
@@
 ```
-The important sample information is described in Table S1 from Towfic et al. the its usage in ComBat is described briefly in the methods;
+The important sample information is described in Table S1 from Towfic et al. and its usage in ComBat is described briefly in the methods;
 >  Each microarray’s chip designation was supplied a batch label; there were 18 batches in all.
@@
 
-Based on these descriptions and the annotation for the GEO deposit, we compiled a more complete sample annotation file.
+Based on these descriptions and the annotation for the GEO deposit, we compiled a more complete sample annotation file (sampleannotation.csv).
 
@@
 ```
-The "covariate"-coloumn is made based on the "code"-coloumn, but might not match  to what were actually used.
+The "covariate"-coloumn is made based on the "code"-coloumn, but might not match  to what were actually used for all samples.
 The last two coloumns are from the GEO deposit and reveal that 3 of the samples do not have data and are taken out.
+
+The nameing convention seeme to differ slightly between the Table S1 and the text in Towfic et al. This is our interpretation of covariate labels and its corresponding name in the text.
+- **DP** refferd to as GA (but not GA as in table S1)
+- **N** refferd to as "generic"
+- **M** reffered to as "medium"
+- **RS** reffered to as "reference standard""
 
@@
 ```{r}
-# the batch/covariate design
 table(sampleannotation[, c("chip", "covariate")])
@@
 ```{r}
-table(sampleannotation[sampleannotation$covariate %in% c("DP", "N"), c("chip", "covariate")])
+table(sampleannotation[sampleannotation$covariate %in% c("DP", "N"), 
+                       c("chip", "covariate")])
 ```
@@
 (personal communication).
-The raw data linked to as [GSE40566_RAW.tar](http://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE40566&format=file) and [GSE40566_non_normalized.txt.gz](http://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE40566&format=file&file=GSE40566%5Fnon%5Fnormalized%2Etxt%2Egz) were thus used, although it is not confirmed that this is the version of the data actually used in Towfic et al.
+The files linked to as [GSE40566_RAW.tar](http://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE40566&format=file) and [GSE40566_non_normalized.txt.gz](http://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE40566&format=file&file=GSE40566%5Fnon%5Fnormalized%2Etxt%2Egz) were thus used, although it is not confirmed that this is the version of the data actually used in Towfic et al.
+
+A recap of the important input files used in this report:
+- **sampleannotation.csv** holds the important sample/batch/covariate assignment. Compiled based on inforamtion in Towfic et al, the geo deposit and personal communication.
+- **GSE40566_non_normalized.txt** the measurments in a sample vs probe matrix.
+From the [GEO deposit](http://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE40566&format=file&file=GSE40566%5Fnon%5Fnormalized%2Etxt%2Egz).
+- **GPL6887_MouseWG-6_V2_0_R3_11278593_A.txt** the probe annotation needed for the datarow to probe to gene matching.
+From the GEO deposit inside the [GSE40566_RAW.tar](http://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE40566&format=file).
+
+Reading the data.
 ```{r}
@@
 
-Then some formatting is needed to match the probe information to the measurments and the measurments to the sample information.
+Some formatting is needed to match the probe information to the measurments and the measurments to the sample information.
 ```{r}
@@
 
-Several tables with results are presented in the Supporting Information of in Towfic et al.
+Several tables with results are presented in the Supporting Information in Towfic et al.
 We are aiming to reproduce these.
@@
 We had to resort to a ad-hoc method of cut-and-paste from pdf into text files which were somewhat polluted by pdf-formatting code.
-For some tables about 10% of the rows will be lost, but the rest will suffice.
+For some tables about 10% of the rows will be lost, but the rest will suffice for our purpose.
 
 ```{r}
-
+# Table S2
 table_s2 = read.table("data/table_s2.csv", sep="\t", header=TRUE, stringsAsFactors=FALSE, fill=TRUE)
@@
 
+# Table S3
 table_s3 = read.table("data/table_s3.csv", sep="\t", header=TRUE, stringsAsFactors=FALSE, fill=TRUE) 
@@
 
+# Table S4
 table_s4 = read.table("data/table_s4.csv", sep="\t", header=TRUE, stringsAsFactors=FALSE, fill=TRUE)
@@
 
+# Table S5
 table_s5 = read.table("data/table_s5.csv", sep="\t", header=TRUE, stringsAsFactors=FALSE, fill=TRUE)
@@
 
+> Starting with background-corrected bead-level signals, we quantile normalized the extracted data for all samples across all 46,547 probes via the “preprocessCore” package in R.
+
 ```{r}
-set.seed(100)
 datamatrices = list()
@@
 dimnames(datamatrices[["real_qnorm"]]) = dimnames(datamatrices[["real_raw"]])
+```
 
+> We then corrected for batch variation with ComBat [17] as implemented in the SVA package of R [18].
+Each microarray’s chip designation was supplied a batch label; there were 18 batches in all.
+The labels for the treatments (i.e.
+drug product, reference standard…) were added as covariates.
+
+```{r}
 mod = model.matrix(~as.factor(sampleannotation$covariate))
@@
 ```
-The nameing convention seeme to differ slightly between the Table S1 and the text. This is the covariate labels with its corresponging name in the text.
-- **DP** refferd to as GA (but not GA as in table S1)
-- **N** refferd to as "generic"
-- **M** reffered to as "medium"
-- **RS** reffered to as "reference standard""
 
@@
 ```
-# The PRobes from table_s3 are at least enriched for low p-values.
-# But table_s3 is not fully reproduced. More detailed explanation is needed
+The probes from Table S3 are at least enriched for low p-values. A more detailed explanation is needed for a better reproducablity.
 
@@
 anova_p = apply(datamatrices[["real_combat_covariates"]], 1, function(x){ summary(aov(x[DP|N]~sampleannotation$covariate[DP|N]))[[1]][1,5]}  )  # aov er kanskje feil for dette, Er ikke live mange pr?ver i guppene.
-wilcoxon_p = apply(datamatrices[["real_combat_covariates"]], 1, function(x) { wilcox.test(x[sampleannotation$covariate=="DP"], x[sampleannotation$covariate=="N"])$p.value } )
+
+# wilcox.test gives a few hundred warnings for ties.
+# Warning: cannot compute exact p-value with ties
+wilcoxon_p = suppressWarnings(apply(datamatrices[["real_combat_covariates"]], 1, function(x) { wilcox.test(x[sampleannotation$covariate=="DP"], x[sampleannotation$covariate=="N"])$p.value } )
+)
 ```

### 2014-02-27 reanalysis/Fadi_Towfic_generic_medicine/towfic_et_at.rmd
more towfic

@@
 
-The processed data from the above GEO deposit was deposited for another publication which did not processes the data as described in Towfic et al.
+The processed data from the above GEO deposit was from another publication which did not processes the data as described in Towfic et al.
 (personal communication).
-The files linked to as [GSE40566_RAW.tar](http://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE40566&format=file) and [GSE40566_non_normalized.txt.gz](http://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE40566&format=file&file=GSE40566%5Fnon%5Fnormalized%2Etxt%2Egz) were thus used, although it is not confirmed that this is the version of the data actually used in Towfic et al.
+The lesser processed linked to as [GSE40566_non_normalized.txt.gz](http://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE40566&format=file&file=GSE40566%5Fnon%5Fnormalized%2Etxt%2Egz) in the supplementary data was thus used in this report, although it is not confirmed that this is the version of the data actually used in Towfic et al.
+
+Probe annotation was also found in the supplementary data as [GSE40566_RAW.tar](http://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE40566&format=file).
 
@@
 
+Or at least some of it.
+
 > Starting with background-corrected bead-level signals, we quantile normalized the extracted data for all samples across all 46,547 probes via the “preprocessCore” package in R.
@@
 table(table_s2$ID %in% top1000_RS_vs_M) #  851 out of 909. For some reason it is not a 100% match.
+```
+Out of Table S2's **`r nrow(table_s2)`** probes, **`r sum(table_s2$ID %in% top1000_RS_vs_M)`** were reproduced. Scatter plots of the actual values shows a correlation.
 
-#notin =!(table_s2$ID %in% top1000_RS_vs_M)
-#notinnames = table_s2$ID[notin]
-#cbind(table_s2[notin,], meanRS[notinnames], meanM[notinnames], FC=meanRS[notinnames]-meanM[notinnames])
-```
 ```{r reproduced_tables2, dev='png', fig.width=8, fig.height=8, fig.show='hold', tidy=FALSE}
@@
 
-There is not description of what was used as a significance threshold here, but for the Identification of differentially expressed genes (TAble S5) a 0.05 FDR cut-off.
+There is no description of what was used as a significance threshold in this table, but for the Identification of differentially expressed genes (Table S5), a FDR cut-off of 0-05 was applied.
 
@@
 ```
-The probes from Table S3 are at least enriched for low p-values. A more detailed explanation is needed for a better reproducablity.
+The probes from Table S3 are at least enriched for low p-values. A more detailed explanation is needed for a better reproduction.
 
@@
 
+> Ranked list of probes by ratio of the variance in generic-activated samples to the variance in GA-activated samples.<cite> Figure text, Towfic et al.
+
+> Thus, we calculated for each probe the ratio of the variance in generic to the variance in GA
+
 ```{r}
 # Table s4
-# "Ranked list of probes by ratio of the variance in generic-activated samples to the variance in GA-activated samples."
-# "Thus, we calculated for each probe the ratio of the variance in generic to the variance in GA"
-
-# do a f-test between  N and DP. Correlate with table s4.
 # f-test statistic is the same as ratio of variance?
@@
 table(names(fstatistics)[1:100] %in% table_s4$Probe) # not a complete reproduction.
-# look at the f-statistic for the probes found in table_s3 vs all f-statistics
 ```
+Out of Table S4's **`r nrow(table_s4)`** probes, **`r sum(names(fstatistics)[1:100] %in% table_s4$Probe)`** were reproduced.
 
@@
 par(mfrow=c(2, 1))
-plot(table_s4[,3], fstatistics[table_s4$Probe], main="F-statitics from table S4 vs. reproduced")
+plot(table_s4[,3], fstatistics[table_s4$Probe], main="F-statistic from table S4 vs. reproduced")
 thiscolors = c("black", "red")
-plot(density(fstatistics), col=thiscolors[1], main="Reproduced F-statistics for N vs. DP")
-lines(density(fstatistics[table_s4$Probe]), col=thiscolors[2])
+hist(fstatistics, breaks=100, freq=F, border=thiscolors[1], main="Reproduced F-statistic for N vs. DP")
+hist(fstatistics[table_s4$Probe], breaks=100, add=T, freq=F, border=thiscolors[2])
 legend("topright", legend=c("All probes", "Table S4 probes"), text.col=thiscolors)
+
+hist(fstatistics[table_s4$Probe], breaks=100, add=F, freq=T, border=thiscolors[2])
 ```
+The scatter plots of the values shows a weak correlation, but they are all high. And the histogram shows that the probes from Table S4 all have a high reproduced F-statistic compared to the overall.
 
 **Table S5**
+> Comparison of expression in GA to expression in GA for each probe, including fold change, ANOVA, LIMMA with background subtraction, comparative marker selection by signal-to-noise ratio, comparative marker selection by t-test, and the Wilcoxon non-parametric method.
+<cite> Figure text, Towfic et al.
 
+How this fold change was calculated is not explained. A common way is to take the difference of geometric means that were calculated for Table S2 above.
 ```{r dev='png', fig.width=8, fig.height=8, fig.show='hold', tidy=FALSE}
-# Table s5
-#"Comparison of expression in GA to expression in GA for each probe, including fold change, ANOVA, LIMMA with background subtraction, comparative marker selection by signal-to-noise ratio, comparative marker selection by t-test, and the Wilcoxon non-parametric method."
-#"Finally, genes that have significantly higher expression in samples activated by generic than in samples activated by GA by 4 different parametric methods (Table S5)"
+foldchange_DP_N = meanDP - meanN
+
+ # make the axis equal.
+lim =c( min(   c(min(table_s5$Fold_Change), min(foldchange_DP_N[table_s5$Probe]))  ), 
+        max( c(max(table_s5$Fold_Change), max(foldchange_DP_N[table_s5$Probe])) ))
+
+# plot
+plot(table_s5$Fold_Change, foldchange_DP_N[table_s5$Probe],
+     ylim=lim, xlim=lim, pch=".", xlab="Table_S5", ylab="reproduced",
+     main="FC comparison")
+
+```
+The plot reveals that the reproduced fold changes are much more compressed than the ones in Table S5.
+There seems to two linear relationship, one for neagtive FC's and another for positve FC's. They seems to have different intercepts but comparable slope.
+The methods part is more specific;
+>  First, we computed the statistical significance of differential expression between treatments using the Analysis of Variance (ANOVA) method for each probe [21], adjusting for multiple hypothesis testing using the Benjamini-Hochberg False Discovery Rate (FDR) correction [22].
+
 #"Probes with Fold Change <  1.10 notdisplayed"

### 2014-03-03 reanalysis/ComBat_original/combat_paper_dataset2_reproduced.rmd
ttttest
mer

@@
 Type = as.factor(sampleannotation$Type)
-Block = as.factor(sampleannotation$Batch)
 design = model.matrix(~0 + Type)
@@
 
-Additional sanity checks
----------
+### Additional sanity checks
+
 To substantiate that the result from the use of ComBat is less trustworthy than the alternative analysis we provide a few additional sanity checks.
@@
 generation ended `r as.character(Sys.time())`. Time spent `r  as.integer(round(difftime(Sys.time(),starttime, units="mins")))` minutes .
-
-table(sampleannotation[ sampleannotation$Batch!="3", c("Batch", "Cell")]  )
-table(Batch45_annot[, c("Batch", "Cell")])
-table(sampleannotation[, c("Batch", "Cell")])

### 2014-03-03 commonscripts/helper_functions.r

@@ -92,7 +92,8 @@ getdifftab_limma = function(edata, condition,  contrast, block=NULL, threshold=0
 	if(!is.null(block))
 	{
 		block = as.factor(block)
-		design <- model.matrix(~block+fac)
+		#design <- model.matrix(~block+fac)
+		design = model.matrix(~0+fac+block)
 	}
 	colnames(design)=make.names(colnames(design))
 	#contrast = paste(unique(sa$treatment)[1], "-", unique(sa$treatment)[2], sep="")
@@ -101,18 +102,14 @@ getdifftab_limma = function(edata, condition,  contrast, block=NULL, threshold=0
 	contrast = paste("fac", contrast, sep="")
 	contrast = sub("-", "-fac", contrast)
 	cont.matrix = makeContrasts ( contrasts=contrast, levels=design)	
-	fit2 = contrasts.fit(fit, cont.matrix)
-		
-	#fit <- eBayes(fit)
-	#topTable(fit, coef="facM")
-	
-	fit2 <- eBayes(fit2)
-	#topTable(fit2)	
-	ret = data.frame(gene=names(fit2$Amean), p=fit2$p.value[,1], padjusted=p.adjust(fit2$p.value[,1] , method="fdr"), fc=fit2$coefficients[,1], stringsAsFactors=FALSE)
+	fit2 = contrasts.fit(fit, cont.matrix)	
+	fit2 <- eBayes(fit2)	
+	ret = data.frame(gene=names(fit2$Amean), 
+                   p=fit2$p.value[,1], 
+                   padjusted=p.adjust(fit2$p.value[,1] , method="fdr"),
+                   fc=fit2$coefficients[,1],
+                   stringsAsFactors=FALSE)
 	dimnames(ret)[[1]] = ret$gene 
-	# non eBayes
-	#nonbayestab=toptable(fit2, number=9999999)[names(fit2$Amean),]	
-	#ret = data.frame(gene=names(fit2$Amean), p=nonbayestab[,"P.Value"], padjusted=p.adjust(nonbayestab$P.Value , method="fdr"), fc=fit2$coefficients[,1])
 	return(ret)
 }
 

### 2014-03-04 reanalysis/Fadi_Towfic_generic_medicine/towfic_et_al.rmd
permutation towfic

Reproduction and alternative analysis of "Comparing the Biological Impact of Glatiramer Acetate with the Biological Impact of a Generic"
========================================================

```{r, results='hide', echo=FALSE}
starttime = Sys.time()
```
`r as.character(starttime)`

### Overview
This report aims to show that the use of the statistical tool ComBat from [Johnson et al.](http://biostatistics.oxfordjournals.org/content/8/1/118.abstract) led to false results in the analysis performed in [Towfic et al.'s ](http://www.plosone.org/article/info%3Adoi%2F10.1371%2Fjournal.pone.0083757) "Comparing the Biological Impact of Glatiramer Acetate with the Biological Impact of a Generic".

This document has five main parts
- Getting and formatting the data
- Reproduce some of the results to show that we are working on the same data and analysis workflow
- Remove the use of ComBat and perform an similar analysis with alternative established tools
- Estimate the error introduced by ComBat and the consequences for the conclusion of the study
- Perform a few more sanity checks to substantiate that the difference in results for the two above analysis is mainly introduced by ComBat

### Getting the data

```{r, results='hide', message=FALSE}
includelibs = c("Biobase", "GEOquery", "sva", "limma",
                 "preprocessCore")
lapply(includelibs, require, character.only=T)
```
The important sample information is described in Table S1 from Towfic et al. and its usage in ComBat is described briefly in the methods;
>  Each microarray’s chip designation was supplied a batch label; there were 18 batches in all. The labels for the treatments (i.e. drug product, reference standard…) were added as covariates.<cite> Towfic et al.

Table S1 does have a "Chip"- coloumn, unfortunatly there is no dedicated "treatment"-coloumn.
Communication with the corresponding author yielded this explanation
> ..and the only thing we used as a covariate was the unique treatment names themselves (e.g. ³Medium² or ³RS").

Based on these descriptions and the annotation for the GEO deposit, we compiled a more complete sample annotation file (sampleannotation.csv).

```{r}
sampleannotation = read.table("data/sampleannotation.csv", sep="\t", header=TRUE,  stringsAsFactors=FALSE)
sampleannotation$code = make.names(sampleannotation$code)
sampleannotation$chip = as.character(sampleannotation$chip)
dimnames(sampleannotation)[[1]] = sampleannotation$code
head(sampleannotation)
```
The "covariate"-coloumn is made based on the "code"-coloumn, but might not match  to what were actually used for all samples. The last two coloumns are from the GEO deposit and reveal that 3 of the samples do not have data and are taken out.

The nameing convention seeme to differ slightly between the Table S1 and the text in Towfic et al. This is our interpretation of covariate labels and its corresponding name in the text.
- **DP** refferd to as GA (but not GA as in table S1)
- **N** refferd to as "generic"
- **M** reffered to as "medium"
- **RS** reffered to as "reference standard""

```{r}
# take out 3 samples that are not assign to a geoaccession. Failed QC?
sampleannotation = sampleannotation[!is.na(sampleannotation$geoaccession),] 
```
The batch/covariate design shows many batches and covariate groups.
```{r}
table(sampleannotation[, c("chip", "covariate")])
```
A look at the most interesting groups, "DP"(GA) and "N"(generic) reveals a lack of balance.
```{r}
table(sampleannotation[sampleannotation$covariate %in% c("DP", "N"), 
                       c("chip", "covariate")])
```

> The microarray data have been deposited in the Gene Expression Omnibus, under accession number [GSE40566](http://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE40566)

The processed data from the above GEO deposit was from another publication which did not processes the data as described in Towfic et al. (personal communication). The lesser processed linked to as [GSE40566_non_normalized.txt.gz](http://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE40566&format=file&file=GSE40566%5Fnon%5Fnormalized%2Etxt%2Egz) in the supplementary data was thus used in this report, although it is not confirmed that this is the version of the data actually used in Towfic et al.

Probe annotation was also found in the supplementary data as [GSE40566_RAW.tar](http://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE40566&format=file).

A recap of the important input files used in this report:
- **sampleannotation.csv** holds the important sample/batch/covariate assignment. Compiled based on inforamtion in Towfic et al, the geo deposit and personal communication.
- **GSE40566_non_normalized.txt** the measurments in a sample vs probe matrix. From the [GEO deposit](http://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE40566&format=file&file=GSE40566%5Fnon%5Fnormalized%2Etxt%2Egz).
- **GPL6887_MouseWG-6_V2_0_R3_11278593_A.txt** the probe annotation needed for the datarow to probe to gene matching. From the GEO deposit inside the [GSE40566_RAW.tar](http://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE40566&format=file).

Reading the data.
```{r}
# the output from Illumina software.
rawdata = read.table("not_in_github/GSE40566_non_normalized.txt", sep="\t", header=TRUE, stringsAsFactors=FALSE)

# the probe annotation file found inside GSE40566_RAW.tar
rawannotation = read.table("not_in_github/GPL6887_MouseWG-6_V2_0_R3_11278593_A.txt", sep="\t", header=TRUE, stringsAsFactors=FALSE, skip=8, comment.char="", quote="", fill=TRUE)
```

Some formatting is needed to match the probe information to the measurments and the measurments to the sample information.
```{r}
# annotation file has two separate tables, one for experimental and one at the end for controls.
# splitting and fixing the tables:
tmp =  rawannotation$Species=="Mus musculus"
experimentalannot = rawannotation[tmp,]
experimentalannot$Array_Address_Id = as.numeric(experimentalannot$Array_Address_Id)
controlannot = rawannotation[!tmp,]
dimnames(controlannot)[[2]] = rawannotation[rawannotation[,2]=="Array_Address_Id",]
controlannot$Array_Address_Id = suppressWarnings(as.numeric(controlannot$Array_Address_Id))
controlannot = controlannot[!is.na(controlannot$Array_Address_Id),]
controlannot=controlannot[,1:6]
probeannotation = merge(experimentalannot, controlannot, all=TRUE )
#dim(probeannotation)
rm(tmp, experimentalannot, controlannot)

probeannotation = probeannotation[!duplicated(probeannotation$Array_Address_Id),]
probeannotation = probeannotation[probeannotation$Array_Address_Id %in% rawdata$ID_REF, ] # 
probeannotation$Symbol=tolower(probeannotation$Symbol)
dimnames(probeannotation)[[1]] = probeannotation$Probe_Id
#dim(probeannotation)

#sort and filter probe and data similar.
datamatrix_raw = as.matrix(rawdata[,-1])
datamatrix_raw = datamatrix_raw[match( probeannotation$Array_Address_Id , rawdata$ID_REF), ]
dimnames(datamatrix_raw)[[1]] = probeannotation$Probe_Id
#dim(datamatrix_raw)
#dim(probeannotation)

#and match data to samples.
#table(sampleannotation$code %in% dimnames(datamatrix_raw)[[2]])# check
#table(dimnames(datamatrix_raw)[[2]] %in% sampleannotation$code)# check
datamatrix_raw = datamatrix_raw[, match(sampleannotation$code , dimnames(datamatrix_raw)[[2]])]
```

Several tables with results are presented in the Supporting Information in Towfic et al. We are aiming to reproduce two of these. Unfortunatly they were only presented in a pdf-format not easely parsed by a computer. We had to resort to a ad-hoc method of cut-and-paste from pdf into text files which were somewhat polluted by pdf-formatting code. For some tables about 10% of the rows will be lost, but the rest will suffice for our purpose.

```{r}
# Table S2
table_s2 = read.table("data/table_s2.csv", sep="\t", header=TRUE, stringsAsFactors=FALSE, fill=TRUE)
table_s2 = as.matrix(table_s2)
table_s2[,3:6] = as.numeric(table_s2[,3:6])
# not able to paste the pdf whitout a lot of gibberish clutter the data. Some probes are lost!
a =rowSums(is.na(table_s2[,3:6])) > 0
print(paste("Lost rows table s2: ", sum(a)))
table_s2=data.frame(table_s2[!a,], stringsAsFactors=FALSE)
# data.frame makes the columns characters again.
for(n in 3:6)
{
  table_s2[,n]=as.numeric(table_s2[,n])
}

# Table S5
table_s5 = read.table("data/table_s5.csv", sep="\t", header=TRUE, stringsAsFactors=FALSE, fill=TRUE)
tmp = as.matrix(table_s5)
tmp[,3:14] = as.numeric(tmp[,3:14])
# not able to paste the pdf whitout a lot of gibberish clutter the data. Some probes are lost!
a =rowSums(is.na(tmp[,3:14])) > 0
print(paste("Lost rows table s5: ", sum(a)))
table_s5=data.frame(table_s5[!a,], stringsAsFactors=FALSE)
table_s5$Fold_Change = as.numeric(table_s5$Fold_Change)
```

### Reproduce some of the original results

Towfic et al. performed many different test on the data and it proved difficult and to labouros to reproduce all of their results without any source code. Insted we foucus on the key part of testing for differentially expressed genes between "GA" (DP) and "generic"(N) as described in Table S2 and Table S5. But before those tests we have to pre-process according to the description.

> Starting with background-corrected bead-level signals, we quantile normalized the extracted data for all samples across all 46,547 probes via the “preprocessCore” package in R.

```{r}
datamatrices = list()
datamatrices[["real_raw"]] = datamatrix_raw#[1:1000,]

# used in paper, but seems to do the same as the limma version. Except it loses dimnames
datamatrices[["real_qnorm"]] = normalize.quantiles(datamatrices[["real_raw"]]) 
dimnames(datamatrices[["real_qnorm"]]) = dimnames(datamatrices[["real_raw"]])
```

> We then corrected for batch variation with ComBat [17] as implemented in the SVA package of R [18]. Each microarray’s chip designation was supplied a batch label; there were 18 batches in all. The labels for the treatments (i.e. drug product, reference standard…) were added as covariates.

```{r}
combatmod = model.matrix(~as.factor(sampleannotation$covariate))
datamatrices[["real_combat_covariates"]]= as.matrix(ComBat(dat=datamatrices[["real_qnorm"]],
                                                           batch=sampleannotation$chip,
                                                           mod=combatmod,
                                                           numCovs=NULL,
                                                           par.prior=TRUE,
                                                           prior.plots=FALSE))

```

First we will try to reproduce Table S2.

> Genes utilized for the tolerance method illustrated in Figure 1B. (figure text)

> This standard of comparison was constructed by first identifying the top 1000 probes by absolute fold change of reference standard compared to the medium (Table S2). The list includes both upregulated and downregulated probes compared to medium. Probes were filtered such that ones upregulated by reference standard needed to have an average reference standard expression of 6.00 or higher and ones downregulated by reference standard needed to have an average medium expression of 6.00 or higher.

This is how the table looks
```{r}
head(table_s2)
```

**Table S2**
```{r}
thiscovariatename = "covariate"
thisdata = datamatrices[["real_combat_covariates"]]

# the mean uses log2 numbers and will be a geometric mean.
meanM = rowMeans(thisdata[, sampleannotation[, "covariate"]=="M"]) 
meanRS = rowMeans(thisdata[, sampleannotation[, "covariate"]=="RS"]) 
meanDP = rowMeans(thisdata[, sampleannotation[, "covariate"]=="DP"])
meanN = rowMeans(thisdata[, sampleannotation[, "covariate"] =="N"])

foldchange_RS_vs_M = meanRS - meanM
foldchange_RS_vs_M = foldchange_RS_vs_M[meanM>6 | meanRS>6]
top1000_RS_vs_M = names(foldchange_RS_vs_M[order(abs(foldchange_RS_vs_M), decreasing=TRUE)])[1:1000]
table(table_s2$ID %in% top1000_RS_vs_M) #  851 out of 909. For some reason it is not a 100% match.
```
Out of Table S2's **`r nrow(table_s2)`** probes, **`r sum(table_s2$ID %in% top1000_RS_vs_M)`** were reproduced. Scatter plots of the actual values shows a correlation.

```{r reproduced_tables2, dev='png', fig.width=8, fig.height=8, fig.show='hold', tidy=FALSE}
# scatter plot correlation of the 4 columns in table_S2 vs. the reproduced numbers. String correlation.
par(mfrow=c(2, 2))
plot(table_s2$AVG_Medium, meanM[table_s2$ID], pch=".")
plot(table_s2$AVG_Reference_Standard, meanRS[table_s2$ID], pch=".")
plot(table_s2$AVG_GA, meanDP[table_s2$ID], pch=".")
plot(table_s2$AVG_generic, meanN[table_s2$ID], pch=".")
```
The numbers in Table S2 is almost reproduced.

**Table S5**
> Comparison of expression in GA to expression in GA for each probe, including fold change, ANOVA, LIMMA with background subtraction, comparative marker selection by signal-to-noise ratio, comparative marker selection by t-test, and the Wilcoxon non-parametric method. <cite> Figure text, Towfic et al.

There must be a typo in the figure text, they ment "Comparison of expression in GA to expression in **generic** for each probe, ...".

> To find differentially expressed probes between generic and GA, we utilized various statistical tests at the probe level and merged the results across the different methods. <cite> Towfic et al.

How the fold change was calculated is not explained. A common way is to take the difference of geometric means that were calculated for Table S2 above.
```{r dev='png', fig.width=8, fig.height=8, fig.show='hold', tidy=FALSE}
foldchange_DP_N = meanDP - meanN

 # make the axis equal.
lim =c( min(   c(min(table_s5$Fold_Change), min(foldchange_DP_N[table_s5$Probe]))  ), 
        max( c(max(table_s5$Fold_Change), max(foldchange_DP_N[table_s5$Probe])) ))

# plot
plot(table_s5$Fold_Change, foldchange_DP_N[table_s5$Probe],
     ylim=lim, xlim=lim, pch=".", xlab="Table_S5", ylab="reproduced",
     main="FC comparison")
rm(lim)
```
The plot reveals that the reproduced fold changes are much more compressed than the ones in Table S5.
There seems to two linear relationship, one for neagtive FC's and another for positve FC's. They seems to have different intercepts but comparable slope.

The average expression for "GA" (DP) and "generic" (N) are also given in Table S2. The fold change calculated from those numbers compared to the ones given in Table S5 shows a similar pattern:

```{r dev='png', fig.width=8, fig.height=8, fig.show='hold', tidy=FALSE}
# table S2 has the AVG for both DP and N listed
fc_table_s2 = table_s2$AVG_GA - table_s2$AVG_generic
names(fc_table_s2) = table_s2$ID
x = table_s5$Fold_Change
y = fc_table_s2[table_s5$Probe]
lim =c( min(   c(min(x, na.rm=T), min(y, na.rm=T))  ), max( c(max(x, na.rm=T), max(y, na.rm=T)) ))
plot(x, y , ylim=lim, xlim=lim, pch="." ,  xlab="Table_S5", ylab="Table_S2",
     main="FC comparison, given values.") #  almost perfect match with offset.
rm(x, y, lim)
```

```{r dev='png', fig.width=10, fig.height=5, fig.show='hold', tidy=FALSE}
par(mfrow=c(1, 2))
plot(table_s5$Fold_Change, foldchange_DP_N[table_s5$Probe],
     pch=".", xlab="Original Table S5", ylab="Reproduced",
     main="FC comparison")

thiscolors = c("black", "red")
hist(foldchange_DP_N, breaks=100, freq=F, border=thiscolors[1], main="Reproduced FC for DP vs. N")
hist(foldchange_DP_N[table_s5$Probe], breaks=100, add=T, freq=F, border=thiscolors[2])
legend("topright", legend=c("All probes", "Table S5 probes"), text.col=thiscolors)
```

The reproduced fold changes for the probes listed in Table S5 are clearly more extreme than the rest.

The next test is a test for differentially expressed genes using ANOVA.
>  First, we computed the statistical significance of differential expression between treatments using the Analysis of Variance (ANOVA) method for each probe [21]...<cite> Towfic et al.

```{r}
anova_p = apply(datamatrices[["real_combat_covariates"]], 1,
              function(x){ summary(aov(x[sampleannotation$covariate %in% c("DP", "N")]~
                                         sampleannotation$covariate[sampleannotation$covariate %in% c("DP", "N")]))[[1]][1,5]})

par(mfrow=c(1, 2))
plot(table_s5[,"anova_p"], anova_p[table_s5$Probe], xlab="Original Table S5", 
     ylab="Reproduced", main="ANOVA P-values from table S5 vs. reproduced")
thiscolors = c("black", "red")
hist(anova_p, breaks=100, freq=T, border=thiscolors[1], main="Reproduced ANOVA p-values for DP vs. N")
hist(anova_p[table_s5$Probe], breaks=100, add=T, freq=T, border=thiscolors[2])
legend("topright", legend=c("All probes", "Table S5 probes"), text.col=thiscolors)
```

> Next, we utilized Linear Models for Microarray (LIMMA) data analysis [23], [24] R package, part of the Bioconductor framework [25], to compare generic and GA samples, fitting a linear model that adjusts for fixed effect from medium (Effect = (GA-generic) – (generic-Medium)). The coefficients for the linear model were tested for significance using a modified t-test (taking into account standard deviation)...<cite> Towfic et al.

The formula "Effect = (GA-generic) – (generic-Medium)" seems strange.

```{r dev='png', fig.width=10, fig.height=6, fig.show='hold', tidy=FALSE}
group = factor(sampleannotation$covariate)
design = model.matrix(~0 + group)
fit = lmFit(datamatrices[["real_combat_covariates"]], design)
cont.matrix = makeContrasts ( contrasts="(groupDP-groupN) - (groupN-groupM)", levels=design)  
fit2 = contrasts.fit(fit, cont.matrix)
limma_ret_org = eBayes(fit2)
limma_p_org = limma_ret_org$p.value
names(limma_p_org) = names(limma_ret_org$Amean)
par(mfrow=c(1, 2))
plot(table_s5[,"limma_p"], limma_p_org[table_s5$Probe], xlab="Original Table S5", 
     ylab="Reproduced", main="LIMMA P-values from table S5 vs. reproduced")
thiscolors = c("black", "red")
hist(limma_p_org, breaks=100, freq=T, border=thiscolors[1], main="Reproduced LIMMA p-values for DP vs. N")
hist(limma_p_org[table_s5$Probe], breaks=100, add=T, freq=T, border=thiscolors[2])
legend("topright", legend=c("All probes", "Table S5 probes"), text.col=thiscolors)
```
A lot of the reproduced p-values are high for the probes in Table S5. A more straightforward contrast of Effect = GA-generic give more similar p-values as in Table S5, and for the rest of this report this is what will be used.

```{r dev='png', fig.width=10, fig.height=6, fig.show='hold', tidy=FALSE}
cont.matrix = makeContrasts ( contrasts="groupDP-groupN", levels=design)  
fit2 = contrasts.fit(fit, cont.matrix)
limma_ret_alt = eBayes(fit2)
limma_p_alt = limma_ret_alt$p.value
names(limma_p_alt) = names(limma_ret_alt$Amean)
par(mfrow=c(1, 2))
plot(table_s5[,"limma_p"], limma_p_alt[table_s5$Probe], xlab="Original Table S5", 
     ylab="Reproduced", main="LIMMA P-values from table S5 vs. reproduced")
thiscolors = c("black", "red")
hist(limma_p_alt, breaks=100, freq=T, border=thiscolors[1], main="Reproduced LIMMA p-values for DP vs. N")
hist(limma_p_alt[table_s5$Probe], breaks=100, add=T, freq=T, border=thiscolors[2])
legend("topright", legend=c("All probes", "Table S5 probes"), text.col=thiscolors)
```
> In parallel, we used Comparative Marker Selection as implemented in GenePattern [26] to directly compare probes between generic and GA. We applied two separate techniques within this framework; a traditional T-test and a Signal-to-Noise Ratio test (SNR).

The T-test is stright forward.
```{r dev='png', fig.width=10, fig.height=6, fig.show='hold', tidy=FALSE}
ttest_p = apply(datamatrices[["real_combat_covariates"]], 1, 
                function(x) { t.test(x[sampleannotation$covariate=="DP"], 
                                     x[sampleannotation$covariate=="N"])$p.value } )
par(mfrow=c(1, 2))
plot(table_s5[,"t_test_p"], ttest_p[table_s5$Probe], xlab="Original Table S5", 
     ylab="Reproduced", main="t-test P-values from table S5 vs. reproduced")
thiscolors = c("black", "red")
hist(ttest_p, breaks=100, freq=T, border=thiscolors[1], main="Reproduced t-test p-values for DP vs. N")
hist(ttest_p[table_s5$Probe], breaks=100, add=T, freq=T, border=thiscolors[2])
legend("topright", legend=c("All probes", "Table S5 probes"), text.col=thiscolors)
```

> ...we used the Wilcoxon Rank Sum Test [27] as implemented in R ...

```{r dev='png', fig.width=10, fig.height=6, fig.show='hold', tidy=FALSE}
# wilcox.test gives a few hundred warnings for ties:
# Warning: cannot compute exact p-value with ties
wilcoxon_p = suppressWarnings(
  apply(datamatrices[["real_combat_covariates"]], 1, 
        function(x) { wilcox.test(x[sampleannotation$covariate=="DP"],
                                  x[sampleannotation$covariate=="N"])$p.value } ))
par(mfrow=c(1, 2))
plot(table_s5[,"wilcoxon_p"], wilcoxon_p[table_s5$Probe], xlab="Original Table S5", 
     ylab="Reproduced", main="wilcoxon P-values from table S5 vs. reproduced")
thiscolors = c("black", "red")
hist(wilcoxon_p, breaks=100, freq=T, border=thiscolors[1], 
     main="Reproduced wilcoxon p-values for DP vs. N")
hist(wilcoxon_p[table_s5$Probe], breaks=100, add=T, freq=T, border=thiscolors[2])
legend("topright", legend=c("All probes", "Table S5 probes"), text.col=thiscolors)

```

### Analysis without ComBat and consequences on the results

Table S2 has average expression values for the conditions, but includes no values about how much trust we can have in those average expressions (i.e significance tests, confidence intervals etc.).

Table S5 on the other hand consists of results from significance tests that are strongly influenced by ComBat's data transformation. An alternative way of handling batch effect compared to ComBat's is to include it in the statistical test. This can be done in ANOVA.

```{r tidy=FALSE}

DP = sampleannotation$covariate=="DP"
N = sampleannotation$covariate=="N"
thisdata =datamatrices[["real_qnorm"]]
cov_fac = factor(sampleannotation$covariate[DP|N])
chip_fac = factor(sampleannotation$chip[DP|N])
#anova_p1 = apply(thisdata, 1, function(x){ summary(aov(x[DP|N]~chip_fac*cov_fac))[[1]][2,5]}  ) 
anova_woc = apply(thisdata, 1, function(x){ summary(aov(x[DP|N]~chip_fac+cov_fac))[[1]][2,5]}  ) 
#co = list(cov_fac=contr.sum, chip_fac=contr.sum)
#anova_p3 = apply(thisdata, 1, function(x){ Anova(lm(x[DP|N] ~  cov1_fac + chip_fac,  contrasts=co), type=3)[[4]][2]}  ) 
table(p.adjust(anova_p , method="fdr")<0.05, p.adjust(anova_woc , method="fdr")<0.05, dnn=c("ComBat adj", "ANOVA adj"))
```
The number of significant probes using the given FDR threshold of 0.05 when batch effect is handled by ANOVA is 
**`r sum(p.adjust(anova_woc , method="fdr")<0.05)`**. This is much lower compared to when ComBat is applied, 
**`r sum(p.adjust(anova_p , method="fdr")<0.05)`**.  The p-value distribution is still skewed, but not as much.
```{r dev='png', fig.width=8, fig.height=8, fig.show='hold', tidy=FALSE}
thiscolors = c("black", "red")
hist(anova_p, breaks=100, freq=T, border=thiscolors[1], main="Reproduced ANOVA p-values for DP vs. N")
hist(anova_woc, breaks=100, add=T, freq=T, border=thiscolors[2])
legend("topright", legend=c("ComBat adjusted", "ANOVA adjusted"), text.col=thiscolors)
```

LIMMA also has the possibility to handle batch inside the statistical test.
```{r dev='png', fig.width=8, fig.height=8, fig.show='hold', tidy=FALSE}
group = factor(sampleannotation$covariate)
block = as.factor(sampleannotation$chip)
design = model.matrix(~0+group+block)
fit = lmFit(thisdata, design)
cont.matrix = makeContrasts ( contrasts="groupDP-groupN", levels=design)  
fit2 = contrasts.fit(fit, cont.matrix)
limma_ret_woc = eBayes(fit2)
limma_p_woc = limma_ret_woc$p.value
names(limma_p_woc) = names(limma_ret_woc$Amean)
rm(design, group, block, fit, cont.matrix, fit2)
table(p.adjust(limma_p_alt , method="fdr")<0.05, p.adjust(limma_p_woc , method="fdr")<0.05, dnn=c("ComBat adj", "LIMMA adj"))
```
The number of significant probes using the given FDR threshold of 0.05 when batch effect is handled by LIMMA is 
**`r sum(p.adjust(limma_p_woc , method="fdr")<0.05)`**. This is much lower compared to when ComBat is applied, 
**`r sum(p.adjust(limma_p_alt , method="fdr")<0.05)`**.  The p-value distribution is still skewed, but not as much.
```{r dev='png', fig.width=8, fig.height=8, fig.show='hold', tidy=FALSE}
hist(limma_p_alt, breaks=100, freq=T, border=thiscolors[1], main="Reproduced ANOVA p-values for DP vs. N")
hist(limma_p_woc, breaks=100, add=T, freq=T, border=thiscolors[2])
legend("topright", legend=c("ComBat adjusted", "LIMMA adjusted"), text.col=thiscolors)
```

The T-test can not handle batch inside the test. An alternative way of performing a t-test is to subset the data and do the comparison whithin each batch. For each of the 7 batches that have both a DP ("GA") and a N ("generic") sample, we can do a direct comparison and in effect ending up with 7 samples. Only 7 samples will give less power but the caveats because of batch effect and adjustments are also far less.
```{r tidy=FALSE, results='hold'}

# subsetting the samples so that they have a balanced conditon distribution across batches. i.e each batch has the same compitition of conditions (example 70% treated vs 30% untreated)
drawbatchbalanceddsamples = function(samplenames, batches, conditions)
{
  # 1. find the minimal count for each condition across batches.
	# 2 find the minimal given count / minimal count inside each batch
	# 3 If greater than 1, scale up all conditions for that batch.
	# 4 draw samples randomly from the above calculated condition-count for each batch
	# 5 create and return the data subset.

	#1
	starttab = table( data.frame(conditions, batches ))	
	tab = starttab[, apply(starttab, 2,  min)>0] # do not use batches with missing conditions
	if(dim(tab)[2]==0)
	{
		warning(  paste("Unable to find a balanced subset that has at least one of each condition. Remove one or more conditions and try again")) 
		return(vector())
	}
	if(dim(tab)[1]<dim(starttab)[1])
	{
		warning(  paste("Lost batches during balancing due to missing conditions in those. ",   "Tried: ", paste( dimnames(starttab)[[2]], collapse=" "), "Got:", paste( dimnames(tab)[[2]], collapse=" "  ) ))
	}

	mincond = apply(tab, 1,  min)

	#2
	tab2 = mincond/tab

	#3
	scalings = apply( 1/tab2, 2, min)
	scaledtab = tab
	scaledtab[,] = mincond	
	scaledtab = t(t(scaledtab) * scalings)
	scaledtab = round(scaledtab,0)

	winnersamples= vector()
	for(c in dimnames(scaledtab)[[1]])
	{
		for(b in dimnames(scaledtab)[[2]])
		{
			a = batches==b & conditions==c
			winnersamples = c( winnersamples, sample( samplenames[a], scaledtab[c,b]))
		}
	}
	return(winnersamples)	
}

a = sampleannotation$covariate %in% c("N", "DP")
nbalanceddatasets = 3
pairedpvals = list()
for(i in 1:nbalanceddatasets)
{
  bs = drawbatchbalanceddsamples(sampleannotation$code[a], sampleannotation$chip[a], sampleannotation$covariate[a])
  print(table(sampleannotation[bs,c("chip", "covariate")])) 
  bs = bs[order(sampleannotation[bs, "covariate"], sampleannotation[bs, "chip"])]
  sampleannotation[bs, c("covariate", "chip")] # the samples are ordered
  thispaireddata = datamatrices[["real_qnorm"]][,bs][,1:7] -  datamatrices[["real_qnorm"]][,bs][,8:14]
  pairedpvals[[i]] = apply(thispaireddata, 1, 
                function(x) { t.test( x)$p.value } )
  print(table(p.adjust(pairedpvals[[i]] , method="fdr")<0.05))
  rm(bs,thispaireddata)  
}
```

```{r dev='png', fig.width=8, fig.height=8, fig.show='hold', tidy=FALSE}
par(mfrow=c(2, 2))
hist(ttest_p, breaks=100, freq=T, border="black", main="P-values t-test on ComBat adjusted data")
for(i in 1:length(pairedpvals))
{
  hist(pairedpvals[[i]], breaks=100, freq=T, border="black", main=paste("P-values paired t-test draw ", i, sep=""))
}

```

### Additional sanity checks

```{r}
set.seed(100)
datamatrices[["random_raw"]] = matrix(rnorm(length(datamatrices[["real_raw"]])), 
                                      nrow=nrow(datamatrices[["real_raw"]]), 
                                      ncol=ncol(datamatrices[["real_raw"]]), 
                                      dimnames=dimnames(datamatrices[["real_raw"]]))
datamatrices[["random_combat_covariates"]]= as.matrix(ComBat(dat=datamatrices[["random_raw"]],
                                                             batch=sampleannotation$chip, 
                                                             mod=combatmod, 
                                                             numCovs=NULL, 
                                                             par.prior=TRUE, 
                                                             prior.plots=FALSE))
```

```{r}
group = factor(sampleannotation$covariate)
design = model.matrix(~0 + group)
fit = lmFit(datamatrices[["random_combat_covariates"]], design)
cont.matrix = makeContrasts ( contrasts="groupDP-groupN", levels=design)  
fit2 = contrasts.fit(fit, cont.matrix)
limma_ret_rand = eBayes(fit2)
limma_p_rand = limma_ret_rand$p.value
names(limma_p_rand) = names(limma_ret_rand$Amean)
table(p.adjust(limma_p_rand, method="fdr")<0.05)
```

```{r dev='png', fig.width=8, fig.height=8, fig.show='hold', tidy=FALSE}
par(mfrow=c(1, 1))
thiscolors = c("black", "red")
hist(limma_p_alt, breaks=100, freq=T, border=thiscolors[1], main="LIMMA p-values for DP vs. N")
hist(limma_p_rand, breaks=100, add=T, freq=T, border=thiscolors[2])
legend("topright", legend=c("Reproduced from Real data", "Random data ComBat adjusted"), text.col=thiscolors)
```

```{r}

anova_p_rand = apply(datamatrices[["random_combat_covariates"]], 1,
              function(x){ summary(aov(x[sampleannotation$covariate %in% c("DP", "N")]~
                                         sampleannotation$covariate[sampleannotation$covariate %in% c("DP", "N")]))[[1]][1,5]})
table(p.adjust(anova_p_rand, method="fdr")<0.05)
```

```{r dev='png', fig.width=8, fig.height=8, fig.show='hold', tidy=FALSE}
par(mfrow=c(1, 1))
thiscolors = c("black", "red")
hist(anova_p, breaks=100, freq=T, border=thiscolors[1], main="ANOVA p-values for DP vs. N")
hist(anova_p_rand, breaks=100, add=T, freq=T, border=thiscolors[2])
legend("topright", legend=c("Reproduced from Real data", "Random data ComBat adjusted"), text.col=thiscolors)
```

The p-value plots from the ComBat adjusted random data shows a skewed distribution, although not as much as for the real data. The random data has no batch effect or other effects like the real data which might affect the ComBat adjustment.

Another sanity check is by permuting the DP and N labels and look how the result changes. In order to retain the batch effect we will only swap labels within a batch. This should in theory give fewer significant probes.

```{r}
shufflesamplesinbatch = function(samplenames, batch, covariate, shufflecovariates)
{
  ret = samplenames
	b = covariate %in% shufflecovariates
	for(thisbatch in unique(batch))
	{
		a = batch== thisbatch		
		ret[a & b] = sample( samplenames[a&b],sum(a&b),  replace=FALSE)		
	}
	return(ret)
}

nshuffleddatasets = 10
datamatrices_permuted = list()
for(i in 1:nshuffleddatasets)
{
  x = shufflesamplesinbatch( sampleannotation$code, sampleannotation$chip, sampleannotation$covariate, c("N", "DP") )
  print(  sum(sampleannotation[,"covariate"] != sampleannotation[x,"covariate"]  ))  # see how many got change. Has 11 N's)  
  datamatrices_permuted[[i]] = as.matrix(ComBat(dat=datamatrices[["real_qnorm"]][, x], 
                                                batch=sampleannotation$chip, 
                                                mod=combatmod, 
                                                numCovs=NULL, 
                                                par.prior=TRUE, 
                                                prior.plots=FALSE))
}

permuted_pvalcounts=list()
for(i in 1:nshuffleddatasets)
{ 
  group = factor(sampleannotation$covariate)
  design = model.matrix(~0 + group)
  fit = lmFit(datamatrices_permuted[[i]], design)
  cont.matrix = makeContrasts ( contrasts="groupDP-groupN", levels=design)  
  fit2 = contrasts.fit(fit, cont.matrix)  
  this_p = eBayes(fit2)$p.value
  permuted_pvalcounts[[i]] = hist(this_p, plot=FALSE, breaks=100)$counts
  print(table(p.adjust(this_p, method="fdr")<0.05))
}
```

```{r, dev='svg', fig.width=8, fig.height=8, tidy=FALSE}
real_pvalcounts=hist(limma_p_alt, plot=FALSE, breaks=100)$counts
plot((1:20)/100, real_pvalcounts[1:20], col="red",
     main="P-values, permuted real data", 
      xlab="p-value", ylab="Frequency", type="l", lwd=2, 
      ylim=c(0, max(unlist( c(permuted_pvalcounts,real_pvalcounts)))))
for(i in 1:length(permuted_pvalcounts))
{
  lines((1:20)/100,permuted_pvalcounts[[i]][1:20], col="blue")
}
legend("topright", legend=c("Reproduced p-values real labels", "Permuted DP and N labels"), text.col=c("red", "blue"))
```
### References

Johnson, WE, Rabinovic, A, and Li, C (2007). Adjusting batch effects in microarray expression data using Empirical Bayes methods. Biostatistics 8(1):118-127.

Leek JT, Johnson WE, Parker HS, Jaffe AE, Storey JD.(2012) The sva package for removing batch effects and other unwanted variation in high-throughput experiments. Bioinformatics. 2012 Mar 15;28(6):882-3.

Smyth, GK (2005). Limma: linear models for microarray data. In: 'Bioinformatics and Computational Biology Solutions
  using R and Bioconductor'. R. Gentleman, V. Carey, S. Dudoit, R. Irizarry, W. Huber (eds), Springer, New York, pages
  397-420.

  R Core Team (2013). R: A language and environment for statistical computing. R Foundation for Statistical Computing,
  Vienna, Austria. URL http://www.R-project.org/

  Yihui Xie (2013). knitr: A general-purpose package for dynamic report generation in R. R package version 1.5.

  Yihui Xie (2013) Dynamic Documents with R and knitr. Chapman and Hall/CRC. ISBN 978-1482203530

  Yihui Xie (2013) knitr: A Comprehensive Tool for Reproducible Research in R. In Victoria Stodden, Friedrich Leisch and
  Roger D. Peng, editors, Implementing Reproducible Computational Research. Chapman and Hall/CRC. ISBN 978-1466561595

  RStudio Team (2012). RStudio: Integrated Development for R. RStudio, Inc., Boston, MA URL http://www.rstudio.com/.

```{r sessionInfo, comment=""}
sessionInfo()
```

generation ended `r as.character(Sys.time())`. Time spent `r  as.integer(round(difftime(Sys.time(),starttime, units="mins")))` minutes .

### 2014-03-10 docs/article.rmd
mer

Methods that remove batch effects while retaining group differences may induce false confidence in downstream analyses
========================================================

## Abstract
Removal of, or adjustment for, batch effects or centre differences is generally required when such effects are present in data. In particular, when preparing microarray gene expression data from multiple cohorts, array platforms, or batches for later analyses, batch effects can have confounding effects. Many methods and tools exist for this purpose. One method, ComBat which is part of the R package sva, is particularly popular due to its ability to remove batch differences even when batches are small and heterogeneous, while retaining the difference between study groups, e.g. treatments. Unfortunately, this recommended and frequently used approach to batch correction will systematically induce incorrect group differences in downstream analyses when groups are unevenly distributed between the batches. The scientific community seems to be largely unaware of this problem, which most likely has contributed to false discoveries being presented in the published literature.

## Introduction

```{r}
"
Kort oppsummere behovet for batch-korrigering: generelt og i microarraydata.

Kort beskrive metoder som gjoer dette (batch-sentrering, ANOVA, ComBat).

Forklare tilsiktet effekt disse paa batch+gruppe-forskjeller (balansert versus ikke-balansert).

Forklare kort hva faktisk effekt er: ANOVA gir redusert gruppe-forskjeller, mens for ComBat overestimeres den statistiske styrken/sikkerheten av forskjellen.
"
```

## Simple sanity check
1. Vise til sanity checken med random numbers paa eksempelet fra sva
pakka (bladderbatch) med p-verdi-histogram og mange DE-gener.

## Formler og ord
2. Forklare hva som gaar galt med formler og ord.

## Examples of Undesired consequences
As the amount of false positive results as a consequence of trying to retain group differences depends on the group/batch balance, we will show two examples with varying degree of unbalancedness. The first is a "worst-case"scenario where a few batches are completly missing groups coupled with a a priori assumption of no group difference (24421904). The second compares different cell types and is only mildly unbalanced ("data set 2" from 16632515). We tested their most highlighted group difference for differentially expressed genes using the described approach with ComBat and an alternative approach handling batch inside the test. In addition we performed one sanity check permuting the group labels whitin batches, and one sanity check using random numbers. The different p-value distributions are shown in Figure 2.

![fig. 1](../reanalysis/ComBat_original/figure/pvaluesjohnson.svg)

```{r}
"
3. Vise at det er variasjon i stoerrelsen paa feilen introdusert av ComBat
ved a)worst-case scenario (MS-artikel)og b) mild-case (data set 2 fra
ComBat). Vises med p-verdi plot med/uten ComBat og eventuelt antall
signifikante prober.
"
```
## Discussion

## Sliding application area of ComBat
In the original ComBat article(2007) it is clear that the primary motivation behind ComBat was to employ an Emprical Bayes method for batch-effect removal. The feature of retaining group differences for unbalanced designs is optional and seems to be subordinate, only examplified in the supplementary information. However, over the years this feature became more important judging from advice given by the author to users
(
[1](https://groups.google.com/d/msg/combat-user-forum/eSVSKwGtuyE/ZIWV2juYmmAJ)
,[2](https://groups.google.com/d/msg/combat-user-forum/Vkb9p7wekd4/h5Etie7FTVQJ)
,[3](https://groups.google.com/d/msg/combat-user-forum/fpBTcgDjiR8/uo4QIZL4sZgJ)
,[4](https://groups.google.com/d/msg/combat-user-forum/26FZlgU2LFQ/W6U_Lhh_64EJ)
).
And when ComBat was incorporated in the sva package ( [22257669](http://bioinformatics.oxfordjournals.org/cgi/pmidlookup?view=long&pmid=22257669), 2012), inclusion of group labels was made almost mandatory (an undocumented option of passing a NULL value exists). Thus, this problematic use has likely been common.

## Motivation for this Warning
Our knowledge of ComBat came through a typical use case when trying to salvage unbalanced data which had batch effects. Upon realizing that the confidence on our group differences were exagerated, the literature was searched for a better understanding of correct use and potential overseen limitations of ComBat. But the authors of ComBat and the sva package recomended our usage (16632515, 22257669). In addition, other works looking into the problem of batch effects were mostly recomending ComBat without much concern (22682473, 22133085). A brief inquiry into  some of the articles citing ComBat (574 Google Scholar) revealed few problems, and their method descriptions regarding ComBat were mostly sparse, limited to one or two sentences (24391845, 18414638, 21731603,23630272, 24584070). A further indication of their carefree use of this potentially devastating procedure was the frequent neglect to state the program parameters, i.e batch-labels (18414638, 21731603) or group labels (24391845, 18414638, 21731603, 23630272, 23482648, 24584070). Often no effort was done in order to substantiate the existance of batch effects in their data, ecxept for stating the presence of batches (24391845?, 18414638, 21731603, 23630272). In one instance ComBat was even applied on data were effects due to batch were investigated, but not found (husker ikke ref). The incorporation of the method into analysis pipelines (16642009,TCGA) and other packages (23452776, 21937664) could make its usage even more trivial and parameters setting harder to perceive.
Taken together we fear that many published results from data adjusted by ComBat are completly or partially false. And knowing that scientist don't give up when one analysis fails, a method that almost ensures a result given a sufficently unbalanced design will continue to be used.

```{r}
"

Mange siteringer
I pipelines 16642009,TCGA  pakker.   23452776, 21937664
recomondations, review vinnere. 22682473, 22133085
mangelfull forstaaelse fra forfattere. 16632515, 22257669
Faa setninger, 24391845, 18414638, 21731603,23630272, 24584070
ikke oppgit parametere, dvs batch(18414638, 21731603) eller covariates(24391845, 18414638, 21731603, 23630272, 23482648, 24584070),
Ikke vist batch problem (24391845?, 18414638, 21731603, 23630272).
Brukt uansett.
P-hacking. Gjoer ComBat populaer.
"
```

## Guide for "worried" scientist
6. Gi en rettledning for hvordan en bekymret forsker kanskje kan
estimere feilen p.g.a ComBat-bruk i en eller flere utvalgte artikler.

likely small or no difference (24391845, 18414638?, 21731603, 23630272)

## Reproducible research

## References

Combat org
http://biostatistics.oxfordjournals.org/content/8/1/118.abstract
Adjusting batch effects in microarray expression data using empirical Bayes methods.
Johnson WE, Li C, Rabinovic A.
Biostatistics. 2007 Jan;8(1):118-27. Epub 2006 Apr 21.
PMID: 16632515 [PubMed - indexed for MEDLINE] Free Article

1
Molecular profiling of single Sca-1+/CD34+,- cells--the putative murine lung stem cells.
Hittinger M, Czyz ZT, Huesemann Y, Maneck M, Botteron C, Kaeufl S, Klein CA, Polzer B.
PLoS One. 2013 Dec 31;8(12):e83917. doi: 10.1371/journal.pone.0083917. eCollection 2013.
PMID: 24391845 [PubMed - in process] 
http://www.plosone.org/article/info%3Adoi%2F10.1371%2Fjournal.pone.0083917
" Based on the hybridization date the data set is separated into two main batches of microarrays (early 2007/2008 and late 2009), which degrade further into small sub-batches. We used the ComBat algorithm [15] to adjust for the main batches. "

2
Drinking-water arsenic exposure modulates gene expression in human lymphocytes from a U.S. population.
Andrew AS, Jewell DA, Mason RA, Whitfield ML, Moore JH, Karagas MR.
Environ Health Perspect. 2008 Apr;116(4):524-31. doi: 10.1289/ehp.10861.
PMID: 18414638 [PubMed - indexed for MEDLINE] Free PMC Article
http://www.ncbi.nlm.nih.gov/pmc/articles/PMC2290973/
"followed by empirical Bayes adjustment procedures implemented on the R platform, as described previously (Johnson et al. 2007)"

3
Epigenetic predictor of age.
Bocklandt S, Lin W, Sehl ME, Sánchez FJ, Sinsheimer JS, Horvath S, Vilain E.
PLoS One. 2011;6(6):e14821. doi: 10.1371/journal.pone.0014821. Epub 2011 Jun 22.
PMID: 21731603 [PubMed - indexed for MEDLINE] Free PMC Article
http://www.plosone.org/article/info%3Adoi%2F10.1371%2Fjournal.pone.0014821
"Batch effect were removed using the Combat algorithm [9], and one outlier sample was removed."

4.
Childhood maltreatment is associated with distinct genomic and epigenetic profiles in posttraumatic stress disorder.
Mehta D, Klengel T, Conneely KN, Smith AK, Altmann A, Pace TW, Rex-Haffner M, Loeschner A, Gonik M, Mercer KB, Bradley B, Müller-Myhsok B, Ressler KJ, Binder EB.
Proc Natl Acad Sci U S A. 2013 May 14;110(20):8302-7. doi: 10.1073/pnas.1217750110. Epub 2013 Apr 29.
PMID: 23630272 [PubMed - indexed for MEDLINE] Free PMC Article
http://www.pnas.org/content/110/20/8302.full
"To correct for confounding as a result of batch effects, the data were normalized using an empirical Bayes method for batch correction " (expression)
"Hybridization and chip batch effects were accounted for using an empirical Bayes method (43)." (methylation)

5.
Patterns of regulatory activity across diverse human cell types predict tissue identity, transcription factor binding, and long-range interactions.
Sheffield NC, Thurman RE, Song L, Safi A, Stamatoyannopoulos JA, Lenhard B, Crawford GE, Furey TS.
Genome Res. 2013 May;23(5):777-88. doi: 10.1101/gr.152140.112. Epub 2013 Mar 12.
PMID: 23482648 [PubMed - indexed for MEDLINE] Free PMC Article
http://genome.cshlp.org/content/23/5/777.full
"We improved on the previously published open chromatin measurements by accounting for batch affects that grouped the data by laboratory rather than by biological signal (see Methods). We used ComBat (Johnson et al. 2007) to remove these batch effects, after which both the DNase I and expression data clustered according to expected biological relationships (Supplemental Fig. S1)."
"Counts were quantile-normalized and scaled, and protocol batch effects were corrected using ComBat (Supplemental Fig. S1; Johnson et al. 2007)."
"Artifactual differences between protocols (Figure S1) were corrected using ComBat (Johnson et al. 2007). Because the data was too large for the ComBat model, we divided it into subsets of about 300,000 DHSs and corrected the batch affect individually on each subset. "
"To make the arrays comparable, we used ComBat to correct for this batch affect (Johnson et al. 2007). After correction with ComBat, the arrays grouped according to expected biological similarity (Figure S1). "

6.
A recurrent inactivating mutation in RHOA GTPase in angioimmunoblastic T cell lymphoma.
Yoo HY, Sung MK, Lee SH, Kim S, Lee H, Park S, Kim SC, Lee B, Rho K, Lee JE, Cho KH, Kim W, Ju H, Kim J, Kim SJ, Kim WS, Lee S, Ko YH.
Nat Genet. 2014 Mar 2. doi: 10.1038/ng.2916. [Epub ahead of print]
PMID: 24584070 [PubMed - as supplied by publisher]
http://www.nature.com/ng/journal/vaop/ncurrent/pdf/ng.2916.pdf
http://www.nature.com/ng/journal/vaop/ncurrent/full/ng.2916.html
http://www.nature.com/ng/journal/vaop/ncurrent/extref/ng.2916-S1.pdf
"We combined the RNA-seq and microarray data using the Combat algorithm36 in the Bioconductor sva package to remove batch biases due to platform differences"

## anbefaler ComBat

10
Batch correction of microarray data substantially improves the identification of genes differentially expressed in rheumatoid arthritis and osteoarthritis.
Kupfer P, Guthke R, Pohlers D, Huber R, Koczan D, Kinne RW.
BMC Med Genomics. 2012 Jun 8;5:23. doi: 10.1186/1755-8794-5-23.
PMID: 22682473 [PubMed - indexed for MEDLINE] Free PMC Article
Related citations
http://www.biomedcentral.com/1755-8794/5/23

11
Relative impact of key sources of systematic noise in Affymetrix and Illumina gene-expression microarray experiments.
Kitchen RR, Sabine VS, Simen AA, Dixon JM, Bartlett JM, Sims AH.
BMC Genomics. 2011 Dec 1;12:589. doi: 10.1186/1471-2164-12-589.
PMID: 22133085 [PubMed - indexed for MEDLINE] Free PMC Article
http://www.biomedcentral.com/1471-2164/12/589/

## skeptisk til ComBat

20 - HAr kjoert permutering av covariatelabler og finner at combat finner masse gener uansett. Fokus paa artikkel er ikke combat spesifikkt.
Expanding the understanding of biases in development of clinical-grade molecular signatures: a case study in acute respiratory viral infections.
Lytkin NI, McVoy L, Weitkamp JH, Aliferis CF, Statnikov A.
PLoS One. 2011;6(6):e20662. doi: 10.1371/journal.pone.0020662. Epub 2011 Jun 1.
PMID: 21673802 [PubMed - indexed for MEDLINE] Free PMC Article
http://www.plosone.org/article/info%3Adoi%2F10.1371%2Fjournal.pone.0020662

## ComBat i pipelines

TCGA Batch Effects Tool
http://bioinformatics.mdanderson.org/main/TCGABatchEffects:Overview

GenePattern 2.0.
Reich M, Liefeld T, Gould J, Lerner J, Tamayo P, Mesirov JP.
Nat Genet. 2006 May;38(5):500-1. No abstract available.
PMID: 16642009 [PubMed - indexed for MEDLINE]
http://www.nature.com/ng/journal/v38/n5/full/ng0506-500.html

Uverifisert.
insilicodb? bioconductopakke eller hva
http://bib.oxfordjournals.org/content/early/2012/07/31/bib.bbs037.short

The sva package for removing batch effects and other unwanted variation in high-throughput experiments.
Leek JT, Johnson WE, Parker HS, Jaffe AE, Storey JD.
Bioinformatics. 2012 Mar 15;28(6):882-3. doi: 10.1093/bioinformatics/bts034. Epub 2012 Jan 17.
PMID: 22257669 [PubMed - indexed for MEDLINE] Free PMC Article
http://bioinformatics.oxfordjournals.org/cgi/pmidlookup?view=long&pmid=22257669

virtualArray: a R/bioconductor package to merge raw data from different microarray platforms.
Heider A, Alt R.
BMC Bioinformatics. 2013 Mar 2;14:75. doi: 10.1186/1471-2105-14-75.
PMID: 23452776 [PubMed - indexed for MEDLINE] Free PMC Article
http://bioconductor.org/packages/release/bioc/html/virtualArray.html

inSilicoDb: an R/Bioconductor package for accessing human Affymetrix expert-curated datasets from GEO.
Taminau J, Steenhoff D, Coletta A, Meganck S, Lazar C, de Schaetzen V, Duque R, Molter C, Bersini H, Nowé A, Weiss Solís DY.
Bioinformatics. 2011 Nov 15;27(22):3204-5. doi: 10.1093/bioinformatics/btr529. Epub 2011 Sep 21.
PMID: 21937664 [PubMed - indexed for MEDLINE] Free Article

## eksempler med reporduserte p-value

Comparing the biological impact of glatiramer acetate with the biological impact of a generic.
Towfic F, Funt JM, Fowler KD, Bakshi S, Blaugrund E, Artyomov MN, Hayden MR, Ladkani D, Schwartz R, Zeskind B.
PLoS One. 2014 Jan 8;9(1):e83757. doi: 10.1371/journal.pone.0083757. eCollection 2014.
PMID: 24421904 [PubMed - in process] Free PMC Article
http://www.plosone.org/article/info%3Adoi%2F10.1371%2Fjournal.pone.0083757

### user group recomandations
https://groups.google.com/d/msg/combat-user-forum/eSVSKwGtuyE/ZIWV2juYmmAJ
https://groups.google.com/d/msg/combat-user-forum/Vkb9p7wekd4/h5Etie7FTVQJ
https://groups.google.com/d/msg/combat-user-forum/fpBTcgDjiR8/uo4QIZL4sZgJ
https://groups.google.com/d/msg/combat-user-forum/26FZlgU2LFQ/W6U_Lhh_64EJ

Alternativer til tittel
Methods that remove batch effects while retaining group differences may lead to incorrect conclusions in downstream analyses
Methods that remove batch effects while retaining group differences may lead to incorrect conclusions

```{r}

```

### 2014-03-10 reanalysis/ComBat_original/combat_paper_dataset2_reproduced.rmd
mer

@@
 EB3_limma_res = eBayes(fit2)
-rm(thismat, Type, Block, design, fit, cont.matrix, fit2)
+rm(thismat, Type, design, fit, cont.matrix, fit2)
 print(table(qvalue(EB3_limma_res$p.value[,1])$qvalue<0.05))
@@
 
+```{r}
+source("../../commonscripts/helper_functions.r")
+nshuffleddatasets = 10
+datamatrices_permuted = list()
+for(i in 1:nshuffleddatasets)
+{
+  x = shufflesamplesinbatch( rownames(sampleannotation), sampleannotation$Batch, sampleannotation$Type, c("C", "R") )
+  print(  sum(sampleannotation[,"Type"] != sampleannotation[x,"Type"]  ))  # see how many changed
+  datamatrices_permuted[[i]] = as.matrix(ComBat(dat=datamatrix[, x], 
+                                                batch=sampleannotation$Batch, 
+                                                mod=model.matrix(~as.factor(sampleannotation[colnames(datamatrix),"Type"])),  
+                                                numCovs=NULL, 
+                                                par.prior=TRUE, 
+                                                prior.plots=FALSE))
+}
+
+permuted_pvalcounts=list()
+for(i in 1:nshuffleddatasets)
+{ 
+  this_p = apply(datamatrices_permuted[[i]], 1 , 
+                  FUN=function(x){t.test(
+                                    x[sampleannotation[, "Type"]=="C"],
+                                    x[sampleannotation[, "Type"]=="R"]
+                                    )$p.value})
+
+  permuted_pvalcounts[[i]] = hist(this_p, plot=FALSE, breaks=100)$counts
+  print(table(p.adjust(this_p, method="fdr")<0.05))
+}
+```
+
+Forslag figur brukt i artikkel.
+```{r pvaluesjohnson, dev='svg', fig.width=10, fig.height=10, tidy=FALSE}
+# figure used in our article
+# NB! bruker orgniale data som ikke er floored eller log2!
+par(mfrow=c(2, 2))
+
+# reproduced ComBat + limma
+hist(EB3_limma_res$p.value[,1], main="P-values, real data, ComBat followed by Limma", breaks=100, xlab="p-value", ylim=c(0,3000))
+
+# batch handled in limma
+hist(datamatrix_limma_res$p.value[,1], breaks=100, main="P-values, real data, batch handled by Limma", xlab="p-value" , ylim=c(0,3000))
+
+# random ComBat + limma
+hist(EBrand_limma_pvalues, main="P-values, Random data, ComBat followed by Limma", breaks=100, xlab="p-value" , ylim=c(0,3000))
+
+# permuted labels whitin batch
+
+### permutation sanity test
+real_pvalcounts=hist(EB3_limma_res$p.value[,1], plot=FALSE, breaks=100)$counts
+plot((1:20)/100, real_pvalcounts[1:20], col="red",
+     main=paste("P-values, real data and ", length(permuted_pvalcounts), " permutations" ,sep=""),
+      xlab="p-value", ylab="Frequency", type="l", lwd=2, 
+      ylim=c(0, max(unlist( c(permuted_pvalcounts,real_pvalcounts)))))
+for(i in 1:length(permuted_pvalcounts))
+{
+  lines((1:20)/100,permuted_pvalcounts[[i]][1:20], col="blue")
+}
+legend("topright", legend=c("Reproduced p-values real labels", "Permuted C and R labels"),
+       text.col=c("red", "blue"))
+```
+
 ### References

### 2014-03-10 reanalysis/Fadi_Towfic_generic_medicine/towfic_et_al.rmd
mer

@@
 datamatrices = list()
-datamatrices[["real_raw"]] = datamatrix_raw#[1:1000,]
+datamatrices[["real_raw"]] = datamatrix_raw#[1:1000,] # debug
 
@@
  # make the axis equal.
-lim =c( min(   c(min(table_s5$Fold_Change), min(foldchange_DP_N[table_s5$Probe]))  ), 
-        max( c(max(table_s5$Fold_Change), max(foldchange_DP_N[table_s5$Probe])) ))
+lim =c( min(   c(min(table_s5$Fold_Change, na.rm=T), min(foldchange_DP_N[table_s5$Probe], na.rm=T))  ), 
+        max( c(max(table_s5$Fold_Change, na.rm=T), max(foldchange_DP_N[table_s5$Probe], na.rm=T)) ))
 
@@
 ```{r}
-shufflesamplesinbatch = function(samplenames, batch, covariate, shufflecovariates)
-{
-  ret = samplenames
-	b = covariate %in% shufflecovariates
-	for(thisbatch in unique(batch))
-	{
-		a = batch== thisbatch		
-		ret[a & b] = sample( samplenames[a&b],sum(a&b),  replace=FALSE)		
-	}
-	return(ret)
-}
+source("../../commonscripts/helper_functions.r")
+#shufflesamplesinbatch = function(samplenames, batch, covariate, shufflecovariates)
+#{
+  #ret = samplenames
+	#b = covariate %in% shufflecovariates
+	#for(thisbatch in unique(batch))
+	#{
+	#	a = batch== thisbatch		
+#		ret[a & b] = sample( samplenames[a&b],sum(a&b),  replace=FALSE)		
+	#}
+	#return(ret)
+#}
 
@@
 ```
+
+Forslag figur brukt i artikkel.
+```{r pvaluestowfic, dev='svg', fig.width=10, fig.height=10, tidy=FALSE}
+# figure used in our article
+# NB! bruker orgniale data som ikke er floored eller log2!
+par(mfrow=c(2, 2))
+
+thisylim = c(0, max(unlist( c(permuted_pvalcounts,real_pvalcounts))))
+# reproduced ComBat + limma
+hist(limma_p_alt, main="P-values, real data, ComBat followed by Limma", breaks=100, xlab="p-value", ylim=thisylim)
+
+# batch handled in limma
+hist(limma_p_woc, breaks=100, main="P-values, real data, batch handled by Limma", xlab="p-value" , ylim=thisylim)
+
+# random ComBat + limma
+hist(limma_p_rand, main="P-values, Random data, ComBat followed by Limma", breaks=100, xlab="p-value" , ylim=thisylim)
+
+# permuted labels whitin batch
+
+### permutation sanity test
+real_pvalcounts=hist(limma_p_alt, plot=FALSE, breaks=100)$counts
+plot((1:20)/100, real_pvalcounts[1:20], col="red",
+     main=paste("P-values, real data and ", length(permuted_pvalcounts), " permutations" ,sep=""),
+      xlab="p-value", ylab="Frequency", type="l", lwd=2, 
+      ylim=thisylim)
+for(i in 1:length(permuted_pvalcounts))
+{
+  lines((1:20)/100,permuted_pvalcounts[[i]][1:20], col="blue")
+}
+legend("topright", legend=c("Reproduced p-values real labels", "Permuted DP and N labels"),
+       text.col=c("red", "blue"))
+```
+
 ### References

### 2014-03-10 commonscripts/helper_functions.r

@@ -285,3 +285,6 @@ shufflesamplesinbatch = function(samplenames, batch, covariate, shufflecovariate
 	}
 	return(ret)
 }
+
+
+

### 2014-03-11 docs/article.rmd
mer

@@
 
-```{r}
+```{r echo=FALSE}
 "
@@
 As the amount of false positive results as a consequence of trying to retain group differences depends on the group/batch balance, we will show two examples with varying degree of unbalancedness.
-The first is a "worst-case"scenario where a few batches are completly missing groups coupled with a a priori assumption of no group difference (24421904).
-The second compares different cell types and is only mildly unbalanced ("data set 2" from 16632515).
-We tested their most highlighted group difference for differentially expressed genes using the described approach with ComBat and an alternative approach handling batch inside the test.
-In addition we performed one sanity check permuting the group labels whitin batches, and one sanity check using random numbers.
-The different p-value distributions are shown in Figure 2.
 
-![fig. 1](../reanalysis/ComBat_original/figure/pvaluesjohnson.svg)
+1) In the first experiment(24421904), cells are treated glatiramer acetate (a medicine for multiple sclerosis) or a generic and mRNA was measured on microarrays alongside control samples.
+A batch effect correlated to the array platform (Illumina WG-6\_V2) was observed and adjusted for with ComBat whereafter the data was tested for differetially expressed genes, yielding a list of 1000 genes (Table S5 24421904).
+Unfortunately the treatment/batch design was unbalanced with several batches having only one of the main treatment of interest.
+When we re-analyzed their data without using ComBat, but instead blocked for batch effect in limma, only 9 genes were found (FDR<0.05).
+Additional sanity checks were also applied and the distribution of p-values for differents settings are shown in Fig.
+2.
+Our conclusion is that most of the genes reported as differtially expresssed in (24421904) are false positives.
+This example is a sort of "worst case" scenario for applying ComBat, since it both has a very unbalanced batch/group design and a a priori assumpotion of no difference.
+The R-code for our analysis and a more extensive report can be downloaded from github (https://github.com/vegardny/combat_tests.git).
 
-```{r}
-"
-3. Vise at det er variasjon i stoerrelsen paa feilen introdusert av ComBat
-ved a)worst-case scenario (MS-artikel)og b) mild-case (data set 2 fra
-ComBat). Vises med p-verdi plot med/uten ComBat og eventuelt antall
-signifikante prober.
-"
-```
+![fig. 2](../reanalysis/Fadi_Towfic_generic_medicine/figure/pvaluestowfic.svg)
+**Figure 3**.
+P-value distribution for the main comparison in (data set 2 24421904) glatiramer acetate vs. generic.  
+- a) Real data adjusted by ComBat as described in the article followed by a significance test in limma. 
+- b) Real data not adjusted by ComBat, tested for significance in limma using batch as a blocking factor. 
+- c) Real data is substituted with random numbers from a normal distribution (mean=0, std=1), but the batch/group design is retained, followed by batch adjustment in ComBat and a significance test in limma.
+- d) 10 runs of real data with the glatiramer acetate("DP") and generic("N") labels permuted within batches, followed by ComBat adjustment and significance test in limma.
+
+2) The second example is taken from the supporting information for the original ComBat article (16632515) where it is denoted "data set 2".
+Cells inhibited for the expression of the TAL1 gene were compared to controls on a microarray platform.
+The experiment was conducted on three different time points (used as batches) with a total of 30 samples with a fairly balanced treatment/control across batches (6/2, 3/4 and 9/6).
+ComBat was applied followed by a t-test identifying differentially expressed genes.
+First, we reproduced their analysis including the adjustment by ComBat, but using limma instead of the t-test, resulting in 1003 probes (q<0.05).
+Then, we analysed their data blocking for batch in limma and not using ComBat, resulting in 377 probes (q<0.05).
+In addition the two sanity checks outline above were performed.
+Comparing the P-value distributions for the alternative analyses (Fig.3 a,b) does not indicate a huge difference, in contrast to results above for(24421904).
+
+![fig. 3](../reanalysis/ComBat_original/figure/pvaluesjohnson.svg)  
+**Figure 3**.
+P-value distribution for the main comparison in (data set 2 16632515) for  
+- a) Real data adjusted by ComBat as described in the article followed by a significance test in limma. 
+- b) Real data not adjusted by ComBat, tested for significance in limma using batch as a blocking factor. 
+- c) Real data is substituted with random numbers from a normal distribution (mean=0, std=1), but the batch/group design is retained, followed by batch adjustment in ComBat and a significance test in limma.
+- d) 10 runs of real data with the group labels permuted within batches, followed by ComBat adjustment and significance test in limma.
+
 ## Discussion
@@
 In the original ComBat article(2007) it is clear that the primary motivation behind ComBat was to employ an Emprical Bayes method for batch-effect removal.
-The feature of retaining group differences for unbalanced designs is optional and seems to be subordinate, only examplified in the supplementary information.
+The feature of retaining group differences for unbalanced designs is optional and seems to be subordinate, only exemplified in the supplementary information.
 However, over the years this feature became more important judging from advice given by the author to users
@@
 Our knowledge of ComBat came through a typical use case when trying to salvage unbalanced data which had batch effects.
-Upon realizing that the confidence on our group differences were exagerated, the literature was searched for a better understanding of correct use and potential overseen limitations of ComBat.
-But the authors of ComBat and the sva package recomended our usage (16632515, 22257669).
-In addition, other works looking into the problem of batch effects were mostly recomending ComBat without much concern (22682473, 22133085).
+Upon realizing that the confidence on our group differences were exaggerated, the literature was searched for a better understanding of correct use and potential overseen limitations of ComBat.
+But the authors of ComBat and the sva package recommended our usage (16632515, 22257669).
+In addition, other works looking into the problem of batch effects were mostly recommending ComBat without much concern (22682473, 22133085).
 A brief inquiry into  some of the articles citing ComBat (574 Google Scholar) revealed few problems, and their method descriptions regarding ComBat were mostly sparse, limited to one or two sentences (24391845, 18414638, 21731603,23630272, 24584070).
 A further indication of their carefree use of this potentially devastating procedure was the frequent neglect to state the program parameters, i.e batch-labels (18414638, 21731603) or group labels (24391845, 18414638, 21731603, 23630272, 23482648, 24584070).
-Often no effort was done in order to substantiate the existance of batch effects in their data, ecxept for stating the presence of batches (24391845?, 18414638, 21731603, 23630272).
+Often no effort was done in order to substantiate the existence of batch effects in their data, except for stating the presence of batches (24391845?, 18414638, 21731603, 23630272).
 In one instance ComBat was even applied on data were effects due to batch were investigated, but not found (husker ikke ref).
 The incorporation of the method into analysis pipelines (16642009,TCGA) and other packages (23452776, 21937664) could make its usage even more trivial and parameters setting harder to perceive.
-Taken together we fear that many published results from data adjusted by ComBat are completly or partially false.
-And knowing that scientist don't give up when one analysis fails, a method that almost ensures a result given a sufficently unbalanced design will continue to be used.
+Taken together we fear that many published results from data adjusted by ComBat are completely or partially false.
+And knowing that scientist don't give up when one analysis fails, a method that almost ensures a result given a sufficiently unbalanced design will continue to be used.
 
-```{r}
+```{r echo=FALSE}
 "

### 2014-03-13 docs/article.rmd
sva

@@
 ## Simple sanity check
-1. Vise til sanity checken med random numbers paa eksempelet fra sva
-pakka (bladderbatch) med p-verdi-histogram og mange DE-gener.
+The undesired consequnces of preserving group effects when correcting for batch effect is readily illustrated with a sanity check using random numbers .
+The documentation accompanying the sva library has a runnable example demonstrating how to adjust a data set with ComBat followed by a F-test.
+Swapping the real data with random numbers from a normal distribution (mean=0, sd=1), but otherwise follow the instructions, will generate the skewed p-value distribution shown in Figure 1, clearly indicating that this approach may have unintentional adverse effect.
+
+![fig. 1](../reanalysis/leek_sva/figure/pvaluesleeksva.svg)  
+**Figure 1**  
+Two sanity checks were the recomended use of ComBat fails. Adapted from the user guide in the sva package. 
+- a) Real data is substituted with random numbers from a normal distribution (mean=0, sd=1), but the batch/group design is retained, followed by batch adjustment in ComBat and a F-test.
+- b) 10 runs of real data with the "cancer" labels permuted within batches, followed by ComBat adjustment and a F-test. All permutations produces a skewed p-value distribution.
 
@@
 
-## Examples of Undesired consequences
+## Examples of Undesired Consequences
 As the amount of false positive results as a consequence of trying to retain group differences depends on the group/batch balance, we will show two examples with varying degree of unbalancedness.
 
-1) In the first experiment(24421904), cells are treated glatiramer acetate (a medicine for multiple sclerosis) or a generic and mRNA was measured on microarrays alongside control samples.
-A batch effect correlated to the array platform (Illumina WG-6\_V2) was observed and adjusted for with ComBat whereafter the data was tested for differetially expressed genes, yielding a list of 1000 genes (Table S5 24421904).
-Unfortunately the treatment/batch design was unbalanced with several batches having only one of the main treatment of interest.
+1) In the first experiment(24421904), cells were treated with glatiramer acetate (a medicine for multiple sclerosis) or a generic and mRNA was measured using microarrays alongside control samples.
+A batch effect correlated to the array platform (Illumina WG-6_V2) was observed and adjusted for with ComBat whereafter the data was tested for differetially expressed genes, yielding a list of 1000 genes (Table S5 24421904).
+Unfortunately the treatment/batch design was unbalanced with several batches having only one of the main treatments of interest.
 When we re-analyzed their data without using ComBat, but instead blocked for batch effect in limma, only 9 genes were found (FDR<0.05).
@@
 ![fig. 2](../reanalysis/Fadi_Towfic_generic_medicine/figure/pvaluestowfic.svg)
-**Figure 3**.
+**Figure 2**.
 P-value distribution for the main comparison in (data set 2 24421904) glatiramer acetate vs. generic.  
@@
 - b) Real data not adjusted by ComBat, tested for significance in limma using batch as a blocking factor. 
-- c) Real data is substituted with random numbers from a normal distribution (mean=0, std=1), but the batch/group design is retained, followed by batch adjustment in ComBat and a significance test in limma.
+- c) Real data is substituted with random numbers from a normal distribution (mean=0, sd=1), but the batch/group design is retained, followed by batch adjustment in ComBat and a significance test in limma.
 - d) 10 runs of real data with the glatiramer acetate("DP") and generic("N") labels permuted within batches, followed by ComBat adjustment and significance test in limma.
@@
 The experiment was conducted on three different time points (used as batches) with a total of 30 samples with a fairly balanced treatment/control across batches (6/2, 3/4 and 9/6).
-ComBat was applied followed by a t-test identifying differentially expressed genes.
+ComBat was applied followed by a t-test in order to identify differentially expressed genes.
 First, we reproduced their analysis including the adjustment by ComBat, but using limma instead of the t-test, resulting in 1003 probes (q<0.05).
-Then, we analysed their data blocking for batch in limma and not using ComBat, resulting in 377 probes (q<0.05).
+Then, we analysed their data while blocking for batch in limma and not using ComBat, resulting in 377 probes (q<0.05).
 In addition the two sanity checks outline above were performed.
@@
 - b) Real data not adjusted by ComBat, tested for significance in limma using batch as a blocking factor. 
-- c) Real data is substituted with random numbers from a normal distribution (mean=0, std=1), but the batch/group design is retained, followed by batch adjustment in ComBat and a significance test in limma.
+- c) Real data is substituted with random numbers from a normal distribution (mean=0, sd=1), but the batch/group design is retained, followed by batch adjustment in ComBat and a significance test in limma.
 - d) 10 runs of real data with the group labels permuted within batches, followed by ComBat adjustment and significance test in limma.
+
+3) Her kan vi fylle paa med flere eksempler etter samme mal hvis det trengs.
+Feks en fersk artikkel fra Nature Genetic([24584070](http://www.nature.com/ng/journal/vaop/ncurrent/full/ng.2916.html)).
+Der har de kombinert sekvenseringsdata  (deres data) fra 9 lymfom pasienter med et (eksternt) microarray data set som har 6 lymfom proever og 5 kontroller, dvs kontroll gruppen er helt fravaerende hos dem.
+Dvs veldig ubalansert (9/0, 6/5), MEN skille mellom phenotypene er saa stor at de sikkert hadde funnet omtrent de samme genene signifikante med aa bare bruke de balanserte dataene (men det var jo ikke deres.) Dette er ikke en viktig del av paperet, men det er veldig feil selvom konsekvensense ikke blir store.
+Dog gir det et falskt inntrykk av at dataene deres ligner tidligere data, se heatmap i suppl..
+Dessuten lager de GO analyser av dette som egentlig er 7 aar gamle andres data og ikke deres
 
@@
 
-## Sliding application area of ComBat
+## Increased emphasis on preserving group difference
 In the original ComBat article(2007) it is clear that the primary motivation behind ComBat was to employ an Emprical Bayes method for batch-effect removal.
@@
 
-```{r echo=FALSE}
-"
-
-Mange siteringer
-I pipelines 16642009,TCGA  pakker.   23452776, 21937664
-recomondations, review vinnere. 22682473, 22133085
-mangelfull forstaaelse fra forfattere. 16632515, 22257669
-Faa setninger, 24391845, 18414638, 21731603,23630272, 24584070
-ikke oppgit parametere, dvs batch(18414638, 21731603) eller covariates(24391845, 18414638, 21731603, 23630272, 23482648, 24584070),
-Ikke vist batch problem (24391845?, 18414638, 21731603, 23630272).
-Brukt uansett.
-P-hacking. Gjoer ComBat populaer.
-"
-```
-
-## Guide for "worried" scientist
-6. Gi en rettledning for hvordan en bekymret forsker kanskje kan
-estimere feilen p.g.a ComBat-bruk i en eller flere utvalgte artikler.
-
-likely small or no difference (24391845, 18414638?, 21731603, 23630272)
+## Practical advice
+We have shown that adjusting for batch effects while preserve the group difference may lead to varying degree of false results.
+Knowing this, to what degree can one trust a result  from a work applying such a method?
+Essentially, when the batch/group configuration is balanced, or group difference is ignored  (i.e.
+no group labels not given as parameters to ComBat), problems related to preserving group differences will not occur.
+For other cases, a re-analysis without using this approach is the most rigorous path.
+However, this thoroughness is not feasible if the downstream  analysis can not adjust for batch effects by it self.
+To reach a reliable result, batch effects need to be handled in some way or another.
+To make matters worse, a re-analysis relies on the availability of the raw data and a description of processing and analysis steps taken in the original work.
+If this is available, the necessary bioinformatic skills and work hours could still be in short supply.
+For such situations a superficial assessment could be performed, taking special note of batches were groups of interest are near missing and how likely a group difference is.
+In essence asking if the balanced parts (effective sample size?) of the data has enough power to detect the presumed effects and if this is the case, treat the results more like an ordered list with the most likely true positives on top,de-emphasizing the somewhat deflated p-values (16632515, 24584070).
+In contrast, if a group effect is a priori considered to be unlikely  (24391845, 18414638?, 21731603, 23630272), a intermediate lack of batch/group balance  could lead to a mostly false result.
 
@@
 "We combined the RNA-seq and microarray data using the Combat algorithm36 in the Bioconductor sva package to remove batch biases due to platform differences"
+
+A recurrent inactivating mutation in RHOA GTPase in angioimmunoblastic T cell lymphoma.
+Yoo HY, Sung MK, Lee SH, Kim S, Lee H, Park S, Kim SC, Lee B, Rho K, Lee JE, Cho KH, Kim W, Ju H, Kim J, Kim SJ, Kim WS, Lee S, Ko YH.
+Nat Genet. 2014 Mar 2. doi: 10.1038/ng.2916. [Epub ahead of print]
+PMID: 24584070 [PubMed - as supplied by publisher]
 

### 2014-03-13 commonscripts/helper_functions.r

@@ -274,8 +274,10 @@ drawbatchbalanceddsamples = function(samplenames, batches, conditions)
 #covariate = sampleannotation$covariate1
 #shufflecovariates = c("N", "DP")
 # shuffle samples whithin a batch for selected covariates
-shufflesamplesinbatch = function(samplenames, batch, covariate, shufflecovariates)
+shufflesamplesinbatch = function(samplenames, batch, covariate, shufflecovariates=NULL)
 {
+  if(is.null(shufflecovariates))
+    shufflecovariates=unique(covariate)
 	ret = samplenames
 	b = covariate %in% shufflecovariates
 	for(thisbatch in unique(batch))

### 2014-03-14 docs/article.rmd
mer

@@
 ## Introduction
+Ikke ferdig. notater
+Molecular biololgy experiments chained of laboartory procedures. A change or perturbation in any of the sequenctial manufactoring or laboratory procedures generating measurments
+ other effects than uncontrolled variables
+related to the condition tested
+(20838408)
+'
+"mean-centering and standardization methods may run the risk of distorting biological differences after removing batch effects." (20676067)
+eksempler på batch effects
+http://www.nature.com/tpj/journal/v10/n4/full/tpj201057a.html
+
+Confounding variables have the potiential to mislead researcher to make reach wrong conclusions if left unaccounted for.
+In Molecular Biology confounding variables is often called "batch effects", probably due to the fact that reagents and microarray chips are made in batches, and this is frequently observed as an effect in the measurements.
+See 20676067 for more examples.
+
+For a typical microaray experiment comparing groups differences, a balanced group/batch design, batch effects will lower the statistical power, while for an unbalanced design it may in additon lead to an increase in false positve results.
+In different field Nuisance Factor
+Blocking in the statistical test
 
@@
 
-## Simple sanity check
+## A simple sanity check
 The undesired consequnces of preserving group effects when correcting for batch effect is readily illustrated with a sanity check using random numbers .
@@
 
-## Examples of Undesired Consequences
+## Examples of undesired consequences
 As the amount of false positive results as a consequence of trying to retain group differences depends on the group/batch balance, we will show two examples with varying degree of unbalancedness.
@@
 1) In the first experiment(24421904), cells were treated with glatiramer acetate (a medicine for multiple sclerosis) or a generic and mRNA was measured using microarrays alongside control samples.
-A batch effect correlated to the array platform (Illumina WG-6_V2) was observed and adjusted for with ComBat whereafter the data was tested for differetially expressed genes, yielding a list of 1000 genes (Table S5 24421904).
+A batch effect correlating to the array platform (Illumina WG-6_V2) was observed and adjusted for with ComBat whereafter the data was tested for differetially expressed genes, yielding a list of 1000 genes (Table S5 24421904).
 Unfortunately the treatment/batch design was unbalanced with several batches having only one of the main treatments of interest.
@@
 ,[4](https://groups.google.com/d/msg/combat-user-forum/26FZlgU2LFQ/W6U_Lhh_64EJ)
-).
+). _Aa linke direkte er kanskje litt ufint, alternativt kan det linkes til forumet uten konkrete innlegg_.  
 And when ComBat was incorporated in the sva package ( [22257669](http://bioinformatics.oxfordjournals.org/cgi/pmidlookup?view=long&pmid=22257669), 2012), inclusion of group labels was made almost mandatory (an undocumented option of passing a NULL value exists).
@@
 
-## Motivation for this Warning
+## Motivation for this warning
 Our knowledge of ComBat came through a typical use case when trying to salvage unbalanced data which had batch effects.
@@
 A brief inquiry into  some of the articles citing ComBat (574 Google Scholar) revealed few problems, and their method descriptions regarding ComBat were mostly sparse, limited to one or two sentences (24391845, 18414638, 21731603,23630272, 24584070).
-A further indication of their carefree use of this potentially devastating procedure was the frequent neglect to state the program parameters, i.e batch-labels (18414638, 21731603) or group labels (24391845, 18414638, 21731603, 23630272, 23482648, 24584070).
+A further indication of their carefree use of this potentially devastating procedure was the frequent neglect to state the program parameters, i.e batch labels (18414638, 21731603) or group labels (24391845, 18414638, 21731603, 23630272, 23482648, 24584070).
 Often no effort was done in order to substantiate the existence of batch effects in their data, except for stating the presence of batches (24391845?, 18414638, 21731603, 23630272).
@@
 Taken together we fear that many published results from data adjusted by ComBat are completely or partially false.
-And knowing that scientist don't give up when one analysis fails, a method that almost ensures a result given a sufficiently unbalanced design will continue to be used.
+And knowing that scientists don't give up where one analysis fails, a method that almost ensures a result given a sufficiently unbalanced design will continue to be used.
 
@@
 To make matters worse, a re-analysis relies on the availability of the raw data and a description of processing and analysis steps taken in the original work.
-If this is available, the necessary bioinformatic skills and work hours could still be in short supply.
-For such situations a superficial assessment could be performed, taking special note of batches were groups of interest are near missing and how likely a group difference is.
-In essence asking if the balanced parts (effective sample size?) of the data has enough power to detect the presumed effects and if this is the case, treat the results more like an ordered list with the most likely true positives on top,de-emphasizing the somewhat deflated p-values (16632515, 24584070).
-In contrast, if a group effect is a priori considered to be unlikely  (24391845, 18414638?, 21731603, 23630272), a intermediate lack of batch/group balance  could lead to a mostly false result.
+Even when this is available, the necessary bioinformatic skills and work hours could still be in short supply.
+For such situations, a superficial assessment can be performed, taking special note of batches were groups of interest are near missing and how likely a group difference is.
+In essence asking if the balanced parts (effective sample size?) of the data has enough power to detect the presumed effects and if this is the case (16632515, 24584070), treat the results more like an ordered list with the most likely true positives on top while de-emphasizing the somewhat deflated p-values .
+In contrast, if biological knowledge suggest that a group effect is unlikely  (24391845, 18414638?, 21731603, 23630272), an intermediate lack of batch/group balance  could lead to a mostly false result.
 
@@
 
+Tackling the widespread and critical impact of batch effects in high-throughput data.
+Leek JT, Scharpf RB, Bravo HC, Simcha D, Langmead B, Johnson WE, Geman D, Baggerly K, Irizarry RA.
+Nat Rev Genet. 2010 Oct;11(10):733-9. doi: 10.1038/nrg2825. Epub 2010 Sep 14.
+PMID: 20838408 [PubMed - indexed for MEDLINE] Free PMC Article
+http://www.nature.com/nrg/journal/v11/n10/abs/nrg2825.html
+
+A comparison of batch effect removal methods for enhancement of prediction performance using MAQC-II microarray gene expression data.
+Luo J, Schumacher M, Scherer A, Sanoudou D, Megherbi D, Davison T, Shi T, Tong W, Shi L, Hong H, Zhao C, Elloumi F, Shi W, Thomas R, Lin S, Tillinghast G, Liu G, Zhou Y, Herman D, Li Y, Deng Y, Fang H, Bushel P, Woods M, Zhang J.
+Pharmacogenomics J. 2010 Aug;10(4):278-91. doi: 10.1038/tpj.2010.57.
+PMID: 20676067 [PubMed - indexed for MEDLINE] Free PMC Article
+http://www.nature.com/tpj/journal/v10/n4/full/tpj201057a.html
+
 ## skeptisk til ComBat

### 2014-03-17 docs/article.rmd
mer artikkel

@@
-Methods that remove batch effects while retaining group differences may induce false confidence in downstream analyses
+Methods that remove batch effects while retaining group differences may lead to exaggerated confidence in downstream analyses
 ========================================================
+
+```{r echo=FALSE, results='hide'}
+"
+Alternativer til tittel
+Methods that remove batch effects while retaining group differences may lead to incorrect conclusions in downstream analyses
+Methods that remove batch effects while retaining group differences may lead to incorrect conclusions
+"
+```
 
@@
 ## Introduction
-Ikke ferdig. notater
-Molecular biololgy experiments chained of laboartory procedures. A change or perturbation in any of the sequenctial manufactoring or laboratory procedures generating measurments
- other effects than uncontrolled variables
-related to the condition tested
-(20838408)
-'
-"mean-centering and standardization methods may run the risk of distorting biological differences after removing batch effects." (20676067)
-eksempler på batch effects
-http://www.nature.com/tpj/journal/v10/n4/full/tpj201057a.html
 
-Confounding variables have the potiential to mislead researcher to make reach wrong conclusions if left unaccounted for.
-In Molecular Biology confounding variables is often called "batch effects", probably due to the fact that reagents and microarray chips are made in batches, and this is frequently observed as an effect in the measurements.
+Extraneous variables, if left unaccounted for, have the potential to lead an investigator into drawing wrong conclusions.
+In molecular biology, extraneous variables are often called "batch effects", probably due to the fact that reagents and other equipment, for instance microarray chips, are made in batches, and this is frequently observed as an effect in the measurements.
 See 20676067 for more examples.
 
-For a typical microaray experiment comparing groups differences, a balanced group/batch design, batch effects will lower the statistical power, while for an unbalanced design it may in additon lead to an increase in false positve results.
-In different field Nuisance Factor
-Blocking in the statistical test
+For a typical experiment comparing group differences, the presence of batch effects will decrease the statistical power.
+If the batch/group design is unbalanced - i.e the proportion of one groups samples belonging to each batch is not equal for all groups - batch may act as a confounder and lead to false positive results as well (20838408).
 
-```{r echo=FALSE}
+The standard way to handle an extraneous variable is to include it in the statistical model employed in the inquiry.
+However, many analytic tools for high throughput data does not cater for this option, and when available it could still be outside the competence of the investigator.
+Therefore, an alternative two step procedure has emerged, first the batch effects are estimated and removed, creating a "batch effect free" data set.
+In the next step, the statistical analysis is performed on the adjusted data as there never were batch effects in the first place.
+This appealing compartmentalization is also convenient for practical purposes, for example when data-processing and statistical analyses are performed by different personnel.
+
+The first step can be achieved by subtracting the mean of the measurements in one batch from all measurements in that batch, i.e mean adjustment or one-way ANOVA adjustment as implemented in the method pamr.batchadjust from the pamr package in R.
+When the batch/group design is balanced, mean-adjustment will take out some, but not necessarily all, variance attributed to batch and leave the between group variance, thus increasing the statistical power.
+However, when batch is a confounder, some of the variance removed will be group variance and statistical power will be lost.
+In addition if there are many groups, some effect will be transferred between groups that share batches, leading to false positive results (_Dette problemet er ikke beskrevet i de andre artiklene eller i einars mail, men saa vidt jeg skjoenner saa kan dette skje_).
+
+To mitigate the above problems, group differences can, simplified explained, be estimated based on the balanced part of the data and applied to all samples in the adjusted data set.
+If all the samples, and not only the balanced ones, are used as sample size in later a group comparison, the confidence will be overestimated.
+
+The ComBat method described in 16632515 and also included in the sva package (22257669) implements the above described procedure in an attempt to retain group differences when adjusting for batch effects.
+Based on actual use of ComBat by the authors and others, we suspect the adjusted data predominantly is treated as "batch effect free" in the subsequent analysis.
+And as a consequence, confidence is overestimated and false results are reported.
+
+```{r echo=FALSE, results='hide'}
 "
@@
 ## A simple sanity check
-The undesired consequnces of preserving group effects when correcting for batch effect is readily illustrated with a sanity check using random numbers .
+The undesired consequences of preserving group effects when correcting for batch effect is readily illustrated with a sanity check using random numbers .
 The documentation accompanying the sva library has a runnable example demonstrating how to adjust a data set with ComBat followed by a F-test.
-Swapping the real data with random numbers from a normal distribution (mean=0, sd=1), but otherwise follow the instructions, will generate the skewed p-value distribution shown in Figure 1, clearly indicating that this approach may have unintentional adverse effect.
+Swapping the real data with random numbers from a normal distribution (mean=0, sd=1), but otherwise following the instructions, will generate the p-value distribution shown in Figure 1.
+The skewed distribution is for us a clear indication that this approach may have a unintentional adverse effect.
+_Jeg har ikke beskrevet permutatsjon checken fordi jeg er usikker paa om vi skal bruke den._
 
@@
 **Figure 1**  
-Two sanity checks were the recomended use of ComBat fails. Adapted from the user guide in the sva package. 
+Two sanity checks were the recommended use of ComBat fails. Adapted from the user guide in the sva package. 
 - a) Real data is substituted with random numbers from a normal distribution (mean=0, sd=1), but the batch/group design is retained, followed by batch adjustment in ComBat and a F-test.
@@
 ## Formler og ord
-2. Forklare hva som gaar galt med formler og ord.
+_Forklare hva som gaar galt med formler og ord. Overlater til Einar aa skrive dette_
 
 ## Examples of undesired consequences
-As the amount of false positive results as a consequence of trying to retain group differences depends on the group/batch balance, we will show two examples with varying degree of unbalancedness.
+As the amount of false positive results when trying to retain group differences depends on the group/batch balance, we will show two examples with varying degree of unbalancedness.
 
 1) In the first experiment(24421904), cells were treated with glatiramer acetate (a medicine for multiple sclerosis) or a generic and mRNA was measured using microarrays alongside control samples.
-A batch effect correlating to the array platform (Illumina WG-6_V2) was observed and adjusted for with ComBat whereafter the data was tested for differetially expressed genes, yielding a list of 1000 genes (Table S5 24421904).
-Unfortunately the treatment/batch design was unbalanced with several batches having only one of the main treatments of interest.
+A batch effect correlating to the chip (Illumina WG-6_V2, six samples per chip) was observed and adjusted for with ComBat, whereafter the data was tested for differetially expressed genes, yielding a list of 1000 genes (Table S5 24421904).
+Unfortunately the batch/treatment design was unbalanced with several batches having only one of the main treatments of interest.
 When we re-analyzed their data without using ComBat, but instead blocked for batch effect in limma, only 9 genes were found (FDR<0.05).
-Additional sanity checks were also applied and the distribution of p-values for differents settings are shown in Fig.
+Additional sanity checks with random numbers or perumted labels were also carried out and the distribution of p-values for differents settings are shown in Fig.
 2.
@@
 - d) 10 runs of real data with the glatiramer acetate("DP") and generic("N") labels permuted within batches, followed by ComBat adjustment and significance test in limma.  
+_jeg tok med begge sanitycheckene her. Er d) slik einar foreslo?. Merk jeg har ikke permutert alle labeler, bare de mellom DP og N. Det er en del andre._
 
@@
 Cells inhibited for the expression of the TAL1 gene were compared to controls on a microarray platform.
-The experiment was conducted on three different time points (used as batches) with a total of 30 samples with a fairly balanced treatment/control across batches (6/2, 3/4 and 9/6).
+The experiment was conducted on three different time points (used as batches) with a total of 30 samples and a fairly balanced batch/treatment set up (6-2, 3-4 and 9-6).
 ComBat was applied followed by a t-test in order to identify differentially expressed genes.
 First, we reproduced their analysis including the adjustment by ComBat, but using limma instead of the t-test, resulting in 1003 probes (q<0.05).
-Then, we analysed their data while blocking for batch in limma and not using ComBat, resulting in 377 probes (q<0.05).
+Then, we analysed their data without batch adjustment in ComBat, but blocking for batch in limma, resulting in 377 probes (q<0.05).
 In addition the two sanity checks outline above were performed.
-Comparing the P-value distributions for the alternative analyses (Fig.3 a,b) does not indicate a huge difference, in contrast to results above for(24421904).
+In contrast to the above results obtained for(24421904), the P-value distributions for the alternative analyses (Fig.3 a,b) does not indicate a huge difference.
+Nevertheless, we belive that  P-values are deflated for the ComBat adjusted analysis.
 
@@
 - d) 10 runs of real data with the group labels permuted within batches, followed by ComBat adjustment and significance test in limma.   
+_problemet med permutatsjons sanitychechen slik jeg har utfoert og plottet den her er at den neppe kan sies aa feile.
+Dessuten kan de 3 foerst plottene bli plottet i ett som vist under i en alternativ version jeg synes er finere, men dette gaar ikke for permutasjonstesten tror jeg_
 
-3) Her kan vi fylle paa med flere eksempler etter samme mal hvis det trengs.
+![fig. 4](../reanalysis/ComBat_original/figure/pvaluesjohnson2.svg)   
+_Alternativ illustrasjon av de tre p-value distribusjonene_
+
+3) _Her kan vi fylle paa med flere eksempler etter samme mal hvis det trengs.
 Feks en fersk artikkel fra Nature Genetic([24584070](http://www.nature.com/ng/journal/vaop/ncurrent/full/ng.2916.html)).
@@
 Dog gir det et falskt inntrykk av at dataene deres ligner tidligere data, se heatmap i suppl..
-Dessuten lager de GO analyser av dette som egentlig er 7 aar gamle andres data og ikke deres
+Dessuten lager de GO analyser av dette som egentlig er 7 aar gamle andres data og ikke deres_
 
@@
 Often no effort was done in order to substantiate the existence of batch effects in their data, except for stating the presence of batches (24391845?, 18414638, 21731603, 23630272).
-In one instance ComBat was even applied on data were effects due to batch were investigated, but not found (husker ikke ref).
+In one instance ComBat was even applied on data were effects due to batch were investigated, but not found (_husker ikke ref_).
 The incorporation of the method into analysis pipelines (16642009,TCGA) and other packages (23452776, 21937664) could make its usage even more trivial and parameters setting harder to perceive.
@@
 We have shown that adjusting for batch effects while preserve the group difference may lead to varying degree of false results.
-Knowing this, to what degree can one trust a result  from a work applying such a method?
+Knowing this, to what degree can an investigator trust a result from a work applying such a method?
 Essentially, when the batch/group configuration is balanced, or group difference is ignored  (i.e.
@@
 ## Reproducible research
+_vi boer klare aa tilfredstille kravene i http://biostatistics.oxfordjournals.org/content/10/3/405.full_
 
@@
 
-Combat org
 http://biostatistics.oxfordjournals.org/content/8/1/118.abstract
@@
 
-1
 Molecular profiling of single Sca-1+/CD34+,- cells--the putative murine lung stem cells.
@@
 http://www.plosone.org/article/info%3Adoi%2F10.1371%2Fjournal.pone.0083917
-" Based on the hybridization date the data set is separated into two main batches of microarrays (early 2007/2008 and late 2009), which degrade further into small sub-batches.
-We used the ComBat algorithm [15] to adjust for the main batches.
-"
 
-2
 Drinking-water arsenic exposure modulates gene expression in human lymphocytes from a U.S. population.
@@
 http://www.ncbi.nlm.nih.gov/pmc/articles/PMC2290973/
-"followed by empirical Bayes adjustment procedures implemented on the R platform, as described previously (Johnson et al. 2007)"
 
-3
 Epigenetic predictor of age.
@@
 http://www.plosone.org/article/info%3Adoi%2F10.1371%2Fjournal.pone.0014821
-"Batch effect were removed using the Combat algorithm [9], and one outlier sample was removed."
 
-4.
 Childhood maltreatment is associated with distinct genomic and epigenetic profiles in posttraumatic stress disorder.
@@
 http://www.pnas.org/content/110/20/8302.full
-"To correct for confounding as a result of batch effects, the data were normalized using an empirical Bayes method for batch correction " (expression)
-"Hybridization and chip batch effects were accounted for using an empirical Bayes method (43)." (methylation)
 
-5.
 Patterns of regulatory activity across diverse human cell types predict tissue identity, transcription factor binding, and long-range interactions.
@@
 http://genome.cshlp.org/content/23/5/777.full
-"We improved on the previously published open chromatin measurements by accounting for batch affects that grouped the data by laboratory rather than by biological signal (see Methods).
-We used ComBat (Johnson et al.
-2007) to remove these batch effects, after which both the DNase I and expression data clustered according to expected biological relationships (Supplemental Fig.
-S1)."
-"Counts were quantile-normalized and scaled, and protocol batch effects were corrected using ComBat (Supplemental Fig. S1; Johnson et al. 2007)."
-"Artifactual differences between protocols (Figure S1) were corrected using ComBat (Johnson et al.
-2007).
-Because the data was too large for the ComBat model, we divided it into subsets of about 300,000 DHSs and corrected the batch affect individually on each subset.
-"
-"To make the arrays comparable, we used ComBat to correct for this batch affect (Johnson et al.
-2007).
-After correction with ComBat, the arrays grouped according to expected biological similarity (Figure S1).
-"
 
-6.
 A recurrent inactivating mutation in RHOA GTPase in angioimmunoblastic T cell lymphoma.
@@
 PMID: 24584070 [PubMed - as supplied by publisher]
-http://www.nature.com/ng/journal/vaop/ncurrent/pdf/ng.2916.pdf
 http://www.nature.com/ng/journal/vaop/ncurrent/full/ng.2916.html
-http://www.nature.com/ng/journal/vaop/ncurrent/extref/ng.2916-S1.pdf
-"We combined the RNA-seq and microarray data using the Combat algorithm36 in the Bioconductor sva package to remove batch biases due to platform differences"
 
@@
 
-## anbefaler ComBat
-
-10
 Batch correction of microarray data substantially improves the identification of genes differentially expressed in rheumatoid arthritis and osteoarthritis.
@@
 
-11
 Relative impact of key sources of systematic noise in Affymetrix and Illumina gene-expression microarray experiments.
@@
 
-## skeptisk til ComBat
+Removing batch effects in analysis of expression microarray data: an evaluation of six batch adjustment methods.
+Chen C, Grennan K, Badner J, Zhang D, Gershon E, Jin L, Liu C.
+PLoS One. 2011 Feb 28;6(2):e17238. doi: 10.1371/journal.pone.0017238.
+PMID: 21386892 [PubMed - indexed for MEDLINE] Free PMC Article
+http://www.plosone.org/article/info%3Adoi%2F10.1371%2Fjournal.pone.0017238
 
-20 - HAr kjoert permutering av covariatelabler og finner at combat finner masse gener uansett. Fokus paa artikkel er ikke combat spesifikkt.
+T. Hastie, R. Tibshirani, Balasubramanian Narasimhan and Gil Chu (2013). pamr: Pam: prediction analysis for microarrays. R package version
+  1.54.1. http://CRAN.R-project.org/package=pamr
+
 Expanding the understanding of biases in development of clinical-grade molecular signatures: a case study in acute respiratory viral infections.
@@
 http://www.plosone.org/article/info%3Adoi%2F10.1371%2Fjournal.pone.0020662
-
-## ComBat i pipelines
 
@@
 http://www.nature.com/ng/journal/v38/n5/full/ng0506-500.html
-
-Uverifisert.
-insilicodb? bioconductopakke eller hva
-http://bib.oxfordjournals.org/content/early/2012/07/31/bib.bbs037.short
 
@@
 
-## eksempler med reporduserte p-value
-
 Comparing the biological impact of glatiramer acetate with the biological impact of a generic.
@@
 http://www.plosone.org/article/info%3Adoi%2F10.1371%2Fjournal.pone.0083757
-
-### user group recomandations
-https://groups.google.com/d/msg/combat-user-forum/eSVSKwGtuyE/ZIWV2juYmmAJ
-https://groups.google.com/d/msg/combat-user-forum/Vkb9p7wekd4/h5Etie7FTVQJ
-https://groups.google.com/d/msg/combat-user-forum/fpBTcgDjiR8/uo4QIZL4sZgJ
-https://groups.google.com/d/msg/combat-user-forum/26FZlgU2LFQ/W6U_Lhh_64EJ
-
-Alternativer til tittel
-Methods that remove batch effects while retaining group differences may lead to incorrect conclusions in downstream analyses
-Methods that remove batch effects while retaining group differences may lead to incorrect conclusions
-
-```{r}
-
-```

### 2014-03-17 reanalysis/ComBat_original/combat_paper_dataset2_reproduced.rmd
mer artikkel

@@
 
+Forslag 2 figur brukt i artikkel.
+```{r pvaluesjohnson2, dev='svg', fig.width=10, fig.height=10, tidy=FALSE}
+# figure used in our article
+# NB! bruker orgniale data som ikke er floored eller log2!
+par(mfrow=c(1, 1))
+
+# reproduced ComBat + limma
+hist(EB3_limma_res$p.value[,1], main="P-values, 3 settings", breaks=100, xlab="p-value", ylim=c(0,3000), border="red")
+
+# batch handled in limma
+hist(datamatrix_limma_res$p.value[,1], add=T, breaks=100, xlab="p-value" , ylim=c(0,3000), border="blue")
+
+# random ComBat + limma
+hist(EBrand_limma_pvalues, add=T, breaks=100, xlab="p-value" , ylim=c(0,3000), border="black")
+
+legend("topright", legend=c("Reald data ComBat adjusted", "Reald data Limma adjusted", "Random data ComBat adjusted"),
+       text.col=c("red", "blue", "black"))
+```
+
 ### References

### 2014-03-17 commonscripts/helper_functions.r

@@ -44,17 +44,17 @@ addbatcheffect = function(df, batches, thismean=1, thissd=1)
 }
 
 # 
-addconditioneffect = function(df, conditions, ndiffgenes, thismean=1, betweensd=1, insidesd=1)
+addconditioneffect = function(df, labels, ndiffgenes, thismean=1, betweensd=1, insidesd=1, affectedconditions)
 {
 	ret = df
 	if(ndiffgenes>0)
 	{
-		for(i in 1:length(unique(conditions)))
+		for(i in 1:length(affectedconditions))
 		{
 			for(s in 1:ndiffgenes)
 			{
-				thisgeneseffect = rnorm(thismean, sd=betweensd)
-				a=conditions==conditions[i]
+				thisgeneseffect = rnorm(1, thismean, sd=betweensd)
+				a=labels==affectedconditions[i]
 				#ret[s, a] = ret[s, a]+ rnorm(length(ret[s, a]), mean=thisgeneseffect)		
 				ret[s, a] = rnorm(length(ret[s, a]), mean=thisgeneseffect, sd=insidesd)		
 			}
@@ -87,7 +87,7 @@ getdifftab_sva = function(edata, sa, mod, threshold=0.05)
 getdifftab_limma = function(edata, condition,  contrast, block=NULL, threshold=0.05)
 {
 	require(limma)
-	fac = as.factor(condition)	
+	fac = factor(condition)	
 	design = model.matrix(~0 + fac)
 	if(!is.null(block))
 	{

### 2014-03-17 reanalysis/ComBat_original/combat_paper_dataset2_reproduced.rmd
fra imac

--- 2014-03-03 reanalysis/ComBat_original/combat_paper_dataset2_reproduced.rmd
@@
 Type = as.factor(sampleannotation$Type)
+Block = as.factor(sampleannotation$Batch)
 design = model.matrix(~0 + Type)
@@
 
-### Additional sanity checks
-
+Additional sanity checks
+---------
 To substantiate that the result from the use of ComBat is less trustworthy than the alternative analysis we provide a few additional sanity checks.
@@
 generation ended `r as.character(Sys.time())`. Time spent `r  as.integer(round(difftime(Sys.time(),starttime, units="mins")))` minutes .
+
+table(sampleannotation[ sampleannotation$Batch!="3", c("Batch", "Cell")]  )
+table(Batch45_annot[, c("Batch", "Cell")])
+table(sampleannotation[, c("Batch", "Cell")])

### 2014-03-17 docs/vedlegg.rmd

--- 2014-03-13 docs/article.rmd
@@
-Methods that remove batch effects while retaining group differences may induce false confidence in downstream analyses
+Title
 ========================================================
 
-## Abstract
-Removal of, or adjustment for, batch effects or centre differences is generally required when such effects are present in data.
-In particular, when preparing microarray gene expression data from multiple cohorts, array platforms, or batches for later analyses, batch effects can have confounding effects.
-Many methods and tools exist for this purpose.
-One method, ComBat which is part of the R package sva, is particularly popular due to its ability to remove batch differences even when batches are small and heterogeneous, while retaining the difference between study groups, e.g.
-treatments.
-Unfortunately, this recommended and frequently used approach to batch correction will systematically induce incorrect group differences in downstream analyses when groups are unevenly distributed between the batches.
-The scientific community seems to be largely unaware of this problem, which most likely has contributed to false discoveries being presented in the published literature.
-
-## Introduction
-
-```{r echo=FALSE}
-"
-Kort oppsummere behovet for batch-korrigering: generelt og i microarraydata.
-
-Kort beskrive metoder som gjoer dette (batch-sentrering, ANOVA, ComBat).
-
-Forklare tilsiktet effekt disse paa batch+gruppe-forskjeller (balansert versus ikke-balansert).
-
-Forklare kort hva faktisk effekt er: ANOVA gir redusert gruppe-forskjeller, mens for ComBat overestimeres den statistiske styrken/sikkerheten av forskjellen.
-"
-```
-
-## Simple sanity check
-The undesired consequnces of preserving group effects when correcting for batch effect is readily illustrated with a sanity check using random numbers .
-The documentation accompanying the sva library has a runnable example demonstrating how to adjust a data set with ComBat followed by a F-test.
-Swapping the real data with random numbers from a normal distribution (mean=0, sd=1), but otherwise follow the instructions, will generate the skewed p-value distribution shown in Figure 1, clearly indicating that this approach may have unintentional adverse effect.
-
-![fig. 1](../reanalysis/leek_sva/figure/pvaluesleeksva.svg)  
-**Figure 1**  
-Two sanity checks were the recomended use of ComBat fails. Adapted from the user guide in the sva package. 
-- a) Real data is substituted with random numbers from a normal distribution (mean=0, sd=1), but the batch/group design is retained, followed by batch adjustment in ComBat and a F-test.
-- b) 10 runs of real data with the "cancer" labels permuted within batches, followed by ComBat adjustment and a F-test. All permutations produces a skewed p-value distribution.
-
-## Formler og ord
-2. Forklare hva som gaar galt med formler og ord.
-
-## Examples of Undesired Consequences
-As the amount of false positive results as a consequence of trying to retain group differences depends on the group/batch balance, we will show two examples with varying degree of unbalancedness.
-
-1) In the first experiment(24421904), cells were treated with glatiramer acetate (a medicine for multiple sclerosis) or a generic and mRNA was measured using microarrays alongside control samples.
-A batch effect correlated to the array platform (Illumina WG-6_V2) was observed and adjusted for with ComBat whereafter the data was tested for differetially expressed genes, yielding a list of 1000 genes (Table S5 24421904).
-Unfortunately the treatment/batch design was unbalanced with several batches having only one of the main treatments of interest.
-When we re-analyzed their data without using ComBat, but instead blocked for batch effect in limma, only 9 genes were found (FDR<0.05).
-Additional sanity checks were also applied and the distribution of p-values for differents settings are shown in Fig.
-2.
-Our conclusion is that most of the genes reported as differtially expresssed in (24421904) are false positives.
-This example is a sort of "worst case" scenario for applying ComBat, since it both has a very unbalanced batch/group design and a a priori assumpotion of no difference.
-The R-code for our analysis and a more extensive report can be downloaded from github (https://github.com/vegardny/combat_tests.git).
-
-![fig. 2](../reanalysis/Fadi_Towfic_generic_medicine/figure/pvaluestowfic.svg)
-**Figure 2**.
-P-value distribution for the main comparison in (data set 2 24421904) glatiramer acetate vs. generic.  
-- a) Real data adjusted by ComBat as described in the article followed by a significance test in limma. 
-- b) Real data not adjusted by ComBat, tested for significance in limma using batch as a blocking factor. 
-- c) Real data is substituted with random numbers from a normal distribution (mean=0, sd=1), but the batch/group design is retained, followed by batch adjustment in ComBat and a significance test in limma.
-- d) 10 runs of real data with the glatiramer acetate("DP") and generic("N") labels permuted within batches, followed by ComBat adjustment and significance test in limma.
-
-2) The second example is taken from the supporting information for the original ComBat article (16632515) where it is denoted "data set 2".
-Cells inhibited for the expression of the TAL1 gene were compared to controls on a microarray platform.
-The experiment was conducted on three different time points (used as batches) with a total of 30 samples with a fairly balanced treatment/control across batches (6/2, 3/4 and 9/6).
-ComBat was applied followed by a t-test in order to identify differentially expressed genes.
-First, we reproduced their analysis including the adjustment by ComBat, but using limma instead of the t-test, resulting in 1003 probes (q<0.05).
-Then, we analysed their data while blocking for batch in limma and not using ComBat, resulting in 377 probes (q<0.05).
-In addition the two sanity checks outline above were performed.
-Comparing the P-value distributions for the alternative analyses (Fig.3 a,b) does not indicate a huge difference, in contrast to results above for(24421904).
-
-![fig. 3](../reanalysis/ComBat_original/figure/pvaluesjohnson.svg)  
-**Figure 3**.
-P-value distribution for the main comparison in (data set 2 16632515) for  
-- a) Real data adjusted by ComBat as described in the article followed by a significance test in limma. 
-- b) Real data not adjusted by ComBat, tested for significance in limma using batch as a blocking factor. 
-- c) Real data is substituted with random numbers from a normal distribution (mean=0, sd=1), but the batch/group design is retained, followed by batch adjustment in ComBat and a significance test in limma.
-- d) 10 runs of real data with the group labels permuted within batches, followed by ComBat adjustment and significance test in limma.
-
-3) Her kan vi fylle paa med flere eksempler etter samme mal hvis det trengs.
-Feks en fersk artikkel fra Nature Genetic([24584070](http://www.nature.com/ng/journal/vaop/ncurrent/full/ng.2916.html)).
-Der har de kombinert sekvenseringsdata  (deres data) fra 9 lymfom pasienter med et (eksternt) microarray data set som har 6 lymfom proever og 5 kontroller, dvs kontroll gruppen er helt fravaerende hos dem.
-Dvs veldig ubalansert (9/0, 6/5), MEN skille mellom phenotypene er saa stor at de sikkert hadde funnet omtrent de samme genene signifikante med aa bare bruke de balanserte dataene (men det var jo ikke deres.) Dette er ikke en viktig del av paperet, men det er veldig feil selvom konsekvensense ikke blir store.
-Dog gir det et falskt inntrykk av at dataene deres ligner tidligere data, se heatmap i suppl..
-Dessuten lager de GO analyser av dette som egentlig er 7 aar gamle andres data og ikke deres
-
-## Discussion
-
-## Increased emphasis on preserving group difference
-In the original ComBat article(2007) it is clear that the primary motivation behind ComBat was to employ an Emprical Bayes method for batch-effect removal.
-The feature of retaining group differences for unbalanced designs is optional and seems to be subordinate, only exemplified in the supplementary information.
-However, over the years this feature became more important judging from advice given by the author to users
-(
-[1](https://groups.google.com/d/msg/combat-user-forum/eSVSKwGtuyE/ZIWV2juYmmAJ)
-,[2](https://groups.google.com/d/msg/combat-user-forum/Vkb9p7wekd4/h5Etie7FTVQJ)
-,[3](https://groups.google.com/d/msg/combat-user-forum/fpBTcgDjiR8/uo4QIZL4sZgJ)
-,[4](https://groups.google.com/d/msg/combat-user-forum/26FZlgU2LFQ/W6U_Lhh_64EJ)
-).
-And when ComBat was incorporated in the sva package ( [22257669](http://bioinformatics.oxfordjournals.org/cgi/pmidlookup?view=long&pmid=22257669), 2012), inclusion of group labels was made almost mandatory (an undocumented option of passing a NULL value exists).
-Thus, this problematic use has likely been common.
-
-## Motivation for this Warning
-Our knowledge of ComBat came through a typical use case when trying to salvage unbalanced data which had batch effects.
-Upon realizing that the confidence on our group differences were exaggerated, the literature was searched for a better understanding of correct use and potential overseen limitations of ComBat.
-But the authors of ComBat and the sva package recommended our usage (16632515, 22257669).
-In addition, other works looking into the problem of batch effects were mostly recommending ComBat without much concern (22682473, 22133085).
-A brief inquiry into  some of the articles citing ComBat (574 Google Scholar) revealed few problems, and their method descriptions regarding ComBat were mostly sparse, limited to one or two sentences (24391845, 18414638, 21731603,23630272, 24584070).
-A further indication of their carefree use of this potentially devastating procedure was the frequent neglect to state the program parameters, i.e batch-labels (18414638, 21731603) or group labels (24391845, 18414638, 21731603, 23630272, 23482648, 24584070).
-Often no effort was done in order to substantiate the existence of batch effects in their data, except for stating the presence of batches (24391845?, 18414638, 21731603, 23630272).
-In one instance ComBat was even applied on data were effects due to batch were investigated, but not found (husker ikke ref).
-The incorporation of the method into analysis pipelines (16642009,TCGA) and other packages (23452776, 21937664) could make its usage even more trivial and parameters setting harder to perceive.
-Taken together we fear that many published results from data adjusted by ComBat are completely or partially false.
-And knowing that scientist don't give up when one analysis fails, a method that almost ensures a result given a sufficiently unbalanced design will continue to be used.
-
-## Practical advice
-We have shown that adjusting for batch effects while preserve the group difference may lead to varying degree of false results.
-Knowing this, to what degree can one trust a result  from a work applying such a method?
-Essentially, when the batch/group configuration is balanced, or group difference is ignored  (i.e.
-no group labels not given as parameters to ComBat), problems related to preserving group differences will not occur.
-For other cases, a re-analysis without using this approach is the most rigorous path.
-However, this thoroughness is not feasible if the downstream  analysis can not adjust for batch effects by it self.
-To reach a reliable result, batch effects need to be handled in some way or another.
-To make matters worse, a re-analysis relies on the availability of the raw data and a description of processing and analysis steps taken in the original work.
-If this is available, the necessary bioinformatic skills and work hours could still be in short supply.
-For such situations a superficial assessment could be performed, taking special note of batches were groups of interest are near missing and how likely a group difference is.
-In essence asking if the balanced parts (effective sample size?) of the data has enough power to detect the presumed effects and if this is the case, treat the results more like an ordered list with the most likely true positives on top,de-emphasizing the somewhat deflated p-values (16632515, 24584070).
-In contrast, if a group effect is a priori considered to be unlikely  (24391845, 18414638?, 21731603, 23630272), a intermediate lack of batch/group balance  could lead to a mostly false result.
-
-## Reproducible research
+Mer forklaring om artiklene referert til i hovedartikkelen. Spesielt de setningene som ComBat bruk er beskrevet.
 
@@
 
+Tackling the widespread and critical impact of batch effects in high-throughput data.
+Leek JT, Scharpf RB, Bravo HC, Simcha D, Langmead B, Johnson WE, Geman D, Baggerly K, Irizarry RA.
+Nat Rev Genet. 2010 Oct;11(10):733-9. doi: 10.1038/nrg2825. Epub 2010 Sep 14.
+PMID: 20838408 [PubMed - indexed for MEDLINE] Free PMC Article
+http://www.nature.com/nrg/journal/v11/n10/abs/nrg2825.html
+
+A comparison of batch effect removal methods for enhancement of prediction performance using MAQC-II microarray gene expression data.
+Luo J, Schumacher M, Scherer A, Sanoudou D, Megherbi D, Davison T, Shi T, Tong W, Shi L, Hong H, Zhao C, Elloumi F, Shi W, Thomas R, Lin S, Tillinghast G, Liu G, Zhou Y, Herman D, Li Y, Deng Y, Fang H, Bushel P, Woods M, Zhang J.
+Pharmacogenomics J. 2010 Aug;10(4):278-91. doi: 10.1038/tpj.2010.57.
+PMID: 20676067 [PubMed - indexed for MEDLINE] Free PMC Article
+http://www.nature.com/tpj/journal/v10/n4/full/tpj201057a.html
+
+Removing batch effects in analysis of expression microarray data: an evaluation of six batch adjustment methods.
+Chen C, Grennan K, Badner J, Zhang D, Gershon E, Jin L, Liu C.
+PLoS One. 2011 Feb 28;6(2):e17238. doi: 10.1371/journal.pone.0017238.
+PMID: 21386892 [PubMed - indexed for MEDLINE] Free PMC Article
+http://www.plosone.org/article/info%3Adoi%2F10.1371%2Fjournal.pone.0017238
+
+T. Hastie, R. Tibshirani, Balasubramanian Narasimhan and Gil Chu (2013). pamr: Pam: prediction analysis for microarrays. R package version
+  1.54.1. http://CRAN.R-project.org/package=pamr
+
 ## skeptisk til ComBat
@@
 https://groups.google.com/d/msg/combat-user-forum/26FZlgU2LFQ/W6U_Lhh_64EJ
-
-Alternativer til tittel
-Methods that remove batch effects while retaining group differences may lead to incorrect conclusions in downstream analyses
-Methods that remove batch effects while retaining group differences may lead to incorrect conclusions
-
-```{r}
-
-```

### 2014-03-17 reanalysis/Fadi_Towfic_generic_medicine/towfic_et_al_tomuch.rmd

--- 2014-02-27 reanalysis/Fadi_Towfic_generic_medicine/towfic_et_at.rmd
@@
 
-hist(fstatistics[table_s4$Probe], breaks=100, add=F, freq=T, border=thiscolors[2])
+#hist(fstatistics[table_s4$Probe], breaks=100, add=F, freq=T, border=thiscolors[2])
 ```
@@
 
-How this fold change was calculated is not explained. A common way is to take the difference of geometric means that were calculated for Table S2 above.
+There must be a typo in the figure text, they ment "Comparison of expression in GA to expression in **generic** for each probe, ...".
+
+> To find differentially expressed probes between generic and GA, we utilized various statistical tests at the probe level and merged the results across the different methods. <cite> Towfic et al.
+
+How the fold change was calculated is not explained. A common way is to take the difference of geometric means that were calculated for Table S2 above.
 ```{r dev='png', fig.width=8, fig.height=8, fig.show='hold', tidy=FALSE}
@@
      main="FC comparison")
-
+rm(lim)
 ```
@@
 There seems to two linear relationship, one for neagtive FC's and another for positve FC's. They seems to have different intercepts but comparable slope.
-The methods part is more specific;
->  First, we computed the statistical significance of differential expression between treatments using the Analysis of Variance (ANOVA) method for each probe [21], adjusting for multiple hypothesis testing using the Benjamini-Hochberg False Discovery Rate (FDR) correction [22].
 
-#"Probes with Fold Change <  1.10 notdisplayed"
+The average expression for "GA" (DP) and "generic" (N) are also given in Table S2. The fold change calculated from those numbers compared to the ones given in Table S5 shows a similar pattern:
 
-# Table s5 seems to be a comparison of expression in  GA (drug product) vs.
-generic (N) , listing the probes with greatest fold change (>abs(1.1)??) and the p-value calculated by several different tests.
-There seems to be no description of how the fold-change was calculated and after an inspection they look weird.
-
-meanDP = rowMeans(thisdata[, sampleannotation$covariate=="DP"]) # DP is drug product also called GA
-meanN = rowMeans(thisdata[, sampleannotation$covariate =="N"]) # N is Natco the generic,  (or G for alternative covariates)
-
-# fold change between N and DP, calculated as geometric mean.
-x = table_s5$Fold_Change
-y = (meanDP - meanN)[table_s5$Probe]
-lim =c( min(   c(min(x), min(y))  ), max( c(max(x), max(y)) ))
-plot(x, y , ylim=lim, xlim=lim, pch=".", xlab="Table_S5", ylab="reproduced", main="FC comparison") # Not a perfect match. And the table_s5 numbers seems offset.
-```
 ```{r dev='png', fig.width=8, fig.height=8, fig.show='hold', tidy=FALSE}
@@
 lim =c( min(   c(min(x, na.rm=T), min(y, na.rm=T))  ), max( c(max(x, na.rm=T), max(y, na.rm=T)) ))
-plot(x, y , ylim=lim, xlim=lim, pch="." ,  xlab="Table_S5", ylab="Table_S2", main="FC comparison") # perfect match.
+plot(x, y , ylim=lim, xlim=lim, pch="." ,  xlab="Table_S5", ylab="Table_S2",
+     main="FC comparison, given values.") #  almost perfect match with offset.
+rm(x, y, lim, fc_table_s2)
 ```
-But this is odd, we were able to reproduce table s2 and table s2 matches table s5, but table s5 fold changes does not match our reproduced values as good as the match for table s2.
+We are not able to reporduce the foldchanges fully.
 
-The reason that the line is not as straight for the reproduce FC's for table s5 as for the mean expressions in table S2 is because of different scale in the plot and the fact that the small difference will be higher for a ratio (combining to values with a individual deviation from the tables.)
-```{r dev='png', fig.width=8, fig.height=8, fig.show='hold', tidy=FALSE}
-x = (meanDP - meanN)[table_s2$ID]
-y = fc_table_s2
-plot(x, y , ylim=lim, xlim=lim, pch="." ,  xlab="reproduced", ylab="Table_S2", main="FC comparison") # perfect match.
+>  First, we computed the statistical significance of differential expression between treatments using the Analysis of Variance (ANOVA) method for each probe [21]...<cite> Towfic et al.
+
+```{r}
+anova_p = apply(datamatrices[["real_combat_covariates"]], 1,
+              function(x){ summary(aov(x[DP|N]~sampleannotation$covariate[DP|N]))[[1]][1,5]})
 ```
 
+> Next, we utilized Linear Models for Microarray (LIMMA) data analysis [23], [24] R package, part of the Bioconductor framework [25], to compare generic and GA samples, fitting a linear model that adjusts for fixed effect from medium (Effect = (GA-generic) – (generic-Medium)).
+The coefficients for the linear model were tested for significance using a modified t-test (taking into account standard deviation)...<cite> Towfic et al.
+
+The formula "Effect = (GA-generic) – (generic-Medium)" seems strange. A more standard way would be GA-generic.
+
 ```{r}
-# Still there is a offset in the numbers.
-commonprobes = table_s5$Probe[table_s5$Probe %in% table_s2$ID]
-fc_table_s5 = table_s5$Fold_Change[match(commonprobes ,table_s5$Probe)]
-head(cbind(fc_table_s2[commonprobes], fc_table_s5, fc_table_s5-fc_table_s2[commonprobes], fc_table_s5/fc_table_s2[commonprobes]))
-# seems that some unacounted transformation has occured on the fold changes.
+group = factor(sampleannotation$covariate)
+design = model.matrix(~0 + group)
+fit = lmFit(datamatrices[["real_combat_covariates"]], design)
+cont.matrix = makeContrasts ( contrasts="(groupDP-groupN) - (groupN-groupM)", levels=design)  
+fit2 = contrasts.fit(fit, cont.matrix)
+limma_p_org = eBayes(fit2)$p.value
+
+cont.matrix = makeContrasts ( contrasts="groupDP-groupN", levels=design)  
+fit2 = contrasts.fit(fit, cont.matrix)
+limma_p_alt = eBayes(fit2)$p.value
 ```
+> In parallel, we used Comparative Marker Selection as implemented in GenePattern [26] to directly compare probes between generic and GA.
+We applied two separate techniques within this framework; a traditional T-test and a Signal-to-Noise Ratio test (SNR).
+
+The T-test is stright forward.
 ```{r}
-source("../../commonscripts/helper_functions.r")
-getdifftab_limma
-```
-```{r}
-# reproduce the Nom P-value columns
-limma_ret= getdifftab_limma(edata=datamatrices[["real_combat_covariates"]], condition=sampleannotation[,"covariate"], contrast="DP-N" )
-limma_p = limma_ret$p
-names(limma_p) = limma_ret$gene
 DP = sampleannotation$covariate=="DP"
@@
 ttest_p = apply(datamatrices[["real_combat_covariates"]], 1, function(x) { t.test(x[DP], x[N])$p.value } )
-anova_p = apply(datamatrices[["real_combat_covariates"]], 1, function(x){ summary(aov(x[DP|N]~sampleannotation$covariate[DP|N]))[[1]][1,5]}  )  # aov er kanskje feil for dette, Er ikke live mange pr?ver i guppene.
+```
 
-# wilcox.test gives a few hundred warnings for ties.
+> ...we used the Wilcoxon Rank Sum Test [27] as implemented in R ...
+
+```{r}
+# wilcox.test gives a few hundred warnings for ties:
 # Warning: cannot compute exact p-value with ties
-wilcoxon_p = suppressWarnings(apply(datamatrices[["real_combat_covariates"]], 1, function(x) { wilcox.test(x[sampleannotation$covariate=="DP"], x[sampleannotation$covariate=="N"])$p.value } )
+wilcoxon_p = suppressWarnings(apply(datamatrices[["real_combat_covariates"]], 1, 
+                                    function(x) { wilcox.test(x[sampleannotation$covariate=="DP"],
+                                                              x[sampleannotation$covariate=="N"])$p.value } )
 )
 ```
+
 ```{r dev='png', fig.width=8, fig.height=8, fig.show='hold', tidy=FALSE}

### 2014-03-17 reanalysis/test/batch_adjust_no_covariates_can_induce_group_diff.rmd

========================================================

Show that adjusting for batch without trying to retain group differeces also can induce group effects.

```{r}
library(sva)
source("../../commonscripts/helper_functions.r")
```

Creating a unbalances experiment, 3 conditions, 2 batches with some genes true different in condition 3 compared to 1 and 2. But 1 and 2 have no difference. No batch effect added but adjusted for anyways.
```{r}
n=10
ngenes=5000
ndiffgenes = round(ngenes/10)
sampleannotation = createsampleannotation(  list(c(n,n,0), c(n,0,n))) 
#sampleannotation = createsampleannotation(  list(c(n,1), c(1,1), c(1,n))) 
randomdata = createrandomdata(ngenes, sampleannotation, mean=0, sd=1)
conditiondata = addconditioneffect(df=randomdata, labels=sampleannotation$treatment, ndiffgenes=ndiffgenes, thismean=10, betweensd=1, insidesd=1, affectedconditions="treatment3")
print(table(sampleannotation[, c("batch", "treatment")]))
```

Then run ComBat without covariates.
```{r}
batcnnormdata = ComBat(dat=conditiondata, 
                       batch=sampleannotation$batch, 
                       mod=NULL, par.prior=TRUE, prior.plots=FALSE)

```

And plot p-values
```{r pvaluespermuted, dev='svg', fig.width=10, fig.height=10, tidy=FALSE}
ret=getdifftab_limma(batcnnormdata, sampleannotation$treatment, "treatment1-treatment2")
#table(ret$padjusted<0.05)
hist(ret$p, breaks=100, border="red")

ret=getdifftab_limma(conditiondata, sampleannotation$treatment, "treatment1-treatment2", block=sampleannotation$batch)
#table(ret$padjusted<0.05)
hist(ret$p, breaks=100, add=T, border="black")

a = sampleannotation$batch==1
ret=getdifftab_limma(conditiondata[,a], sampleannotation$treatment[a], "treatment1-treatment2")
#table(ret$padjusted<0.05)
hist(ret$p, breaks=100, add=T, border="blue")
legend("topright", legend=c("ComBat adjusted", "Limma adjusted", "Only batch 1, no adjustment"), text.col=c("red", "black", "blue"))

```
Treatment1 and Treatment2 differs. Some of the effect in Treatment3 has been transfered overt to Treatment1 since they both are in batch2.
This problem is not limited to ComBat.

```{r}
library(pamr)

pamrnnormdata = pamr.batchadjust(list(x=conditiondata,
                                      y=factor(sampleannotation$treatment),
                                      batchlabels=factor(sampleannotation$batch)))
ret=getdifftab_limma(pamrnnormdata[["x"]], sampleannotation$treatment, "treatment1-treatment2")
#table(ret$padjusted<0.05)
hist(ret$p, breaks=100, border="red")

```

### 2014-03-17 docs/article.rmd
fra imac

@@
 Swapping the real data with random numbers from a normal distribution (mean=0, sd=1), but otherwise following the instructions, will generate the p-value distribution shown in Figure 1.
-The skewed distribution is for us a clear indication that this approach may have a unintentional adverse effect.
+The skewed distribution is a indication that this approach may have a unintentional adverse effect.
 _Jeg har ikke beskrevet permutatsjon checken fordi jeg er usikker paa om vi skal bruke den._
@@
 ## Examples of undesired consequences
-As the amount of false positive results when trying to retain group differences depends on the group/batch balance, we will show two examples with varying degree of unbalancedness.
+As the amount of false positive results when trying to retain group differences depends on the batch/group balance, we will show two examples with varying degree of unbalancedness.
 
@@
 Nevertheless, we belive that  P-values are deflated for the ComBat adjusted analysis.
+The R-code for our analysis and a more extensive report can be downloaded from github (https://github.com/vegardny/combat_tests.git).
 

### 2014-03-19 reanalysis/Fadi_Towfic_generic_medicine/towfic_et_al.rmd
download data

@@
 
-```{r, results='hide', echo=FALSE}
-starttime = Sys.time()
-```
-`r as.character(starttime)`
+`r as.character(Sys.time())`
 
@@
 - Perform a few more sanity checks to substantiate that the difference in results for the two above analysis is mainly introduced by ComBat
+
+```{r runparameters, results='hide', echo=FALSE}
+starttime = Sys.time()
+debug = FALSE
+downloaddata=TRUE
+```
 
@@
 
-Table S1 does have a "Chip"- coloumn, unfortunatly there is no dedicated "treatment"-coloumn.
+Table S1 does have a "Chip"- column, unfortunately there is no dedicated "treatment"-column.
 Communication with the corresponding author yielded this explanation
-> ..and the only thing we used as a covariate was the unique treatment names themselves (e.g. ³Medium² or ³RS").
+> ..and the only thing we used as a covariate was the unique treatment names themselves (e.g. "Medium"" or "RS").
 
@@
 ```
-The "covariate"-coloumn is made based on the "code"-coloumn, but might not match  to what were actually used for all samples.
-The last two coloumns are from the GEO deposit and reveal that 3 of the samples do not have data and are taken out.
+The "covariate"-column is made based on the "code"-column, but might not match  to what were actually used for all samples.
+The last two columns are from the GEO deposit and reveal that 3 of the samples do not have data and will be removed from the sampleannotation.
 
-The nameing convention seeme to differ slightly between the Table S1 and the text in Towfic et al. This is our interpretation of covariate labels and its corresponding name in the text.
-- **DP** refferd to as GA (but not GA as in table S1)
-- **N** refferd to as "generic"
-- **M** reffered to as "medium"
-- **RS** reffered to as "reference standard""
+The naming convention seem to differ slightly between the Table S1 and the text in Towfic et al. This is our interpretation of covariate labels and its corresponding name in the text.
+- **DP** referred to as GA (but not GA as in table S1)
+- **N** referred to as "generic"
+- **M** referred to as "medium"
+- **RS** referred to as "reference standard""
 
@@
 
-The processed data from the above GEO deposit was from another publication which did not processes the data as described in Towfic et al.
+The processed data from the above GEO deposit ("Series Matrix File") is from another publication which did not processes the data as described in Towfic et al.
 (personal communication).
-The lesser processed linked to as [GSE40566_non_normalized.txt.gz](http://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE40566&format=file&file=GSE40566%5Fnon%5Fnormalized%2Etxt%2Egz) in the supplementary data was thus used in this report, although it is not confirmed that this is the version of the data actually used in Towfic et al.
+The lesser processed data in GEO linked to as [GSE40566_non_normalized.txt.gz](http://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE40566&format=file&file=GSE40566%5Fnon%5Fnormalized%2Etxt%2Egz) in the "Supplementary file" section was thus used in this report.
 
-Probe annotation was also found in the supplementary data as [GSE40566_RAW.tar](http://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE40566&format=file).
+Probe annotation was also found in the "Supplementary file" section as [GSE40566_RAW.tar](http://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE40566&format=file).
 
 A recap of the important input files used in this report:
-- **sampleannotation.csv** holds the important sample/batch/covariate assignment. Compiled based on inforamtion in Towfic et al, the geo deposit and personal communication.
-- **GSE40566_non_normalized.txt** the measurments in a sample vs probe matrix.
+- **sampleannotation.csv** holds the important sample/batch/covariate assignment. Compiled based on information in Towfic et al, the geo deposit and personal communication.
+- **GSE40566_non_normalized.txt** the measurements in a sample vs probe matrix.
 From the [GEO deposit](http://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE40566&format=file&file=GSE40566%5Fnon%5Fnormalized%2Etxt%2Egz).
-- **GPL6887_MouseWG-6_V2_0_R3_11278593_A.txt** the probe annotation needed for the datarow to probe to gene matching.
+Not in github.
+- **GPL6887_MouseWG-6_V2_0_R3_11278593_A.txt** the probe annotation needed for the data row to probe to gene matching.
 From the GEO deposit inside the [GSE40566_RAW.tar](http://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE40566&format=file).
+Not in github.
 
@@
 ```{r}
-# the output from Illumina software.
-rawdata = read.table("not_in_github/GSE40566_non_normalized.txt", sep="\t", header=TRUE, stringsAsFactors=FALSE)
+
+if(downloaddata)
+{
+  temp <- tempfile()
+  download.file(url="http://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE40566&format=file&file=GSE40566%5Fnon%5Fnormalized%2Etxt%2Egz",
+              destfile=temp, mode = "wb")
+  rawdata <- read.table(temp, sep="\t", header=TRUE, stringsAsFactors=FALSE)
+  unlink(temp)
+
+  temp <- tempfile()
+  download.file(url="http://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE40566&format=file",
+              destfile=temp, mode = "wb")
+  tardirtemp = tempdir()
+  untar(temp, exdir = tardirtemp)
+  rawannotation = read.table(paste(tardirtemp ,"/GPL6887_MouseWG-6_V2_0_R3_11278593_A.txt.gz", sep=""), 
+                             sep="\t", header=TRUE, stringsAsFactors=FALSE, 
+                             skip=8, comment.char="", quote="", fill=TRUE)
+
+  unlink(temp)
+  unlink(tardirtemp, recursive=TRUE )
+}else{
+
+  # if download did not work change downloaddata to FALSE
+  # download data from the GEO deposit
+  # unpack and place files in a folder named "not_in_github"
+  rawdata = read.table("not_in_github/GSE40566_non_normalized.txt", 
+                       sep="\t", header=TRUE, stringsAsFactors=FALSE)
 
   # the probe annotation file found inside GSE40566_RAW.tar
-rawannotation = read.table("not_in_github/GPL6887_MouseWG-6_V2_0_R3_11278593_A.txt", sep="\t", header=TRUE, stringsAsFactors=FALSE, skip=8, comment.char="", quote="", fill=TRUE)
+  rawannotation = read.table("not_in_github/GPL6887_MouseWG-6_V2_0_R3_11278593_A.txt", 
+                             sep="\t", header=TRUE, stringsAsFactors=FALSE, 
+                             skip=8, comment.char="", quote="", fill=TRUE)
+}
 ```
 
-Some formatting is needed to match the probe information to the measurments and the measurments to the sample information.
+Some formatting is needed to match the probe information to the measurements and the measurements to the sample information.
 ```{r}
@@
 We are aiming to reproduce two of these.
-Unfortunatly they were only presented in a pdf-format not easely parsed by a computer.
-We had to resort to a ad-hoc method of cut-and-paste from pdf into text files which were somewhat polluted by pdf-formatting code.
+Unfortunately they were only presented in a pdf-format not easily parsed by a computer.
+We had to resort to a adhoc method of cut-and-paste from pdf into text files which were somewhat polluted by pdf-formatting code.
 For some tables about 10% of the rows will be lost, but the rest will suffice for our purpose.
@@
 Towfic et al.
-performed many different test on the data and it proved difficult and to labouros to reproduce all of their results without any source code.
-Insted we foucus on the key part of testing for differentially expressed genes between "GA" (DP) and "generic"(N) as described in Table S2 and Table S5.
-But before those tests we have to pre-process according to the description.
+performed many different test on the data and it is outside the scope of this report to reproduce all of their results.
+We focus on the key part of testing for differentially expressed genes between "GA" (DP) and "generic"(N) as described in Table S2 and Table S5.
+But before those tests we have to preprocess according to the description.
 
@@
 ```
-The numbers in Table S2 is almost reproduced.
+Original values from table S2 on X-axis, reporduced values on Y-axis. The values in Table S2 is almost reproduced.
 
@@
 
-There must be a typo in the figure text, they ment "Comparison of expression in GA to expression in **generic** for each probe, ...".
+There must be a typo in the figure text, they meant "Comparison of expression in GA to expression in **generic** for each probe, ...".
 
@@
 The plot reveals that the reproduced fold changes are much more compressed than the ones in Table S5.
-There seems to two linear relationship, one for neagtive FC's and another for positve FC's. They seems to have different intercepts but comparable slope.
+There seems to two linear relationship, one for negative FC's and another for positive FC's. They seems to have different intercepts but comparable slope.
 
@@
 
-The T-test is stright forward.
+The T-test is straight forward.
 ```{r dev='png', fig.width=10, fig.height=6, fig.show='hold', tidy=FALSE}
@@
 The T-test can not handle batch inside the test.
-An alternative way of performing a t-test is to subset the data and do the comparison whithin each batch.
+An alternative way of performing a t-test is to subset the data and do the comparison within each batch.
 For each of the 7 batches that have both a DP ("GA") and a N ("generic") sample, we can do a direct comparison and in effect ending up with 7 samples.

### 2014-03-24 reanalysis/Fadi_Towfic_generic_medicine/old_rmd/towfic_et_al_tomuch_again.rmd
figure

--- 2014-03-19 reanalysis/Fadi_Towfic_generic_medicine/towfic_et_al.rmd
@@
 - Reproduce some of the results to show that we are working on the same data and analysis workflow
-- Remove the use of ComBat and perform an similar analysis with alternative established tools
+- Remove the use of ComBat and perform a similar analysis with alternative established tools
 - Estimate the error introduced by ComBat and the consequences for the conclusion of the study
@@
 starttime = Sys.time()
-debug = FALSE
+debug = TRUE
 downloaddata=TRUE
 ```
-
-### Getting the data
 
@@
 ```
+
+### Getting the data
+
 The important sample information is described in Table S1 from Towfic et al. and its usage in ComBat is described briefly in the methods;
@@
 ```
-The "covariate"-column is made based on the "code"-column, but might not match  to what were actually used for all samples.
-The last two columns are from the GEO deposit and reveal that 3 of the samples do not have data and will be removed from the sampleannotation.
+The "covariate"-column is made based on the "code"-column, but might not match 100% to what were actually used for all samples.
+The last two columns are from the GEO deposit and reveal the samples that do not have data, which we will remove from the data shortly.
 
-The naming convention seem to differ slightly between the Table S1 and the text in Towfic et al. This is our interpretation of covariate labels and its corresponding name in the text.
+The naming convention seem to differ slightly between the Table S1 and the text in Towfic et al. This is our interpretation of the main covariate labels and its corresponding name in the text.
 - **DP** referred to as GA (but not GA as in table S1)
@@
 ```
-A look at the most interesting groups, "DP"(GA) and "N"(generic) reveals a lack of balance.
+A look at the primary comparison, "DP"(GA) and "N"(generic) reveals a lack of balance.
 ```{r}
@@
 {
-  temp <- tempfile()
+  temp = tempfile()
   download.file(url="http://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE40566&format=file&file=GSE40566%5Fnon%5Fnormalized%2Etxt%2Egz",
               destfile=temp, mode = "wb")
-  rawdata <- read.table(temp, sep="\t", header=TRUE, stringsAsFactors=FALSE)
+  rawdata = read.table(temp, sep="\t", header=TRUE, stringsAsFactors=FALSE)
   unlink(temp)
 
-  temp <- tempfile()
+  temp = tempfile()
   download.file(url="http://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE40566&format=file",
@@
 datamatrices[["real_raw"]] = datamatrix_raw#[1:1000,] # debug
+if(debug)
+  datamatrices[["real_raw"]] = datamatrix_raw[1:1000,]
 
@@
 par(mfrow=c(2, 2))
-plot(table_s2$AVG_Medium, meanM[table_s2$ID], pch=".")
-plot(table_s2$AVG_Reference_Standard, meanRS[table_s2$ID], pch=".")
-plot(table_s2$AVG_GA, meanDP[table_s2$ID], pch=".")
-plot(table_s2$AVG_generic, meanN[table_s2$ID], pch=".")
+plot(table_s2$AVG_Medium, meanM[table_s2$ID], pch=".", na.rm=T)
+plot(table_s2$AVG_Reference_Standard, meanRS[table_s2$ID], pch=".", na.rm=T)
+plot(table_s2$AVG_GA, meanDP[table_s2$ID], pch=".", na.rm=T)
+plot(table_s2$AVG_generic, meanN[table_s2$ID], pch=".", na.rm=T)
 ```
@@
 How the fold change was calculated is not explained. A common way is to take the difference of geometric means that were calculated for Table S2 above.
-```{r dev='png', fig.width=8, fig.height=8, fig.show='hold', tidy=FALSE}
+```{r}
 foldchange_DP_N = meanDP - meanN
+```
 
- # make the axis equal.
-lim =c( min(   c(min(table_s5$Fold_Change, na.rm=T), min(foldchange_DP_N[table_s5$Probe], na.rm=T))  ), 
-        max( c(max(table_s5$Fold_Change, na.rm=T), max(foldchange_DP_N[table_s5$Probe], na.rm=T)) ))
+These reproduced foldchanges can be compared for the probes listed in table S5.
 
-# plot
+```{r dev='png', fig.width=10, fig.height=5, fig.show='hold', tidy=FALSE}
+par(mfrow=c(1, 2))
 plot(table_s5$Fold_Change, foldchange_DP_N[table_s5$Probe],
-     ylim=lim, xlim=lim, pch=".", xlab="Table_S5", ylab="reproduced",
+     pch=".", xlab="Original Table S5", ylab="Reproduced",
      main="FC comparison")
-rm(lim)
+
+thiscolors = c("black", "red")
+hist(foldchange_DP_N, breaks=100, freq=F, border=thiscolors[1], main="Reproduced FC for DP vs. N")
+hist(foldchange_DP_N[table_s5$Probe], breaks=100, add=T, freq=F, border=thiscolors[2])
+legend("topright", legend=c("All probes", "Table S5 probes"), text.col=thiscolors)
 ```
-The plot reveals that the reproduced fold changes are much more compressed than the ones in Table S5.
-There seems to two linear relationship, one for negative FC's and another for positive FC's. They seems to have different intercepts but comparable slope.
+The first plot reveals that the reproduced fold changes are much more compressed than the ones in Table S5.
+There seems to two linear relationship, one for negative FC's and another for positive FC's.
+They seems to have different intercepts but comparable slope.
+The second plot shows that the reproduced fold changes for the probes listed in Table S5 are clearly more extreme than the rest.
 
-The average expression for "GA" (DP) and "generic" (N) are also given in Table S2. The fold change calculated from those numbers compared to the ones given in Table S5 shows a similar pattern:
+The fold changes in table 5S do not match the average expression values listed in table S2 either as a the same plot will show;
 
@@
 ```
+There seems to be some inconsistancy between between the fold changes in table S5 and the reproduced ones, however it is the conficende i.e p-values that are affected by the use of ComBat.
+The rest of table S5 is p-values and we will ty and reproduce some of them.
 
-```{r dev='png', fig.width=10, fig.height=5, fig.show='hold', tidy=FALSE}
-par(mfrow=c(1, 2))
-plot(table_s5$Fold_Change, foldchange_DP_N[table_s5$Probe],
-     pch=".", xlab="Original Table S5", ylab="Reproduced",
-     main="FC comparison")
-
-thiscolors = c("black", "red")
-hist(foldchange_DP_N, breaks=100, freq=F, border=thiscolors[1], main="Reproduced FC for DP vs. N")
-hist(foldchange_DP_N[table_s5$Probe], breaks=100, add=T, freq=F, border=thiscolors[2])
-legend("topright", legend=c("All probes", "Table S5 probes"), text.col=thiscolors)
-```
-
-The reproduced fold changes for the probes listed in Table S5 are clearly more extreme than the rest.
-
-The next test is a test for differentially expressed genes using ANOVA.
+ANOVA is The first test for differentially expressed gene.
 >  First, we computed the statistical significance of differential expression between treatments using the Analysis of Variance (ANOVA) method for each probe [21]...<cite> Towfic et al.

### 2014-03-25 reanalysis/ComBat_original/combat_paper_dataset2_reproduced.rmd
figure
mer

--- 2014-03-17 reanalysis/ComBat_original/combat_paper_dataset2_reproduced.rmd
@@
 ### Overview
-This report aims to show to what extent the use of ComBat led to false results in the second analysis example given in [Johnson et al.](http://biostatistics.oxfordjournals.org/content/8/1/118.abstract) The example named "Data Set 2" and the analysis are described in the [supplementary material](http://biostatistics.oxfordjournals.org/content/suppl/2006/04/21/kxj037.DC1/kxj037supp.pdf) for Johnson et al.
+This report aims to show to what extent the use of ComBat led to false results in the second analysis example given in [Johnson et al.](http://biostatistics.oxfordjournals.org/content/8/1/118.abstract) The example named "Data Set 2" and the analysis is described in the [supplementary material](http://biostatistics.oxfordjournals.org/content/suppl/2006/04/21/kxj037.DC1/kxj037supp.pdf) for Johnson et al.
 Description on how to obtain the data were found [here](https://groups.google.com/forum/#!msg/combat-user-forum/S9vBcXw8RGk/QIoD6oRBM9IJ) which had links to [dataExample2.txt](http://www.bu.edu/jlab/wp-assets/ComBat/data/dataExample2.txt)
@@
 - Estimate the error introduced by ComBat and the consequences for the conclusion of the study
-- Perform a few more sanity checks to substantiate that the difference in results for the two above analysis is mainly due to error introduced by ComBat
+- Perform a few more sanity checks to substantiate that the difference in results for the two above analyses is mainly false and introduced by ComBat
 
@@
 ```
+Three batches and two covariates. There is a "Cell" column in the sample annotation, but it does not seem to have been used.
 
@@
 ```
-Number of differentially expressed probes found with the alternative anaysis is **`r sum(qvalue(datamatrix_limma_res$p.value[,1])$qvalue<0.05)`**.
+Number of differentially expressed probes found with the alternative analysis is **`r sum(qvalue(datamatrix_limma_res$p.value[,1])$qvalue<0.05)`**.
 
@@
 ```
-And if they were to use the top 1000 probes in a gene set test, they would have found many of the same genes regardless of hadling batch effects by ComBat or limma.
+And if they were to use the top 1000 probes in a gene set test, they would have found many of the same genes regardless of handling batch effects by ComBat or limma.
 ```{r, tidy=FALSE}
@@
 ```
-The p-values for ComBat adjusted random numbers are slightly enriched for low p-values. Indicating that the same enrichment seen in the real data is also false.
+The p-values for ComBat adjusted random numbers are slightly enriched for low p-values. Indicating that some of the enrichment seen in the real data is also false.
 
@@
 ```
-Then adjust the data with fictive batches and calculate p-values as done for the unadjusted batch 3.
+Then adjust the data with fictive batches and calculate p-values as done for the un-adjusted batch 3.
 ```{r, results='hide', tidy=FALSE}

### 2014-03-25 reanalysis/Fadi_Towfic_generic_medicine/towfic_et_al.rmd
mer

@@
 starttime = Sys.time()
-debug = FALSE
+debug = TRUE
 downloaddata=TRUE
+set.seed(100)
 ```
-
-### Getting the data
 
@@
 ```
+There seems to be some inconsistancy between between the fold changes in table S5 and the reproduced ones, however it is the conficende i.e p-values that are affected by the use of ComBat.
 
-```{r dev='png', fig.width=10, fig.height=5, fig.show='hold', tidy=FALSE}
-par(mfrow=c(1, 2))
-plot(table_s5$Fold_Change, foldchange_DP_N[table_s5$Probe],
-     pch=".", xlab="Original Table S5", ylab="Reproduced",
-     main="FC comparison")
-
-thiscolors = c("black", "red")
-hist(foldchange_DP_N, breaks=100, freq=F, border=thiscolors[1], main="Reproduced FC for DP vs. N")
-hist(foldchange_DP_N[table_s5$Probe], breaks=100, add=T, freq=F, border=thiscolors[2])
-legend("topright", legend=c("All probes", "Table S5 probes"), text.col=thiscolors)
-```
-
-The reproduced fold changes for the probes listed in Table S5 are clearly more extreme than the rest.
-
-The next test is a test for differentially expressed genes using ANOVA.
->  First, we computed the statistical significance of differential expression between treatments using the Analysis of Variance (ANOVA) method for each probe [21]...<cite> Towfic et al.
-
-```{r}
-anova_p = apply(datamatrices[["real_combat_covariates"]], 1,
-              function(x){ summary(aov(x[sampleannotation$covariate %in% c("DP", "N")]~
-                                         sampleannotation$covariate[sampleannotation$covariate %in% c("DP", "N")]))[[1]][1,5]})
-
-par(mfrow=c(1, 2))
-plot(table_s5[,"anova_p"], anova_p[table_s5$Probe], xlab="Original Table S5", 
-     ylab="Reproduced", main="ANOVA P-values from table S5 vs. reproduced")
-thiscolors = c("black", "red")
-hist(anova_p, breaks=100, freq=T, border=thiscolors[1], main="Reproduced ANOVA p-values for DP vs. N")
-hist(anova_p[table_s5$Probe], breaks=100, add=T, freq=T, border=thiscolors[2])
-legend("topright", legend=c("All probes", "Table S5 probes"), text.col=thiscolors)
-```
+The rest of table S5 is p-values and we will try to reproduce the ones from Limma.
 
@@
 ```
-> In parallel, we used Comparative Marker Selection as implemented in GenePattern [26] to directly compare probes between generic and GA.
-We applied two separate techniques within this framework; a traditional T-test and a Signal-to-Noise Ratio test (SNR).
 
-The T-test is straight forward.
-```{r dev='png', fig.width=10, fig.height=6, fig.show='hold', tidy=FALSE}
-ttest_p = apply(datamatrices[["real_combat_covariates"]], 1, 
-                function(x) { t.test(x[sampleannotation$covariate=="DP"], 
-                                     x[sampleannotation$covariate=="N"])$p.value } )
-par(mfrow=c(1, 2))
-plot(table_s5[,"t_test_p"], ttest_p[table_s5$Probe], xlab="Original Table S5", 
-     ylab="Reproduced", main="t-test P-values from table S5 vs. reproduced")
-thiscolors = c("black", "red")
-hist(ttest_p, breaks=100, freq=T, border=thiscolors[1], main="Reproduced t-test p-values for DP vs. N")
-hist(ttest_p[table_s5$Probe], breaks=100, add=T, freq=T, border=thiscolors[2])
-legend("topright", legend=c("All probes", "Table S5 probes"), text.col=thiscolors)
-```
-
-> ...we used the Wilcoxon Rank Sum Test [27] as implemented in R ...
-
-```{r dev='png', fig.width=10, fig.height=6, fig.show='hold', tidy=FALSE}
-# wilcox.test gives a few hundred warnings for ties:
-# Warning: cannot compute exact p-value with ties
-wilcoxon_p = suppressWarnings(
-  apply(datamatrices[["real_combat_covariates"]], 1, 
-        function(x) { wilcox.test(x[sampleannotation$covariate=="DP"],
-                                  x[sampleannotation$covariate=="N"])$p.value } ))
-par(mfrow=c(1, 2))
-plot(table_s5[,"wilcoxon_p"], wilcoxon_p[table_s5$Probe], xlab="Original Table S5", 
-     ylab="Reproduced", main="wilcoxon P-values from table S5 vs. reproduced")
-thiscolors = c("black", "red")
-hist(wilcoxon_p, breaks=100, freq=T, border=thiscolors[1], 
-     main="Reproduced wilcoxon p-values for DP vs. N")
-hist(wilcoxon_p[table_s5$Probe], breaks=100, add=T, freq=T, border=thiscolors[2])
-legend("topright", legend=c("All probes", "Table S5 probes"), text.col=thiscolors)
-
-```
+In addition to the Limma test, Towfic et al. performed similar tests for significant probes between generic and GA using several other methods. 
+We tried several of them as well, but have skipped them in this report since they showed a similar patterent as for the Limma p-values.
 
@@
 An alternative way of handling batch effect compared to ComBat's is to include it in the statistical test.
-This can be done in ANOVA.
+This can be done in Limma
 
-```{r tidy=FALSE}
-
-DP = sampleannotation$covariate=="DP"
-N = sampleannotation$covariate=="N"
-thisdata =datamatrices[["real_qnorm"]]
-cov_fac = factor(sampleannotation$covariate[DP|N])
-chip_fac = factor(sampleannotation$chip[DP|N])
-#anova_p1 = apply(thisdata, 1, function(x){ summary(aov(x[DP|N]~chip_fac*cov_fac))[[1]][2,5]}  ) 
-anova_woc = apply(thisdata, 1, function(x){ summary(aov(x[DP|N]~chip_fac+cov_fac))[[1]][2,5]}  ) 
-#co = list(cov_fac=contr.sum, chip_fac=contr.sum)
-#anova_p3 = apply(thisdata, 1, function(x){ Anova(lm(x[DP|N] ~  cov1_fac + chip_fac,  contrasts=co), type=3)[[4]][2]}  ) 
-table(p.adjust(anova_p , method="fdr")<0.05, p.adjust(anova_woc , method="fdr")<0.05, dnn=c("ComBat adj", "ANOVA adj"))
-```
-The number of significant probes using the given FDR threshold of 0.05 when batch effect is handled by ANOVA is 
-**`r sum(p.adjust(anova_woc , method="fdr")<0.05)`**. This is much lower compared to when ComBat is applied, 
-**`r sum(p.adjust(anova_p , method="fdr")<0.05)`**.  The p-value distribution is still skewed, but not as much.
-```{r dev='png', fig.width=8, fig.height=8, fig.show='hold', tidy=FALSE}
-thiscolors = c("black", "red")
-hist(anova_p, breaks=100, freq=T, border=thiscolors[1], main="Reproduced ANOVA p-values for DP vs. N")
-hist(anova_woc, breaks=100, add=T, freq=T, border=thiscolors[2])
-legend("topright", legend=c("ComBat adjusted", "ANOVA adjusted"), text.col=thiscolors)
-```
-
-LIMMA also has the possibility to handle batch inside the statistical test.
 ```{r dev='png', fig.width=8, fig.height=8, fig.show='hold', tidy=FALSE}
@@
 
-The T-test can not handle batch inside the test.
-An alternative way of performing a t-test is to subset the data and do the comparison within each batch.
-For each of the 7 batches that have both a DP ("GA") and a N ("generic") sample, we can do a direct comparison and in effect ending up with 7 samples.
-Only 7 samples will give less power but the caveats because of batch effect and adjustments are also far less.
-```{r tidy=FALSE, results='hold'}
-
-# subsetting the samples so that they have a balanced conditon distribution across batches. i.e each batch has the same compitition of conditions (example 70% treated vs 30% untreated)
-drawbatchbalanceddsamples = function(samplenames, batches, conditions)
-{
-  # 1. find the minimal count for each condition across batches.
-	# 2 find the minimal given count / minimal count inside each batch
-	# 3 If greater than 1, scale up all conditions for that batch.
-	# 4 draw samples randomly from the above calculated condition-count for each batch
-	# 5 create and return the data subset.
-
-	#1
-	starttab = table( data.frame(conditions, batches ))	
-	tab = starttab[, apply(starttab, 2,  min)>0] # do not use batches with missing conditions
-	if(dim(tab)[2]==0)
-	{
-		warning(  paste("Unable to find a balanced subset that has at least one of each condition. Remove one or more conditions and try again")) 
-		return(vector())
-	}
-	if(dim(tab)[1]<dim(starttab)[1])
-	{
-		warning(  paste("Lost batches during balancing due to missing conditions in those.
-",   "Tried: ", paste( dimnames(starttab)[[2]], collapse=" "), "Got:", paste( dimnames(tab)[[2]], collapse=" "  ) ))
-	}
-
-	mincond = apply(tab, 1,  min)
-
-	#2
-	tab2 = mincond/tab
-
-	#3
-	scalings = apply( 1/tab2, 2, min)
-	scaledtab = tab
-	scaledtab[,] = mincond	
-	scaledtab = t(t(scaledtab) * scalings)
-	scaledtab = round(scaledtab,0)
-
-	winnersamples= vector()
-	for(c in dimnames(scaledtab)[[1]])
-	{
-		for(b in dimnames(scaledtab)[[2]])
-		{
-			a = batches==b & conditions==c
-			winnersamples = c( winnersamples, sample( samplenames[a], scaledtab[c,b]))
-		}
-	}
-	return(winnersamples)	
-}
-
-a = sampleannotation$covariate %in% c("N", "DP")
-nbalanceddatasets = 3
-pairedpvals = list()
-for(i in 1:nbalanceddatasets)
-{
-  bs = drawbatchbalanceddsamples(sampleannotation$code[a], sampleannotation$chip[a], sampleannotation$covariate[a])
-  print(table(sampleannotation[bs,c("chip", "covariate")])) 
-  bs = bs[order(sampleannotation[bs, "covariate"], sampleannotation[bs, "chip"])]
-  sampleannotation[bs, c("covariate", "chip")] # the samples are ordered
-  thispaireddata = datamatrices[["real_qnorm"]][,bs][,1:7] -  datamatrices[["real_qnorm"]][,bs][,8:14]
-  pairedpvals[[i]] = apply(thispaireddata, 1, 
-                function(x) { t.test( x)$p.value } )
-  print(table(p.adjust(pairedpvals[[i]] , method="fdr")<0.05))
-  rm(bs,thispaireddata)  
-}
-```
-
-```{r dev='png', fig.width=8, fig.height=8, fig.show='hold', tidy=FALSE}
-par(mfrow=c(2, 2))
-hist(ttest_p, breaks=100, freq=T, border="black", main="P-values t-test on ComBat adjusted data")
-for(i in 1:length(pairedpvals))
-{
-  hist(pairedpvals[[i]], breaks=100, freq=T, border="black", main=paste("P-values paired t-test draw ", i, sep=""))
-}
-
-```
-
 ### Additional sanity checks
 
+To substantiate that the result from the use of ComBat is less trustworthy than the alternative analysis we provide a couple of additional sanity checks.
+
+First we use random numbers drawn from the same distribution(mean=0, sd=1) regardless of batch or covariate but retaining the batch/covariate design.
 ```{r}
 set.seed(100)
-datamatrices[["random_raw"]] = matrix(rnorm(length(datamatrices[["real_raw"]])), 
+datamatrices[["random_raw"]] = matrix(rnorm(length(datamatrices[["real_raw"]]), mean=0, sd=1), 
                                       nrow=nrow(datamatrices[["real_raw"]]), 
@@
 
+Limma is then used in 3 ways
+- On the ComBat adjusted random numbers
+- On the random numbers ignoring batch information
+- On the random numbers including batch as a blocking factor
+
 ```{r}
@@
 design = model.matrix(~0 + group)
+cont.matrix = makeContrasts ( contrasts="groupDP-groupN", levels=design)
+
 fit = lmFit(datamatrices[["random_combat_covariates"]], design)  
+fit2 = contrasts.fit(fit, cont.matrix)
+limma_p_rand_combat = eBayes(fit2)$p.value[,1]
+
+fit = lmFit(datamatrices[["random_raw"]], design)  
+fit2 = contrasts.fit(fit, cont.matrix)
+limma_p_rand_nocombat = eBayes(fit2)$p.value[,1]
+
+block = as.factor(sampleannotation$chip)
+design = model.matrix(~0+group+block)
 cont.matrix = makeContrasts ( contrasts="groupDP-groupN", levels=design)
+fit = lmFit(datamatrices[["random_raw"]], design)  
 fit2 = contrasts.fit(fit, cont.matrix)
-limma_ret_rand = eBayes(fit2)
-limma_p_rand = limma_ret_rand$p.value
-names(limma_p_rand) = names(limma_ret_rand$Amean)
-table(p.adjust(limma_p_rand, method="fdr")<0.05)
+limma_p_rand_batchblocked = eBayes(fit2)$p.value[,1]
+
 ```
@@
 par(mfrow=c(1, 1))
-thiscolors = c("black", "red")
-hist(limma_p_alt, breaks=100, freq=T, border=thiscolors[1], main="LIMMA p-values for DP vs. N")
-hist(limma_p_rand, breaks=100, add=T, freq=T, border=thiscolors[2])
-legend("topright", legend=c("Reproduced from Real data", "Random data ComBat adjusted"), text.col=thiscolors)
-```
-
-```{r}
-
-anova_p_rand = apply(datamatrices[["random_combat_covariates"]], 1,
-              function(x){ summary(aov(x[sampleannotation$covariate %in% c("DP", "N")]~
-                                         sampleannotation$covariate[sampleannotation$covariate %in% c("DP", "N")]))[[1]][1,5]})
-table(p.adjust(anova_p_rand, method="fdr")<0.05)
-```
-
-```{r dev='png', fig.width=8, fig.height=8, fig.show='hold', tidy=FALSE}
-par(mfrow=c(1, 1))
-thiscolors = c("black", "red")
-hist(anova_p, breaks=100, freq=T, border=thiscolors[1], main="ANOVA p-values for DP vs. N")
-hist(anova_p_rand, breaks=100, add=T, freq=T, border=thiscolors[2])
-legend("topright", legend=c("Reproduced from Real data", "Random data ComBat adjusted"), text.col=thiscolors)
+thiscolors = c("red", "blue", "black")
+hist(limma_p_rand_combat, breaks=100, freq=T, border=thiscolors[1], main="P-values, Random numbers for DP vs. N")
+hist(limma_p_rand_nocombat, breaks=100, add=T, freq=T, border=thiscolors[2])
+hist(limma_p_rand_batchblocked, breaks=100, add=T, freq=T, border=thiscolors[3])
+legend("topright", legend=c("ComBat adjusted", "Limma adjusted", "No adjustment"), text.col=thiscolors)
 ```
@@
 
-Another sanity check is by permuting the DP and N labels and look how the result changes.
+Another sanity check is by permuting the DP (GA) and N(generic) labels and observe changes in the results.
 In order to retain the batch effect we will only swap labels within a batch.
@@
 
-```{r}
+```{r results='hide'}
 source("../../commonscripts/helper_functions.r")
-#shufflesamplesinbatch = function(samplenames, batch, covariate, shufflecovariates)
-#{
-  #ret = samplenames
-	#b = covariate %in% shufflecovariates
-	#for(thisbatch in unique(batch))
-	#{
-	#	a = batch== thisbatch		
-#		ret[a & b] = sample( samplenames[a&b],sum(a&b),  replace=FALSE)		
-	#}
-	#return(ret)
-#}
 
@@
 datamatrices_permuted = list()
+changedlabels=vector()
 for(i in 1:nshuffleddatasets)
@@
   x = shufflesamplesinbatch( sampleannotation$code, sampleannotation$chip, sampleannotation$covariate, c("N", "DP") )
-  print(  sum(sampleannotation[,"covariate"] != sampleannotation[x,"covariate"]  ))  # see how many got change. Has 11 N's)  
+  changedlabels[i] = (  sum(sampleannotation[,"covariate"] != sampleannotation[x,"covariate"]  ))  # see how many got change. Has 11 N's)  
   datamatrices_permuted[[i]] = as.matrix(ComBat(dat=datamatrices[["real_qnorm"]][, x], 
@@
 }
+```
+**`r nshuffleddatasets`** permutations were created and batch adjusted with ComBat.
 
+Then perform the limma test on the ComBat adjusted data as was done for the real data.
+```{r}
 permuted_pvalcounts=list()
+overviewtab = data.frame(permutation=1:nshuffleddatasets, changedlabels=changedlabels, significantprobes=NA)
 for(i in 1:nshuffleddatasets)
@@
   permuted_pvalcounts[[i]] = hist(this_p, plot=FALSE, breaks=100)$counts 
-  print(table(p.adjust(this_p, method="fdr")<0.05))
+  overviewtab$significantprobes[i] =sum(p.adjust(this_p, method="fdr")<0.05)
 }
+print(overviewtab)
+rm(overviewtab, group, design, fit, cont.matrix, fit2, this_p)
 ```
+The table lists the number of labels that are changed compared to the real ones for each permutation.
+Remember that the totat number of N (generic) samples is **`r sum(sampleannotation$covariate=="N")`**.
+So even when about half of those are swapped with DP(GA), the ComBat adjusted data produces many significant genes.
+The number of significant genes using the same 0.05 FDR cut-off for the real labels was **`r sum(p.adjust(limma_p_alt , method="fdr")<0.05)`**.
 
+The P-value distribution can slo be plotted.
 ```{r, dev='svg', fig.width=8, fig.height=8, tidy=FALSE}
@@
 ```
+Shuffling the N and DP labels whitinh batches produces about the same p-value distribution as the real data.
+This is an indication that the significant genes found in the real data is not associated with the N and DP labels, but a rather an effect of the ComBat adjustments.
 
-Forslag figur brukt i artikkel.
+Forslag 1 figur i artikkel.
 ```{r pvaluestowfic, dev='svg', fig.width=10, fig.height=10, tidy=FALSE}
@@
 # random ComBat + limma
-hist(limma_p_rand, main="P-values, Random data, ComBat followed by Limma", breaks=100, xlab="p-value" , ylim=thisylim)
+hist(limma_p_rand_combat, main="P-values, Random data, ComBat followed by Limma", breaks=100, xlab="p-value" , ylim=thisylim)
 
@@
        text.col=c("red", "blue"))
+```
+
+Forslag 2 figur i artikkel.
+```{r pvaluestowfic2, dev='svg', fig.width=10, fig.height=10, tidy=FALSE}
+# figure used in our article
+# NB! bruker orgniale data som ikke er floored eller log2!
+par(mfrow=c(1, 1))
+thiscolors = c("red", "blue", "black")
+thisylim = c(0, max(unlist( c(permuted_pvalcounts,real_pvalcounts))))
+xrange=1:20
+xrange=1:25
+# reproduced ComBat + limma
+ plot((xrange)/100, hist(limma_p_alt, breaks=100, plot=FALSE)$counts[xrange], 
+       main="P-values",
+       xlab="p-value", 
+       ylab="frequency",
+       ylim=thisylim,
+       type="l",
+       lwd=2,
+       col=thiscolors[1])
+
+# batch handled in limma
+lines((xrange)/100, hist(limma_p_woc, breaks=100, plot=FALSE)$counts[xrange], 
+     col=thiscolors[2], lwd=2)
+
+# random ComBat + limma
+lines((xrange)/100, hist(limma_p_rand_combat, breaks=100, plot=F)$counts[xrange],
+       col=thiscolors[3], lwd=2)
+
+legend("topright", legend=c("Real data, ComBat adjusted",
+                            "Real data, batch handled by Limma",
+                            "Random data, ComBat adjusted"),
+       text.col=thiscolors)
+
 ```

### 2014-03-27 reanalysis/ComBat_original/old_rmd/combat_paper_dataset2_reproduced_ttest.rmd
forenkling data set 2

--- 2014-03-25 reanalysis/ComBat_original/combat_paper_dataset2_reproduced.rmd
@@
 Three batches and two covariates. There is a "Cell" column in the sample annotation, but it does not seem to have been used.
-
+<br/>  
+<br/>  
 ### Reproduce the original results
@@
 ```
-The heatmap does not look exactly as Fig.
-A1 in the paper.
+The heatmap does not look exactly as [Fig.
+A.1](http://biostatistics.oxfordjournals.org/content/suppl/2006/04/21/kxj037.DC1/kxj037supp.pdf) in Johnson et al.
 This could be due to undocumented transformations preformed in Johnson et al., for example log-transformation or standardization.
 And there are different ways of calculating variation for a probe.
-What parameters that were used for the clustering could also be different.
+The parameters used in the clustering could also be different.
 But the purpose of the figure was to show that a batch effect is present in the data and especially for batch 3.
 This is also achieved in the reproduced figure.
-
+<BR/>  
+<BR/>  
 Next 3 data sets were made,
-- **EB2**: Batches 1 and 2 adjusted with non-parametric use of ComBat
-- **EB3**: All thee batches adjusted with non-parametric use of ComBat
+- **EB2**: Batches 1 and 2 adjusted with ComBat
+- **EB3**: All the batches adjusted with ComBat
 - **Batch3**: Only batch 3 (no adjustments)
@@
 ```
-Again the heatmap is not exactly as in Johnson et al, but the batch clustering is broken("Batch, bottom annotation row") and the samples cluster more by cell type("Cell", top annotation row) and treatment type("Type", middle annotation row).
+Again the heatmap is not exactly as in Johnson et al, but the batch clustering is broken(note, "Batch" is bottom annotation row!) and the samples cluster more by cell type("Cell", top annotation row) and treatment type("Type", middle annotation row).
 

### 2014-04-02 reanalysis/ComBat_original/combat_paper_dataset2_reproduced.rmd
figure script added

@@
 rm(tmp)
+debug=FALSE
 ```
-```{r, cache=TRUE, tidy=FALSE}
-datamatrix = as.matrix(read.table("data/dataExample2.txt", sep="\t", header=TRUE))#[1:3000,]#dev.
+```{r loaddata,  tidy=FALSE}
+datamatrix = as.matrix(read.table("data/dataExample2.txt", sep="\t", header=TRUE))
+if(debug)datamatrix = datamatrix[1:3000,]
 sampleannotation = read.table("data/sampleInfoExample2.txt", 
@@
 
-The reproduced numbers are not the same as the reported ones, but they are not completely off. Again this is likely to be due to undocumented steps performed by Johnson et al.
+The reproduced numbers are not the same as the reported ones, but they are not completely off. This is likely due to undocumented steps performed by Johnson et al.
+
+The above effort done in reproducing some of the results for data set 2 is done using non-log transformed data that also includes some negative values.
+In the next part of this report we will do an alternative analysis were batch is included in the analysis using limma instead of using ComBat.
+Since Limma needs log2 transformed data, the results will not be directly comparable to a t-test result using non-log values.
+Later we will also perform a few test using permutation (i.e many runs).
+For these reasons the rest of the analysis, including the significance test on ComBat adjusted data, will be performed on log2 transformed data with negative values set to 1 or removed (if in many samples for that probe).
+This is a fairly standard workflow and more practical for doing comparisons.
+In addition the parametric version of ComBat will be used.
+
+First we make a data matrix with log2 values. And doing so the negative values will be floored and removed if in 
+```{r}
+log2data = datamatrix
+
+# flooring to 1
+log2data[log2data<1]=1
+
+# take out data with to much low/missing values.
+negativeprobesfilter =( rowSums(log2data>1) >= (0.9*ncol(log2data)) )
+log2data = log2data[negativeprobesfilter,]
+
+# quantilenormalize
+log2data=normalizeBetweenArrays(log2(log2data))
+```
+
+```{r}
+
+log2dataComBat = as.matrix( sva::ComBat(
+            dat=log2data, 
+            batch=sampleannotation[colnames(log2data),"Batch"], 
+            mod=model.matrix(~as.factor(sampleannotation[colnames(log2data),"Type"])), 
+            numCovs=NULL, 
+            par.prior=useparprior, 
+            prior.plots=FALSE))
+
+Type = as.factor(sampleannotation$Type)
+design = model.matrix(~0 + Type)
+fit = lmFit(log2dataComBat, design)
+cont.matrix = makeContrasts ( contrasts="TypeR-TypeC", levels=design)  
+fit2 = contrasts.fit(fit, cont.matrix)
+log2ComBat_res = eBayes(fit2)
+rm(Type, design, fit, cont.matrix, fit2)
+print(table(qvalue(log2ComBat_res$p.value[,1])$qvalue<0.05))
+
+```
+The number of genes below the significance threshold is somewhat lower than for the original the t-test, but that comparison is not relevant since the data is processed in different ways before the test.
+But for the explained practical purposes this is the result coming from the ComBat adjusted data.
 
@@
 ```{r}
-negativeprobesfilter =( rowSums(datamatrix>=1) >= (0.5*ncol(datamatrix)) ) & 
-                      ( rowSums(EB3>=1) >= (0.5*ncol(EB3)) )
-thismat = datamatrix[negativeprobesfilter, ]
-thismat[thismat<1]=1
-thismat=normalizeBetweenArrays(log2(thismat))
 Type = as.factor(sampleannotation$Type)
@@
 design = model.matrix(~0+Type+Block)
-fit = lmFit(thismat, design)
+fit = lmFit(log2data, design)
 cont.matrix = makeContrasts ( contrasts="TypeR-TypeC", levels=design)	
 fit2 = contrasts.fit(fit, cont.matrix)
-datamatrix_limma_res = eBayes(fit2)
-rm(thismat, Type, Block, design, fit, cont.matrix, fit2)
-print(table(qvalue(datamatrix_limma_res$p.value[,1])$qvalue<0.05))
+log2limmablock_res = eBayes(fit2)
+rm(Type, Block, design, fit, cont.matrix, fit2)
+print(table(qvalue(log2limmablock_res$p.value[,1])$qvalue<0.05))
 ```
-Number of differentially expressed probes found with the alternative analysis is **`r sum(qvalue(datamatrix_limma_res$p.value[,1])$qvalue<0.05)`**.
+Number of differentially expressed probes found with the alternative analysis is **`r sum(qvalue(log2limmablock_res$p.value[,1])$qvalue<0.05)`**.
 
@@
 
-Since limma was not the tool used in Johnson et al.'s workflow, we will first need to create the corresponding analysis in Limma for the whole ComBat adjusted data set in order to make a fair comparison.
-```{r}
-thismat = EB3[negativeprobesfilter, ]
-thismat[thismat<1]=1
-thismat=normalizeBetweenArrays(log2(thismat))
-Type = as.factor(sampleannotation$Type)
-design = model.matrix(~0 + Type)
-fit = lmFit(thismat, design)
-cont.matrix = makeContrasts ( contrasts="TypeR-TypeC", levels=design)  
-fit2 = contrasts.fit(fit, cont.matrix)
-EB3_limma_res = eBayes(fit2)
-rm(thismat, Type, design, fit, cont.matrix, fit2)
-print(table(qvalue(EB3_limma_res$p.value[,1])$qvalue<0.05))
-
-```
-The number of genes below the significance threshold is fairly similar to the t-test.
-
 Now we can compare the P-values for the two alternative procedures.
 ```{r, dev='svg', fig.width=8, fig.height=8}
-hist(EB3_limma_res$p.value[,1], border="blue", main="P-values, ComBat vs Limma", breaks=100, xlab="p-value")
-hist(datamatrix_limma_res$p.value[,1], border="red", add=T, breaks=100)
+hist(log2ComBat_res$p.value[,1], border="blue", main="P-values, ComBat vs Limma", breaks=100, xlab="p-value")
+hist(log2limmablock_res$p.value[,1], border="red", add=T, breaks=100)
 legend("topright", legend=c("ComBat adjusted", "Limma adjusted"), text.col=c("blue", "red"))
@@
 Lets assume one aim of this study was to obtain a list of genes that differed between control and treatment with a significant cut-off of q<0.05.
-Then handling the batch effect with ComBat adjustment will yield a list of **`r sum(qvalue(EB3_limma_res$p.value[,1])$qvalue<0.05)`** whereas a list of **`r sum(qvalue(datamatrix_limma_res$p.value[,1])$qvalue<0.05)`** probes would have been closer to the truth.
+Then handling the batch effect with ComBat adjustment will yield a list of **`r sum(qvalue(log2ComBat_res$p.value[,1])$qvalue<0.05)`** whereas a list of **`r sum(qvalue(log2limmablock_res$p.value[,1])$qvalue<0.05)`** probes would have been closer to the truth.
 Also it is of interest to note that it is mostly the same probes that are found.
 ```{r, tidy=FALSE}
-table( qvalue(datamatrix_limma_res$p.value[,1])$qvalue<0.05 ,
-         qvalue(EB3_limma_res$p.value[,1])$qvalue<0.05,
+table( qvalue(log2limmablock_res$p.value[,1])$qvalue<0.05 ,
+         qvalue(log2ComBat_res$p.value[,1])$qvalue<0.05,
        dnn=c("limma", "ComBat"))
@@
 ```{r, tidy=FALSE}
-table( rank(datamatrix_limma_res$p.value[,1])  <=1000 & 
-         rank(EB3_limma_res$p.value[,1])<=1000   )
+table( rank(log2limmablock_res$p.value[,1])  <=1000 & 
+         rank(log2ComBat_res$p.value[,1])<=1000   )
 ```
@@
 set.seed(100)
-randommatrix = datamatrix
-randommatrix[,] =rnorm(length(datamatrix), mean=0, sd=1)
-EBrand = as.matrix( sva::ComBat(
-            dat=randommatrix, 
-            batch=sampleannotation[colnames(randommatrix),"Batch"], 
-            mod=model.matrix(~as.factor(sampleannotation[colnames(randommatrix),"Type"])), 
+randomdata = log2data
+randomdata[,] =rnorm(length(randomdata), mean=0, sd=1)
+randomdataComBat = as.matrix( sva::ComBat(
+            dat=randomdata, 
+            batch=sampleannotation[colnames(randomdata),"Batch"], 
+            mod=model.matrix(~as.factor(sampleannotation[colnames(randomdata),"Type"])), 
             numCovs=NULL, 
@@
 
-fit = lmFit(EBrand, design)
+fit = lmFit(randomdataComBat, design)
 fit2 = contrasts.fit(fit, cont.matrix)
-EBrand_limma_pvalues = eBayes(fit2)$p.value[,1]
+randomdataComBat_pvalues = eBayes(fit2)$p.value[,1]
 
-fit = lmFit(randommatrix, design)
+fit = lmFit(randomdata, design)
 fit2 = contrasts.fit(fit, cont.matrix)
-randommatrix_limma_pvalues = eBayes(fit2)$p.value[,1]
+randomdata_pvalues = eBayes(fit2)$p.value[,1]
 
@@
 design = model.matrix(~0+Type+Block)
-fit = lmFit(randommatrix, design)
+fit = lmFit(randomdata, design)
 cont.matrix = makeContrasts ( contrasts="TypeR-TypeC", levels=design) 
@@
 fit2 = eBayes(fit2)
-randommatrix_limma_batch_pvalues = fit2$p.value[,1]
+randomdatalimmablock_pvalues = fit2$p.value[,1]
 
@@
 ```{r, dev='svg', fig.width=8, fig.height=8}
-hist(EBrand_limma_pvalues, border="blue",  main="P-values, Random numbers", breaks=100, xlab="p-value")
-hist(randommatrix_limma_pvalues, border="black", add=T, breaks=100)
-hist(randommatrix_limma_batch_pvalues, border="red", add=T, breaks=100)
+hist(randomdataComBat_pvalues, border="blue",  main="P-values, Random numbers", breaks=100, xlab="p-value")
+hist(randomdata_pvalues, border="black", add=T, breaks=100)
+hist(randomdatalimmablock_pvalues, border="red", add=T, breaks=100)
 legend("topright", legend=c("ComBat adjusted", "Limma adjusted", "No adjustment"), text.col=c("blue", "black", "red"))
@@
 The p-values for ComBat adjusted random numbers are slightly enriched for low p-values. Indicating that some of the enrichment seen in the real data is also false.
-
+<br/>  
+<br/>  
 Another sanity check is to subset the real data into fictive batches.
 The idea behind this check stems from the observation that the EB2 data set contains the same number of samples with the same batch/covariate balance as the third batch alone.
+
 > The third batch was used for comparison against the EB2 analysis results because it was an identical experiment to EB2 other than the fact that it was conducted in a single batch.
@@
 ```{r, dev='svg', fig.width=8, fig.height=8}
-hist(EB2_pvals, border="blue",  main="P-values, real data", breaks=100, xlab="p-value")
-hist(Batch3_pvals, border="red", add=T, breaks=100)
-legend("topright", legend=c("ComBat adjusted batch 1 and 2", "Batch 3"), text.col=c("blue", "red"))
-```
-The 2-batch experiment is able to retrieve more genes as significant than running all in one batch for the same sample size.
-This is also observed by Johnson et al., but not commented.
-We claim that the apparent better result for the two-batch experiment is likely a consequence of the use of ComBat.
-A test of this is to split batch 3 into two imaginary batches with the same design as for batch 1 and 2, and then look at the p-value distribution compared with the real batch 3.
+# need to subset the log2data and do the ComBat adjustment and do the limma tests
 
-```{r, tidy=FALSE}
-Batch45 = Batch3
-Batch45_annot = sampleannotation[dimnames(Batch45)[[2]],]
-Batch45_annot$Batch="4"
-Batch45_annot$Batch[Batch45_annot$Type=="C"][1:4] = "5"
-Batch45_annot$Batch[Batch45_annot$Type=="R"][1:3] = "5"
-print(table(sampleannotation[sampleannotation$Batch %in% c("1", "2"), c("Batch", "Type")]))
-print(table(Batch45_annot[, c("Batch", "Type")]))
-```
-Then adjust the data with fictive batches and calculate p-values as done for the un-adjusted batch 3.
-```{r, results='hide', tidy=FALSE}
-EB45 = as.matrix( sva::ComBat(
-            dat=Batch45, 
-            batch=Batch45_annot[colnames(Batch45),"Batch"], 
-            mod=model.matrix( ~as.factor(Batch45_annot[colnames(Batch45),"Type"])  ), 
+# subset log2data
+mat = log2data[,sampleannotation$Batch %in% c("1","2")]
+log2EB2 = as.matrix( sva::ComBat(
+            dat=mat, 
+            batch=sampleannotation[colnames(mat),"Batch"], 
+            mod=model.matrix( ~as.factor(sampleannotation[colnames(mat),"Type"])  ), 
             numCovs=NULL, 
@@
             prior.plots=FALSE))
+rm(mat)
+log2Batch3 = log2data[,sampleannotation$Batch=="3"]
 
-EB45_pvals = apply(EB45, 1 , 
-                  FUN=function(x){t.test(
-                                    x[Batch45_annot[colnames(EB45), "Type"]=="C"],
-                                    x[Batch45_annot[colnames(EB45), "Type"]=="R"]
-                                    )$p.value})
+# run limma
+Type = as.factor(sampleannotation[colnames(log2EB2), "Type"])
+design = model.matrix(~0 + Type)
+fit = lmFit(log2EB2, design)
+cont.matrix = makeContrasts ( contrasts="TypeR-TypeC", levels=design)  
+fit2 = contrasts.fit(fit, cont.matrix)
+log2EB2ComBat_pvals = eBayes(fit2)$p.value
+rm(Type, design, fit, cont.matrix, fit2)
+
+Type = as.factor(sampleannotation[colnames(log2Batch3), "Type"])
+design = model.matrix(~0 + Type)
+fit = lmFit(log2Batch3, design)
+cont.matrix = makeContrasts ( contrasts="TypeR-TypeC", levels=design)  
+fit2 = contrasts.fit(fit, cont.matrix)
+log2Batch3ComBat_pvals = eBayes(fit2)$p.value
+rm(Type, design, fit, cont.matrix, fit2)
+
+# plot the two p-value distibutions
+hist(log2EB2ComBat_pvals, border="blue",  main="P-values, real data", breaks=100, xlab="p-value")
+hist(log2Batch3ComBat_pvals, border="red", add=T, breaks=100)
+legend("topright", legend=c("ComBat adjusted batch 1 and 2", "Batch 3"), text.col=c("blue", "red"))
 ```
-```{r, dev='svg', fig.width=8, fig.height=8, tidy=FALSE}
-hist(EB45_pvals, border="blue",  main="P-values, real data", breaks=100, xlab="p-value")
-hist(Batch3_pvals, border="red", add=T, breaks=100)
-legend("topright", legend=c("ComBat adjusted fictive-batches of batch 3", "Batch 3 not adjusted"), text.col=c("blue", "red"))
-```
-The fictive-batches version with ComBat seemingly outperforms the original.
-This plot looks very similar to the EB2 vs.
-Batch3 p-values, and for both, the p-values for the ComBat adjusted data are likely deflated due to ComBat use.
-
-The fictive batch asssignment above was just one of several possible. Repeating the selection 10 times shows that the lower p-values was not a coincidence.
+The 2-batch experiment is able to retrieve more genes as significant(`r sum(qvalue(log2EB2ComBat_pvals)$qvalue<0.05)` for q<0.05) than running all in one batch(`r sum(qvalue(log2Batch3ComBat_pvals)$qvalue<0.05)` for q<0.05) for the same sample size.
+This is also observed by Johnson et al., but not commented.
+We claim that the apparent better result for the two-batch experiment is a consequence of the use of ComBat.
+A sanity check supporting this claim is to split batch 3 into two imaginary batches with the same design as for batch 1 and 2, and then look at the p-value distribution compared with the real batch 3.
+In order to obtain some robustness this test in performed for 10 permutations of different ways of sub-setting batch 3 into 2 fictive batches.
 
 ```{r, tidy=FALSE, results='hide'}
+
+# print the batch/covariate balance for the real batch 1 and batch 2
+print(table(sampleannotation[sampleannotation$Batch %in% c("1", "2"), c("Batch", "Type")]))
+
 fictivebatches_pvalcounts=list()
@@
 {
-Batch45 = Batch3
-Batch45_annot = sampleannotation[dimnames(Batch45)[[2]],]
-Batch45_annot$Batch="4"
-Batch45_annot$Batch[Batch45_annot$Type=="C"][sample(1:sum(Batch45_annot$Type=="C"), 4)] = "5"
-Batch45_annot$Batch[Batch45_annot$Type=="R"][sample(1:sum(Batch45_annot$Type=="R"), 3)] = "5"
-print(table(sampleannotation[sampleannotation$Batch %in% c("1", "2"), c("Batch", "Type")]))
-print(table(Batch45_annot[, c("Batch", "Type")]))
+permdata = log2Batch3
+permdata_annot = sampleannotation[dimnames(permdata)[[2]],]
+permdata_annot$Batch="4"
+permdata_annot$Batch[permdata_annot$Type=="C"][sample(1:sum(permdata_annot$Type=="C"), 4)] = "5"
+permdata_annot$Batch[permdata_annot$Type=="R"][sample(1:sum(permdata_annot$Type=="R"), 3)] = "5"
 
-EB45 = as.matrix( sva::ComBat(
-            dat=Batch45, 
-            batch=Batch45_annot[colnames(Batch45),"Batch"], 
-            mod=model.matrix( ~as.factor(Batch45_annot[colnames(Batch45),"Type"])  ), 
+# This permutations balance
+print(table(permdata_annot[, c("Batch", "Type")]))
+
+permdataComBat = as.matrix( sva::ComBat(
+            dat=permdata, 
+            batch=permdata_annot[,"Batch"], 
+            mod=model.matrix( ~as.factor(permdata_annot[,"Type"])  ), 
             numCovs=NULL, 
@@
 
-pvals = apply(EB45, 1 , 
-                  FUN=function(x){t.test(
-                                    x[Batch45_annot[colnames(EB45), "Type"]=="C"],
-                                    x[Batch45_annot[colnames(EB45), "Type"]=="R"]
-                                    )$p.value})
+Type = as.factor(permdata_annot[, "Type"])
+design = model.matrix(~0 + Type)
+fit = lmFit(permdataComBat, design)
+cont.matrix = makeContrasts ( contrasts="TypeR-TypeC", levels=design)  
+fit2 = contrasts.fit(fit, cont.matrix)
+pvals = eBayes(fit2)$p.value
 
@@
 ```
-In order to save some computational time the parameteric version of ComBat was used for this.
+
+The p-values for the 10 splits of Batch 3 adjusted by ComBat is compared to the p-values for the not-split(and not ComBat adjusted) Batch 3.
 ```{r, dev='svg', fig.width=8, fig.height=8, tidy=FALSE}
-plot((1:20)/100, hist(Batch3_pvals, plot=FALSE, breaks=100)$counts[1:20], col="red",  main="P-values, fictive batches", 
+plot((1:20)/100, hist(log2Batch3ComBat_pvals, plot=FALSE, breaks=100)$counts[1:20],
+     col="red",  main="P-values, fictive batches", 
       xlab="p-value", ylab="Frequency", type="l", lwd=2, 
@@
 ```
+The two-batch approach seem to consistently outperform the all-in-one-batch version.
 
+Alternativ sanitycheck som data set 2 ikke feiler paa, saa jeg synes ikke den skal med. Laget for aa vise i figur som eksempel
 ```{r}
@@
 {
-  x = shufflesamplesinbatch( rownames(sampleannotation), sampleannotation$Batch, sampleannotation$Type, c("C", "R") )
-  print(  sum(sampleannotation[,"Type"] != sampleannotation[x,"Type"]  ))  # see how many changed
-  datamatrices_permuted[[i]] = as.matrix(ComBat(dat=datamatrix[, x], 
+  permlabels = shufflesamplesinbatch( rownames(sampleannotation), sampleannotation$Batch, sampleannotation$Type, c("C", "R") )
+  print(  sum(sampleannotation[,"Type"] != sampleannotation[permlabels,"Type"]  ))  # see how many changed
+  datamatrices_permuted[[i]] = as.matrix(ComBat(dat=datamatrix[, permlabels], 
                                                 batch=sampleannotation$Batch, 
@@
 {
-  this_p = apply(datamatrices_permuted[[i]], 1 , 
-                  FUN=function(x){t.test(
-                                    x[sampleannotation[, "Type"]=="C"],
-                                    x[sampleannotation[, "Type"]=="R"]
-                                    )$p.value})
+
+  Type = as.factor(sampleannotation$Type)
+  design = model.matrix(~0 + Type)
+  fit = lmFit(datamatrices_permuted[[i]], design)
+  cont.matrix = makeContrasts ( contrasts="TypeR-TypeC", levels=design)  
+  fit2 = contrasts.fit(fit, cont.matrix)
+  this_p = eBayes(fit2)$p.value[,1]
 
   permuted_pvalcounts[[i]] = hist(this_p, plot=FALSE, breaks=100)$counts
-  print(table(p.adjust(this_p, method="fdr")<0.05))
+  print(table(qvalue(this_p)$qvalue<0.05))
 }
@@
 
-# reproduced ComBat + limma
-hist(EB3_limma_res$p.value[,1], main="P-values, real data, ComBat followed by Limma", breaks=100, xlab="p-value", ylim=c(0,3000))
+# real data, reproduced ComBat + limma
+hist(log2ComBat_res$p.value[,1], main="P-values, real data, ComBat followed by Limma", breaks=100, xlab="p-value", ylim=c(0,3000))
 
-# batch handled in limma
-hist(datamatrix_limma_res$p.value[,1], breaks=100, main="P-values, real data, batch handled by Limma", xlab="p-value" , ylim=c(0,3000))
+# real data, batch handled in limma
+hist(log2limmablock_res$p.value[,1], breaks=100, main="P-values, real data, batch handled by Limma", xlab="p-value" , ylim=c(0,3000))
 
 # random ComBat + limma
-hist(EBrand_limma_pvalues, main="P-values, Random data, ComBat followed by Limma", breaks=100, xlab="p-value" , ylim=c(0,3000))
+hist(randomdataComBat_pvalues, main="P-values, Random data, ComBat followed by Limma", breaks=100, xlab="p-value" , ylim=c(0,3000))
 
@@
 ### permutation sanity test
-real_pvalcounts=hist(EB3_limma_res$p.value[,1], plot=FALSE, breaks=100)$counts
+real_pvalcounts=hist(log2ComBat_res$p.value[,1], plot=FALSE, breaks=100)$counts
 plot((1:20)/100, real_pvalcounts[1:20], col="red",
@@
 # reproduced ComBat + limma
-hist(EB3_limma_res$p.value[,1], main="P-values, 3 settings", breaks=100, xlab="p-value", ylim=c(0,3000), border="red")
+hist(log2ComBat_res$p.value[,1], main="P-values, 3 settings", breaks=100, xlab="p-value", ylim=c(0,3000), border="red")
 
 # batch handled in limma
-hist(datamatrix_limma_res$p.value[,1], add=T, breaks=100, xlab="p-value" , ylim=c(0,3000), border="blue")
+hist(log2limmablock_res$p.value[,1], add=T, breaks=100, xlab="p-value" , ylim=c(0,3000), border="blue")
 
 # random ComBat + limma
-hist(EBrand_limma_pvalues, add=T, breaks=100, xlab="p-value" , ylim=c(0,3000), border="black")
+hist(randomdataComBat_pvalues, add=T, breaks=100, xlab="p-value" , ylim=c(0,3000), border="black")
 

### 2014-04-02 reanalysis/Fadi_Towfic_generic_medicine/towfic_et_al.rmd
figure script added

@@
 meanN = rowMeans(thisdata[, sampleannotation[, "covariate"] =="N"]) 
-
+rm(thisdata)
 foldchange_RS_vs_M = meanRS - meanM
@@
 design = model.matrix(~0+group+block)
-fit = lmFit(thisdata, design)
+fit = lmFit(datamatrices[["real_qnorm"]], design)
 cont.matrix = makeContrasts ( contrasts="groupDP-groupN", levels=design)  

### 2014-04-02 commonscripts/helper_functions.r

@@ -290,3 +290,77 @@ shufflesamplesinbatch = function(samplenames, batch, covariate, shufflecovariate
 
 
 
+
+# Towfic et al.
+# load data and probe annotation and some formatting
+loadtowfic = function(downloaddata=TRUE)
+{
+    
+  if(downloaddata)
+  {
+    temp = tempfile()
+    download.file(url="http://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE40566&format=file&file=GSE40566%5Fnon%5Fnormalized%2Etxt%2Egz",
+                  destfile=temp, mode = "wb")
+    rawdata = read.table(temp, sep="\t", header=TRUE, stringsAsFactors=FALSE)
+    unlink(temp)
+    
+    temp = tempfile()
+    download.file(url="http://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE40566&format=file",
+                  destfile=temp, mode = "wb")
+    tardirtemp = tempdir()
+    untar(temp, exdir = tardirtemp)
+    rawannotation = read.table(paste(tardirtemp ,"/GPL6887_MouseWG-6_V2_0_R3_11278593_A.txt.gz", sep=""), 
+                               sep="\t", header=TRUE, stringsAsFactors=FALSE, 
+                               skip=8, comment.char="", quote="", fill=TRUE)
+    
+    unlink(temp)
+    unlink(tardirtemp, recursive=TRUE )
+  }else{
+    
+    # if download did not work change downloaddata to FALSE  
+    # download data from the GEO deposit
+    # unpack and place files in a folder named "not_in_github"
+    rawdata = read.table("not_in_github/GSE40566_non_normalized.txt", 
+                         sep="\t", header=TRUE, stringsAsFactors=FALSE)
+    
+    # the probe annotation file found inside GSE40566_RAW.tar
+    rawannotation = read.table("not_in_github/GPL6887_MouseWG-6_V2_0_R3_11278593_A.txt", 
+                               sep="\t", header=TRUE, stringsAsFactors=FALSE, 
+                               skip=8, comment.char="", quote="", fill=TRUE)
+  }  
+  
+  tmp =  rawannotation$Species=="Mus musculus"
+  experimentalannot = rawannotation[tmp,]
+  experimentalannot$Array_Address_Id = as.numeric(experimentalannot$Array_Address_Id)
[…]

### 2014-04-02 reanalysis/ComBat_original/figurescriptdataset2.r

@@ -1,0 +1,114 @@
+
+
+starttime = Sys.time()
+
+includelibs = c("sva", "qvalue", "limma")
+tmp = lapply(includelibs, require, character.only=T)
+print(tmp)
+if(any(!unlist(tmp)))
+{
+  stop( paste("Not able to find all packages. Please install ",
+              paste(includelibs[!unlist(tmp)], collapse=", ") )
+  )
+}
+rm(tmp)
+debug=FALSE
+useparprior=TRUE
+
+
+datamatrix = as.matrix(read.table("data/dataExample2.txt", sep="\t", header=TRUE))
+if(debug)datamatrix = datamatrix[1:3000,]
+sampleannotation = read.table("data/sampleInfoExample2.txt", 
+                              sep="\t", header=TRUE, stringsAsFactors=FALSE)
+rownames(sampleannotation)=sampleannotation$ArrayName
+sampleannotation$Batch=factor(as.character(sampleannotation$Batch)) 
+datamatrix=datamatrix[,sampleannotation$Type!="WT"]
+sampleannotation=sampleannotation[sampleannotation$Type!="WT",]
+sampleannotation$Type=factor(sampleannotation$Type)
+
+log2data = datamatrix
+# flooring to 1
+log2data[log2data<1]=1
+# take out data with to much low/missing values.
+negativeprobesfilter =( rowSums(log2data>1) >= (0.9*ncol(log2data)) )
+log2data = log2data[negativeprobesfilter,]
+# quantilenormalize
+log2data=normalizeBetweenArrays(log2(log2data))
+
+log2dataComBat = as.matrix( sva::ComBat(
+  dat=log2data, 
+  batch=sampleannotation[colnames(log2data),"Batch"], 
+  mod=model.matrix(~as.factor(sampleannotation[colnames(log2data),"Type"])), 
+  numCovs=NULL, 
+  par.prior=useparprior, 
+  prior.plots=FALSE))
+
+# DE pvalues ComBat adjusted real data
+Type = as.factor(sampleannotation$Type)
+design = model.matrix(~0 + Type)
+fit = lmFit(log2dataComBat, design)
+cont.matrix = makeContrasts ( contrasts="TypeR-TypeC", levels=design)  
+fit2 = contrasts.fit(fit, cont.matrix)
+log2ComBat_res = eBayes(fit2)
+rm(Type, design, fit, cont.matrix, fit2)
[…]

### 2014-04-02 reanalysis/Fadi_Towfic_generic_medicine/figuretowfic.r

@@ -1,0 +1,131 @@
+
+# source("figuretowfic.r")
+starttime = Sys.time()
+debug = FALSE
+downloaddata=FALSE
+set.seed(100)
+
+
+includelibs = c("Biobase", "GEOquery", "sva", "limma",
+                "preprocessCore")
+lapply(includelibs, require, character.only=T)
+
+
+sampleannotation = read.table("data/sampleannotation.csv", sep="\t", header=TRUE,  stringsAsFactors=FALSE)
+sampleannotation$code = make.names(sampleannotation$code)
+sampleannotation$chip = as.character(sampleannotation$chip)
+dimnames(sampleannotation)[[1]] = sampleannotation$code
+# take out 3 samples that are not assign to a geoaccession. Failed QC?
+sampleannotation = sampleannotation[!is.na(sampleannotation$geoaccession),] 
+
+
+source("../../commonscripts/helper_functions.r")
+
+
+ret = loadtowfic(downloaddata)
+datamatrices = list()
+datamatrices[["real_raw"]] = ret[["data"]]
+probeannotation = ret[["probes"]]
+
+if(debug)
+  datamatrices[["real_raw"]] = datamatrices[["real_raw"]][1:1000,]
+
+
+datamatrices[["real_qnorm"]] = normalize.quantiles(datamatrices[["real_raw"]]) 
+dimnames(datamatrices[["real_qnorm"]]) = dimnames(datamatrices[["real_raw"]])
+
+
+
+# combat adjusted
+combatmod = model.matrix(~as.factor(sampleannotation$covariate))
+datamatrices[["real_combat_covariates"]]= as.matrix(ComBat(dat=datamatrices[["real_qnorm"]],
+                                                           batch=sampleannotation$chip,
+                                                           mod=combatmod,
+                                                           numCovs=NULL,
+                                                           par.prior=TRUE,
+                                                           prior.plots=FALSE))
+
+group = factor(sampleannotation$covariate)
+design = model.matrix(~0 + group)
+fit = lmFit(datamatrices[["real_combat_covariates"]], design)
+cont.matrix = makeContrasts ( contrasts="groupDP-groupN", levels=design)  
+fit2 = contrasts.fit(fit, cont.matrix)
+limma_ret_alt = eBayes(fit2)
+limma_p_alt = limma_ret_alt$p.value
[…]

### 2014-04-03 tex/manuscript.tex
TeX

--- 2014-03-17 docs/article.rmd
@@
-Methods that remove batch effects while retaining group differences may lead to exaggerated confidence in downstream analyses
-========================================================
+\documentclass{bio}
+%\documentclass[oupdraft]{bio}
+\usepackage[colorlinks=true, urlcolor=citecolor, linkcolor=citecolor, citecolor=citecolor]{hyperref}
 
-```{r echo=FALSE, results='hide'}
-"
-Alternativer til tittel
-Methods that remove batch effects while retaining group differences may lead to incorrect conclusions in downstream analyses
-Methods that remove batch effects while retaining group differences may lead to incorrect conclusions
-"
-```
+% Add history information for the article if required
+\history{}
 
-## Abstract
-Removal of, or adjustment for, batch effects or centre differences is generally required when such effects are present in data.
+\begin{document}
+
+% Title of paper
+\title{Methods that remove batch effects while retaining group differences may lead
+to exaggerated confidence in downstream analyses
+}
+
+% List of authors, with corresponding author marked by asterisk
+\author{VEGARD NYGAARD, EINAR ANDREAS R\O{}DLAND, EIVIND HOVIG$^\ast$\\[4pt]
+% Author addresses
+\textit{Department of Tumor Biology,
+Institute for Cancer Research,
+Oslo University Hospital HF - Radiumhospitalet,
+Montebello,
+0310 Oslo,
+Norway}
+\\[2pt]
+% E-mail address for correspondence
+{ehovig@radium.uio.no}}
+
+% Running headers of paper:
+\markboth%
+% First field is the short list of authors
+{V. Nygaard, E. A. R\o{}dland, E. Hovig}
+% Second field is the short title of the paper
+{Exagerated effects after batch correction}
+
+\maketitle
+
+% Add a footnote for the corresponding author if one has been
+% identified in the author list
+\footnotetext{To whom correspondence should be addressed.}
+
+\begin{abstract}
+{Removal of, or adjustment for, batch effects or centre differences is generally required when such effects are present in data.
 In particular, when preparing microarray gene expression data from multiple cohorts, array platforms, or batches for later analyses, batch effects can have confounding effects.
@@
 The scientific community seems to be largely unaware of this problem, which most likely has contributed to false discoveries being presented in the published literature.
+}
+\end{abstract}
 
-## Introduction
+%% NOTES TO SELF:
+% Cite using:
+%    \citet{...}  in text
+%    \citealp{...}   within parantheses
+%
+% Temporary notifications and reminders:
+\newcommand\NB[1]{{\color{red}#1}}
+\newcommand\NOTE[1]{{\color{red}[#1]}}
+\newcommand\CITE[1]{{\color{magenta}[#1]}}
+
+\section{Introduction}\label{intro}
 
@@
 However, when batch is a confounder, some of the variance removed will be group variance and statistical power will be lost.
-In addition if there are many groups, some effect will be transferred between groups that share batches, leading to false positive results (_Dette problemet er ikke beskrevet i de andre artiklene eller i einars mail, men saa vidt jeg skjoenner saa kan dette skje_).
+In addition if there are many groups, some effect will be transferred between groups that share batches, leading to false positive results.
+\NOTE{Dette problemet er ikke beskrevet i de andre artiklene eller i einars mail, men saa vidt jeg skjoenner saa kan dette skje.}
 
@@
 
-```{r echo=FALSE, results='hide'}
-"
-Kort oppsummere behovet for batch-korrigering: generelt og i microarraydata.
+\NB{Kort oppsummere behovet for batch-korrigering: generelt og i microarraydata.}
 
-Kort beskrive metoder som gjoer dette (batch-sentrering, ANOVA, ComBat).
+\NB{Kort beskrive metoder som gjoer dette (batch-sentrering, ANOVA, ComBat).}
 
-Forklare tilsiktet effekt disse paa batch+gruppe-forskjeller (balansert versus ikke-balansert).
+\NB{Forklare tilsiktet effekt disse paa batch+gruppe-forskjeller (balansert versus ikke-balansert).}
 
-Forklare kort hva faktisk effekt er: ANOVA gir redusert gruppe-forskjeller, mens for ComBat overestimeres den statistiske styrken/sikkerheten av forskjellen.
-"
-```
+\NB{Forklare kort hva faktisk effekt er: ANOVA gir redusert gruppe-forskjeller, mens for ComBat overestimeres den statistiske styrken/sikkerheten av forskjellen.}
 
-## A simple sanity check
+\section{Methods for batch effect correction}
+
+\subsection{Model for data with batch effects}
+
+We will base out discussion of data with batch effects on a simple model:
+\begin{equation}
+Y_{ijr}=\alpha+\beta_j+\gamma_i+\epsilon_{ijr}
+\end{equation}
+where $i=1,\ldots,m$ are the different batches, $j=1,\ldots,M$ are different study groups that we wish to compare, and $r=1,\ldots,n_{ij}$ are the different samples within batch $i$ and group $j$.
+
+When combining data from more diverse data sources, e.g. microarray data from different platforms, a more general model is required. One such model, used by \CITE{ComBat}, is
+\begin{equation}
+Y_{ijgr}=\alpha_g+X_r\beta_{g}+\gamma_{ig}+\delta_{ig}\epsilon_{ijgr}
+\end{equation}
+where $g=1,\ldots,G$ are different measurements, e.g.
+genes, performed for each sample, and $X$ is the design matrix which in our case will indicate the study group.
+This permits independent rescaling of data from different batches.
+In addition, \CITE{ComBat} uses an empirical Bayes approach to estimate $\gamma_{ig}$ and $\delta_{ig}$ to stabilise estimates, which is critical for use with small batches.
+
+For simplicity, we consider the case with a single gene and constant scale, i.e.
+$\delta_{ig}=1$.
+We will discuss the effect of empirical Bayes estimation of $\gamma_{ig}$ later, but out main argument is more easily made in the simpler context.
+
+\subsection{Standard batch correction methods}
+
+The main ambition of batch effect adjustments is to be able to remove batch differences in such a way that downstream analyses of the adjusted data may be done without further batch adjustments.
+
+The most common method for removing batch effects is to zero-centre each batch:
+\begin{equation}
+\Delta\widetilde Y_{ijr}=Y_{ijr}-\overline{Y}_{i}
+\quad\text{where}\quad
+\bar Y_i=\frac{\sum_{j=1}^M\sum_{r=1}^{n_{ij}} Y_{ijr}}{\sum_{j=1}^M n_{ij}}.
+\end{equation}
+An alternative is to centre each batch to the common average by adding the average value $\bar Y$ across the entire data set: i.e.
+$\widetilde Y_{ijr}=\Delta\widetilde Y_{ijr}+\bar Y$.
+When comparing groups, the common value $\bar Y$ has no effect, and so this is equilvalent to zero-centring each batch.
+However, if the different groups are unevenly represented in the different batches, the batch average $\overline{Y}_i$ will tend to capture group differences as well as batch effects.
+Thus, batch centering may reduce group differences, and thus reduce the power of downstream analyses.
+
+Removing batch effects while retaining group differences can be done through an ANOVA analysis in which the group effects, $\beta_j$, and the batch effects, $\gamma_i$, are estimated simultaneously.
+Batch adjusted values may then be obtained by subtracting the estimated batch effects, $\hat\beta_j$:
+\begin{equation}
+\widetilde Y_{ijr}=Y_{ijr}-\hat\beta_j=\alpha+(\beta_j-\hat\beta_j)+\gamma_i+\epsilon_{ijr}.
+\end{equation}
+This will yield batch adjusted values where any systematic bias induced by the batch differences has been removed, while the group differences are retained.
+
+Unfortunately, the estimation error $\hat\beta_j-\beta_j$ affects all values within the same batch in the same manner, inducing a dependency between the values.
+If the study groups are evenly represented in all batches, this will not influence estimated group differences as all groups are equally affected.
+However, if groups are unevenly represented, this induced dependency can have severe impact on downstream analyses.
+
+\section{Results}
+
+\subsection{A simple sanity check}
+
 The undesired consequences of preserving group effects when correcting for batch effect is readily illustrated with a sanity check using random numbers .
@@
 The skewed distribution is a indication that this approach may have a unintentional adverse effect.
-_Jeg har ikke beskrevet permutatsjon checken fordi jeg er usikker paa om vi skal bruke den._
+\NOTE{Jeg har ikke beskrevet permutatsjon checken fordi jeg er usikker paa om vi skal bruke den.}
 
-![fig. 1](../reanalysis/leek_sva/figure/pvaluesleeksva.svg)  
-**Figure 1**  
-Two sanity checks were the recommended use of ComBat fails. Adapted from the user guide in the sva package. 
-- a) Real data is substituted with random numbers from a normal distribution (mean=0, sd=1), but the batch/group design is retained, followed by batch adjustment in ComBat and a F-test.
-- b) 10 runs of real data with the "cancer" labels permuted within batches, followed by ComBat adjustment and a F-test. All permutations produces a skewed p-value distribution.
+%% FIGURE 1: Original position
 
-## Formler og ord
-_Forklare hva som gaar galt med formler og ord. Overlater til Einar aa skrive dette_
+\NB{Formler og ord:
+Forklare hva som gaar galt med formler og ord. Overlater til Einar aa skrive dette.}
 
-## Examples of undesired consequences
+\subsection{Examples of undesired consequences}
+
 As the amount of false positive results when trying to retain group differences depends on the batch/group balance, we will show two examples with varying degree of unbalancedness.
@@
 1) In the first experiment(24421904), cells were treated with glatiramer acetate (a medicine for multiple sclerosis) or a generic and mRNA was measured using microarrays alongside control samples.
-A batch effect correlating to the chip (Illumina WG-6_V2, six samples per chip) was observed and adjusted for with ComBat, whereafter the data was tested for differetially expressed genes, yielding a list of 1000 genes (Table S5 24421904).
+A batch effect correlating to the chip (Illumina WG-6\_V2, six samples per chip) was observed and adjusted for with ComBat, whereafter the data was tested for differetially expressed genes, yielding a list of 1000 genes (Table S5 24421904).
 Unfortunately the batch/treatment design was unbalanced with several batches having only one of the main treatments of interest.
@@
 This example is a sort of "worst case" scenario for applying ComBat, since it both has a very unbalanced batch/group design and a a priori assumpotion of no difference.
-The R-code for our analysis and a more extensive report can be downloaded from github (https://github.com/vegardny/combat_tests.git).
+The R-code for our analysis and a more extensive report can be downloaded from github (\href{https://github.com/vegardny/combat_tests.git}).
 
-![fig. 2](../reanalysis/Fadi_Towfic_generic_medicine/figure/pvaluestowfic.svg)
-**Figure 2**.
-P-value distribution for the main comparison in (data set 2 24421904) glatiramer acetate vs. generic.  
-- a) Real data adjusted by ComBat as described in the article followed by a significance test in limma. 
-- b) Real data not adjusted by ComBat, tested for significance in limma using batch as a blocking factor. 
-- c) Real data is substituted with random numbers from a normal distribution (mean=0, sd=1), but the batch/group design is retained, followed by batch adjustment in ComBat and a significance test in limma.
-- d) 10 runs of real data with the glatiramer acetate("DP") and generic("N") labels permuted within batches, followed by ComBat adjustment and significance test in limma.  
-_jeg tok med begge sanitycheckene her. Er d) slik einar foreslo?. Merk jeg har ikke permutert alle labeler, bare de mellom DP og N. Det er en del andre._
+%![fig. 2](../reanalysis/Fadi_Towfic_generic_medicine/figure/pvaluestowfic.svg)
+%**Figure 2**.
+%P-value distribution for the main comparison in (data set 2 24421904) glatiramer acetate vs. generic.  
+%- a) Real data adjusted by ComBat as described in the article followed by a significance test in limma. 
+%- b) Real data not adjusted by ComBat, tested for significance in limma using batch as a blocking factor. 
+%- c) Real data is substituted with random numbers from a normal distribution (mean=0, sd=1), but the batch/group design is retained, followed by batch adjustment in ComBat and a significance test in limma.
+%- d) 10 runs of real data with the glatiramer acetate("DP") and generic("N") labels permuted within batches, followed by ComBat adjustment and significance test in limma.  
+%_jeg tok med begge sanitycheckene her. Er d) slik einar foreslo?. Merk jeg har ikke permutert alle labeler, bare de mellom DP og N. Det er en del andre._
 
@@
 Nevertheless, we belive that  P-values are deflated for the ComBat adjusted analysis.
-The R-code for our analysis and a more extensive report can be downloaded from github (https://github.com/vegardny/combat_tests.git).
+The R-code for our analysis and a more extensive report can be downloaded from github (\href{https://github.com/vegardny/combat_tests.git}).
 
-![fig. 3](../reanalysis/ComBat_original/figure/pvaluesjohnson.svg)  
-**Figure 3**.
-P-value distribution for the main comparison in (data set 2 16632515) for  
-- a) Real data adjusted by ComBat as described in the article followed by a significance test in limma. 
-- b) Real data not adjusted by ComBat, tested for significance in limma using batch as a blocking factor. 
-- c) Real data is substituted with random numbers from a normal distribution (mean=0, sd=1), but the batch/group design is retained, followed by batch adjustment in ComBat and a significance test in limma.
-- d) 10 runs of real data with the group labels permuted within batches, followed by ComBat adjustment and significance test in limma.   
-_problemet med permutatsjons sanitychechen slik jeg har utfoert og plottet den her er at den neppe kan sies aa feile.
+%![fig. 3](../reanalysis/ComBat_original/figure/pvaluesjohnson.svg)  
+%**Figure 3**.
+%P-value distribution for the main comparison in (data set 2 16632515) for  
+%- a) Real data adjusted by ComBat as described in the article followed by a significance test in limma. 
+%- b) Real data not adjusted by ComBat, tested for significance in limma using batch as a blocking factor. 
+%- c) Real data is substituted with random numbers from a normal distribution (mean=0, sd=1), but the batch/group design is retained, followed by batch adjustment in ComBat and a significance test in limma.
+%- d) 10 runs of real data with the group labels permuted within batches, followed by ComBat adjustment and significance test in limma.   
+%_problemet med permutatsjons sanitychechen slik jeg har utfoert og plottet den her er at den neppe kan sies aa feile.
 Dessuten kan de 3 foerst plottene bli plottet i ett som vist under i en alternativ version jeg synes er finere, men dette gaar ikke for permutasjonstesten tror jeg_
 
-![fig. 4](../reanalysis/ComBat_original/figure/pvaluesjohnson2.svg)   
-_Alternativ illustrasjon av de tre p-value distribusjonene_
+%![fig. 4](../reanalysis/ComBat_original/figure/pvaluesjohnson2.svg)   
+%_Alternativ illustrasjon av de tre p-value distribusjonene_
 
-3) _Her kan vi fylle paa med flere eksempler etter samme mal hvis det trengs.
+3) \NOTE{Her kan vi fylle paa med flere eksempler etter samme mal hvis det trengs.
 Feks en fersk artikkel fra Nature Genetic([24584070](http://www.nature.com/ng/journal/vaop/ncurrent/full/ng.2916.html)).
@@
 Dog gir det et falskt inntrykk av at dataene deres ligner tidligere data, se heatmap i suppl..
-Dessuten lager de GO analyser av dette som egentlig er 7 aar gamle andres data og ikke deres_
+Dessuten lager de GO analyser av dette som egentlig er 7 aar gamle andres data og ikke deres}
 
-## Discussion
+\section{Discussion}
 
-## Increased emphasis on preserving group difference
+\subsection{Increased emphasis on preserving group difference}
+
 In the original ComBat article(2007) it is clear that the primary motivation behind ComBat was to employ an Emprical Bayes method for batch-effect removal.
@@
 (
-[1](https://groups.google.com/d/msg/combat-user-forum/eSVSKwGtuyE/ZIWV2juYmmAJ)
-,[2](https://groups.google.com/d/msg/combat-user-forum/Vkb9p7wekd4/h5Etie7FTVQJ)
-,[3](https://groups.google.com/d/msg/combat-user-forum/fpBTcgDjiR8/uo4QIZL4sZgJ)
-,[4](https://groups.google.com/d/msg/combat-user-forum/26FZlgU2LFQ/W6U_Lhh_64EJ)
-). _Aa linke direkte er kanskje litt ufint, alternativt kan det linkes til forumet uten konkrete innlegg_.  
-And when ComBat was incorporated in the sva package ( [22257669](http://bioinformatics.oxfordjournals.org/cgi/pmidlookup?view=long&pmid=22257669), 2012), inclusion of group labels was made almost mandatory (an undocumented option of passing a NULL value exists).
+[1](\href{https://groups.google.com/d/msg/combat-user-forum/eSVSKwGtuyE/ZIWV2juYmmAJ})
+,[2](\href{https://groups.google.com/d/msg/combat-user-forum/Vkb9p7wekd4/h5Etie7FTVQJ})
+,[3](\href{https://groups.google.com/d/msg/combat-user-forum/fpBTcgDjiR8/uo4QIZL4sZgJ})
+,[4](\href{https://groups.google.com/d/msg/combat-user-forum/26FZlgU2LFQ/W6U_Lhh_64EJ})
+). \NOTE{Aa linke direkte er kanskje litt ufint, alternativt kan det linkes til forumet uten konkrete innlegg.}  
+And when ComBat was incorporated in the sva package ( [22257669](\href{http://bioinformatics.oxfordjournals.org/cgi/pmidlookup?view=long&pmid=22257669}), 2012), inclusion of group labels was made almost mandatory (an undocumented option of passing a NULL value exists).
 Thus, this problematic use has likely been common.
 
-## Motivation for this warning
+\subsection{Motivation for this warning}
+
 Our knowledge of ComBat came through a typical use case when trying to salvage unbalanced data which had batch effects.
@@
 Often no effort was done in order to substantiate the existence of batch effects in their data, except for stating the presence of batches (24391845?, 18414638, 21731603, 23630272).
-In one instance ComBat was even applied on data were effects due to batch were investigated, but not found (_husker ikke ref_).
+In one instance ComBat was even applied on data were effects due to batch were investigated, but not found \NOTE{husker ikke ref}.
 The incorporation of the method into analysis pipelines (16642009,TCGA) and other packages (23452776, 21937664) could make its usage even more trivial and parameters setting harder to perceive.
@@
 
-## Practical advice
+\subsection{Practical advice}
+
 We have shown that adjusting for batch effects while preserve the group difference may lead to varying degree of false results.
@@
 
-## Reproducible research
-_vi boer klare aa tilfredstille kravene i http://biostatistics.oxfordjournals.org/content/10/3/405.full_
+\subsection{Reproducible research}
+\NOTE{vi boer klare aa tilfredstille kravene i http://biostatistics.oxfordjournals.org/content/10/3/405.full}
 
-## References
+\section{Supplementary Material}
+\label{supp}
 
-http://biostatistics.oxfordjournals.org/content/8/1/118.abstract
-Adjusting batch effects in microarray expression data using empirical Bayes methods.
-Johnson WE, Li C, Rabinovic A.
-Biostatistics. 2007 Jan;8(1):118-27. Epub 2006 Apr 21.
-PMID: 16632515 [PubMed - indexed for MEDLINE] Free Article
+Supplementary material is available online at
+\href{http://biostatistics.oxfordjournals.org}%
+{http://biostatistics.oxfordjournals.org}.
 
-Molecular profiling of single Sca-1+/CD34+,- cells--the putative murine lung stem cells.
-Hittinger M, Czyz ZT, Huesemann Y, Maneck M, Botteron C, Kaeufl S, Klein CA, Polzer B.
-PLoS One. 2013 Dec 31;8(12):e83917. doi: 10.1371/journal.pone.0083917. eCollection 2013.
-PMID: 24391845 [PubMed - in process] 
-http://www.plosone.org/article/info%3Adoi%2F10.1371%2Fjournal.pone.0083917
+\section*{Acknowledgments}
 
-Drinking-water arsenic exposure modulates gene expression in human lymphocytes from a U.S. population.
-Andrew AS, Jewell DA, Mason RA, Whitfield ML, Moore JH, Karagas MR.
-Environ Health Perspect. 2008 Apr;116(4):524-31. doi: 10.1289/ehp.10861.
-PMID: 18414638 [PubMed - indexed for MEDLINE] Free PMC Article
-http://www.ncbi.nlm.nih.gov/pmc/articles/PMC2290973/
+[...Acknowledgements...]
+{\it Conflict of Interest}: None declared.
 
-Epigenetic predictor of age.
-Bocklandt S, Lin W, Sehl ME, Sánchez FJ, Sinsheimer JS, Horvath S, Vilain E.
-PLoS One. 2011;6(6):e14821. doi: 10.1371/journal.pone.0014821. Epub 2011 Jun 22.
-PMID: 21731603 [PubMed - indexed for MEDLINE] Free PMC Article
-http://www.plosone.org/article/info%3Adoi%2F10.1371%2Fjournal.pone.0014821
+\bibliographystyle{biorefs}
+\bibliography{refs}
 
-Childhood maltreatment is associated with distinct genomic and epigenetic profiles in posttraumatic stress disorder.
-Mehta D, Klengel T, Conneely KN, Smith AK, Altmann A, Pace TW, Rex-Haffner M, Loeschner A, Gonik M, Mercer KB, Bradley B, Müller-Myhsok B, Ressler KJ, Binder EB.
-Proc Natl Acad Sci U S A. 2013 May 14;110(20):8302-7. doi: 10.1073/pnas.1217750110. Epub 2013 Apr 29.
-PMID: 23630272 [PubMed - indexed for MEDLINE] Free PMC Article
-http://www.pnas.org/content/110/20/8302.full
+%\begin{figure}[!p]
+%\centering\includegraphics{fig1.eps}
+%\caption{...}
+%\label{Fig1}
+%\end{figure}
 
-Patterns of regulatory activity across diverse human cell types predict tissue identity, transcription factor binding, and long-range interactions.
-Sheffield NC, Thurman RE, Song L, Safi A, Stamatoyannopoulos JA, Lenhard B, Crawford GE, Furey TS.
-Genome Res. 2013 May;23(5):777-88. doi: 10.1101/gr.152140.112. Epub 2013 Mar 12.
-PMID: 23482648 [PubMed - indexed for MEDLINE] Free PMC Article
-http://genome.cshlp.org/content/23/5/777.full
+\begin{figure}[!p]
+\centering\includegraphics[width=15cm]{Fig/pvaluesleeksva.png}
+\caption{
+Two sanity checks were the recommended use of ComBat fails. Adapted from the user guide in the sva package. 
+a) Real data is substituted with random numbers from a normal distribution (mean=0, sd=1), but the batch/group design is retained, followed by batch adjustment in ComBat and a F-test.
+b) 10 runs of real data with the "cancer" labels permuted within batches, followed by ComBat adjustment and a F-test. All permutations produces a skewed p-value distribution.
+}
+\label{Fig1}
+\end{figure}
 
-A recurrent inactivating mutation in RHOA GTPase in angioimmunoblastic T cell lymphoma.
-Yoo HY, Sung MK, Lee SH, Kim S, Lee H, Park S, Kim SC, Lee B, Rho K, Lee JE, Cho KH, Kim W, Ju H, Kim J, Kim SJ, Kim WS, Lee S, Ko YH.
-Nat Genet. 2014 Mar 2. doi: 10.1038/ng.2916. [Epub ahead of print]
-PMID: 24584070 [PubMed - as supplied by publisher]
-http://www.nature.com/ng/journal/vaop/ncurrent/full/ng.2916.html
+%![fig. 1](../reanalysis/leek_sva/figure/pvaluesleeksva.svg)  
+%**Figure 1**
 
-A recurrent inactivating mutation in RHOA GTPase in angioimmunoblastic T cell lymphoma.
-Yoo HY, Sung MK, Lee SH, Kim S, Lee H, Park S, Kim SC, Lee B, Rho K, Lee JE, Cho KH, Kim W, Ju H, Kim J, Kim SJ, Kim WS, Lee S, Ko YH.
-Nat Genet. 2014 Mar 2. doi: 10.1038/ng.2916. [Epub ahead of print]
-PMID: 24584070 [PubMed - as supplied by publisher]
+%\begin{table}[!p]
+%\tblcaption{... \protect\citet{...} ... \label{Table1}}
+%{\tabcolsep=4.25pt
+%\begin{tabular}{@{}cccccccccc@{}}
+%\tblhead{Dis & Chr & Marker & MLE-OR & MLE & MLE & EB-HM-OR & EB-HM & EB-HM & ES \\
+%&&& (SE) & \textit{P}-value & rank & (SE) & \textit{P}-value & rank & rank}
+%CAD & 2 & rs2943634 & 1.22(0.04) & $1.23 \times 10^{-5}$ & 32 & 1.02(0.01) & 0.19 & 28 & 37 \lastline
+%\end{tabular}}
+%\end{table}
 
-Batch correction of microarray data substantially improves the identification of genes differentially expressed in rheumatoid arthritis and osteoarthritis.
-Kupfer P, Guthke R, Pohlers D, Huber R, Koczan D, Kinne RW.
-BMC Med Genomics. 2012 Jun 8;5:23. doi: 10.1186/1755-8794-5-23.
-PMID: 22682473 [PubMed - indexed for MEDLINE] Free PMC Article
-Related citations
-http://www.biomedcentral.com/1755-8794/5/23
-
-Relative impact of key sources of systematic noise in Affymetrix and Illumina gene-expression microarray experiments.
-Kitchen RR, Sabine VS, Simen AA, Dixon JM, Bartlett JM, Sims AH.
-BMC Genomics. 2011 Dec 1;12:589. doi: 10.1186/1471-2164-12-589.
-PMID: 22133085 [PubMed - indexed for MEDLINE] Free PMC Article
-http://www.biomedcentral.com/1471-2164/12/589/
-
-Tackling the widespread and critical impact of batch effects in high-throughput data.
-Leek JT, Scharpf RB, Bravo HC, Simcha D, Langmead B, Johnson WE, Geman D, Baggerly K, Irizarry RA.
-Nat Rev Genet. 2010 Oct;11(10):733-9. doi: 10.1038/nrg2825. Epub 2010 Sep 14.
-PMID: 20838408 [PubMed - indexed for MEDLINE] Free PMC Article
-http://www.nature.com/nrg/journal/v11/n10/abs/nrg2825.html
-
-A comparison of batch effect removal methods for enhancement of prediction performance using MAQC-II microarray gene expression data.
-Luo J, Schumacher M, Scherer A, Sanoudou D, Megherbi D, Davison T, Shi T, Tong W, Shi L, Hong H, Zhao C, Elloumi F, Shi W, Thomas R, Lin S, Tillinghast G, Liu G, Zhou Y, Herman D, Li Y, Deng Y, Fang H, Bushel P, Woods M, Zhang J.
-Pharmacogenomics J. 2010 Aug;10(4):278-91. doi: 10.1038/tpj.2010.57.
-PMID: 20676067 [PubMed - indexed for MEDLINE] Free PMC Article
-http://www.nature.com/tpj/journal/v10/n4/full/tpj201057a.html
-
-Removing batch effects in analysis of expression microarray data: an evaluation of six batch adjustment methods.
-Chen C, Grennan K, Badner J, Zhang D, Gershon E, Jin L, Liu C.
-PLoS One. 2011 Feb 28;6(2):e17238. doi: 10.1371/journal.pone.0017238.
-PMID: 21386892 [PubMed - indexed for MEDLINE] Free PMC Article
-http://www.plosone.org/article/info%3Adoi%2F10.1371%2Fjournal.pone.0017238
-
-T. Hastie, R. Tibshirani, Balasubramanian Narasimhan and Gil Chu (2013). pamr: Pam: prediction analysis for microarrays. R package version
-  1.54.1. http://CRAN.R-project.org/package=pamr
-
-Expanding the understanding of biases in development of clinical-grade molecular signatures: a case study in acute respiratory viral infections.
-Lytkin NI, McVoy L, Weitkamp JH, Aliferis CF, Statnikov A.
-PLoS One. 2011;6(6):e20662. doi: 10.1371/journal.pone.0020662. Epub 2011 Jun 1.
-PMID: 21673802 [PubMed - indexed for MEDLINE] Free PMC Article
-http://www.plosone.org/article/info%3Adoi%2F10.1371%2Fjournal.pone.0020662
-
-TCGA Batch Effects Tool
-http://bioinformatics.mdanderson.org/main/TCGABatchEffects:Overview
-
-GenePattern 2.0.
-Reich M, Liefeld T, Gould J, Lerner J, Tamayo P, Mesirov JP.
-Nat Genet. 2006 May;38(5):500-1. No abstract available.
-PMID: 16642009 [PubMed - indexed for MEDLINE]
-http://www.nature.com/ng/journal/v38/n5/full/ng0506-500.html
-
-The sva package for removing batch effects and other unwanted variation in high-throughput experiments.
-Leek JT, Johnson WE, Parker HS, Jaffe AE, Storey JD.
-Bioinformatics. 2012 Mar 15;28(6):882-3. doi: 10.1093/bioinformatics/bts034. Epub 2012 Jan 17.
-PMID: 22257669 [PubMed - indexed for MEDLINE] Free PMC Article
-http://bioinformatics.oxfordjournals.org/cgi/pmidlookup?view=long&pmid=22257669
-
-virtualArray: a R/bioconductor package to merge raw data from different microarray platforms.
-Heider A, Alt R.
-BMC Bioinformatics. 2013 Mar 2;14:75. doi: 10.1186/1471-2105-14-75.
-PMID: 23452776 [PubMed - indexed for MEDLINE] Free PMC Article
-http://bioconductor.org/packages/release/bioc/html/virtualArray.html
-
-inSilicoDb: an R/Bioconductor package for accessing human Affymetrix expert-curated datasets from GEO.
-Taminau J, Steenhoff D, Coletta A, Meganck S, Lazar C, de Schaetzen V, Duque R, Molter C, Bersini H, Nowé A, Weiss Solís DY.
-Bioinformatics. 2011 Nov 15;27(22):3204-5. doi: 10.1093/bioinformatics/btr529. Epub 2011 Sep 21.
-PMID: 21937664 [PubMed - indexed for MEDLINE] Free Article
-
-Comparing the biological impact of glatiramer acetate with the biological impact of a generic.
-Towfic F, Funt JM, Fowler KD, Bakshi S, Blaugrund E, Artyomov MN, Hayden MR, Ladkani D, Schwartz R, Zeskind B.
-PLoS One. 2014 Jan 8;9(1):e83757. doi: 10.1371/journal.pone.0083757. eCollection 2014.
-PMID: 24421904 [PubMed - in process] Free PMC Article
-http://www.plosone.org/article/info%3Adoi%2F10.1371%2Fjournal.pone.0083757
+\end{document}

### 2014-04-16 tex/manuscript.tex
Update of TeX manuscript
Have converted manuscript to TeX, added box-plot figure, references,
methods, and made some minor rewrites.

@@
 \usepackage[colorlinks=true, urlcolor=citecolor, linkcolor=citecolor, citecolor=citecolor]{hyperref}
+\usepackage[utf8]{inputenc}
+\usepackage[english]{babel}
 
@@
 % List of authors, with corresponding author marked by asterisk
-\author{VEGARD NYGAARD, EINAR ANDREAS R\O{}DLAND, EIVIND HOVIG$^\ast$\\[4pt]
+\author{VEGARD NYGAARD, EINAR ANDREAS RØDLAND, EIVIND HOVIG$^\ast$\\[4pt]
 % Author addresses
@@
 % First field is the short list of authors
-{V. Nygaard, E. A. R\o{}dland, E. Hovig}
+{V. Nygaard, E. A. Rødland, E. Hovig}
 % Second field is the short title of the paper
@@
 In molecular biology, extraneous variables are often called "batch effects", probably due to the fact that reagents and other equipment, for instance microarray chips, are made in batches, and this is frequently observed as an effect in the measurements.
-See 20676067 for more examples.
+See \citet{Luo2010} for more examples.
 
-For a typical experiment comparing group differences, the presence of batch effects will decrease the statistical power.
-If the batch/group design is unbalanced - i.e the proportion of one groups samples belonging to each batch is not equal for all groups - batch may act as a confounder and lead to false positive results as well (20838408).
+For a typical experiment comparing group differences, the presence of batch effects will decrease the statistical power since it adds variation to the data.
+If the batch--group design is unbalanced, i.e if the study groups are not equally represented in all batches, batch effects may act as a confounder and induce false differences between groups (\citealp{Leek2010}).
 
 The standard way to handle an extraneous variable is to include it in the statistical model employed in the inquiry.
-However, many analytic tools for high throughput data does not cater for this option, and when available it could still be outside the competence of the investigator.
-Therefore, an alternative two step procedure has emerged, first the batch effects are estimated and removed, creating a "batch effect free" data set.
-In the next step, the statistical analysis is performed on the adjusted data as there never were batch effects in the first place.
+However, many analysis tools for high throughput data do not cater for this option, and when available it could still be outside the competence of the investigator.
+Therefore, an alternative two step procedure has emerged.
+First the batch effects are estimated and removed, creating a "batch effect free" data set.
+In the next step, the statistical analyses are performed on the adjusted data without further consideration of batch effects.
 This appealing compartmentalization is also convenient for practical purposes, for example when data-processing and statistical analyses are performed by different personnel.
 
-The first step can be achieved by subtracting the mean of the measurements in one batch from all measurements in that batch, i.e mean adjustment or one-way ANOVA adjustment as implemented in the method pamr.batchadjust from the pamr package in R.
-When the batch/group design is balanced, mean-adjustment will take out some, but not necessarily all, variance attributed to batch and leave the between group variance, thus increasing the statistical power.
-However, when batch is a confounder, some of the variance removed will be group variance and statistical power will be lost.
-In addition if there are many groups, some effect will be transferred between groups that share batches, leading to false positive results.
-\NOTE{Dette problemet er ikke beskrevet i de andre artiklene eller i einars mail, men saa vidt jeg skjoenner saa kan dette skje.}
+The first step can be achieved by subtracting the mean of the measurements in one batch from all measurements in that batch, i.e mean adjustment or one-way ANOVA adjustment as implemented in the method \texttt{pamr.batchadjust} from the \texttt{pamr} package in R.
+When the batch--group design is balanced, mean-adjustment will remove most, but not necessarily all, variance attributed to batch and leave the between group variance, thus increasing the statistical power.
+However, when the batch--group design is unbalanced, batch differences will in part be influences by group differences, and thus batch correction will reduce group differences and thereby reduce the statistical power.
+In very uneven group--batch designs with multiple groups, spurious group differences may even be induced in this way.
+Figure~\ref{fig:boxplots} illustrates both these effects.
 
-To mitigate the above problems, group differences can, simplified explained, be estimated based on the balanced part of the data and applied to all samples in the adjusted data set.
-If all the samples, and not only the balanced ones, are used as sample size in later a group comparison, the confidence will be overestimated.
+To mitigate the above problems, one may simultaneously estimate batch effects and group differences, e.g.
+using a two-way ANOVA, and only remove the batch differences from the data.
+Effectively, this means group differences are estimated based on within batch comparison, and applied to the batch adjusted data.
+In a balanced group--batch design, group differences and batch effects are independent, and this approach becomes identical to the above described zero-centering per batch.
+If the group--batch design is heavily unbalanced, estimation of group differences and batch effects are interdependent.
+Unfortunately, when group differences estimated from an unbalanced group--batch design are applied to the entire data set, if this batch adjusted data set is later analysed for group differences, the confidence will be exaggerated as the estimation of the group differences in the unbalanced design is less accurate than in the corresponding balanced design.
+Figure~\ref{fig:boxplots} illustrates how statistical uncertainties are deflated by this batch adjustment method by comparing them to the uncertainties from the original ANOVA.
 
-The ComBat method described in 16632515 and also included in the sva package (22257669) implements the above described procedure in an attempt to retain group differences when adjusting for batch effects.
-Based on actual use of ComBat by the authors and others, we suspect the adjusted data predominantly is treated as "batch effect free" in the subsequent analysis.
-And as a consequence, confidence is overestimated and false results are reported.
+The ComBat method described in \citet{Johnson2007}, and included in the sva package (\citealp{Leek2012}), can use either of the two above described approaches to estimate batch differences, but uses an empirical Bayes approach to avoid over-correction for batch effects for small batches.
+It has popularised the two-way ANOVA procedure for retaining group differences when adjusting for batch effects, and the inclusion of group difference as a covariate when removing batch effects has been recommended \CITE{reference}.
+Based on actual use of ComBat by the authors and others, we suspect thus adjusted data are commonly treated as "batch effect free" in subsequent analysis.
+And as a consequence, confidence in group effects has been overestimated and false results reported.
 
@@
 
-When combining data from more diverse data sources, e.g. microarray data from different platforms, a more general model is required. One such model, used by \CITE{ComBat}, is
+When combining data from more diverse data sources, e.g. microarray data from different platforms, a more general model is required. One such model, used by \citet{Johnson2007}, is
 \begin{equation}
@@
 This permits independent rescaling of data from different batches.
-In addition, \CITE{ComBat} uses an empirical Bayes approach to estimate $\gamma_{ig}$ and $\delta_{ig}$ to stabilise estimates, which is critical for use with small batches.
+In addition, \citet{Johnson2007} uses an empirical Bayes approach to estimate $\gamma_{ig}$ and $\delta_{ig}$ to stabilise estimates, which is critical for use with small batches.
 
@@
 \quad\text{where}\quad
-\bar Y_i=\frac{\sum_{j=1}^M\sum_{r=1}^{n_{ij}} Y_{ijr}}{\sum_{j=1}^M n_{ij}}.
+\overline{Y}_i=\frac{\sum_{j=1}^M\sum_{r=1}^{n_{ij}} Y_{ijr}}{\sum_{j=1}^M n_{ij}}.
 \end{equation}
-An alternative is to centre each batch to the common average by adding the average value $\bar Y$ across the entire data set: i.e.
-$\widetilde Y_{ijr}=\Delta\widetilde Y_{ijr}+\bar Y$.
-When comparing groups, the common value $\bar Y$ has no effect, and so this is equilvalent to zero-centring each batch.
+An alternative is to centre each batch to the common average by adding the average value $\overline{Y}$ across the entire data set: i.e.
+$\widetilde Y_{ijr}=\Delta\widetilde Y_{ijr}+\overline{Y}$.
+When comparing groups, the common value $\overline{Y}$ has no effect, and so this is equilvalent to zero-centring each batch.
 However, if the different groups are unevenly represented in the different batches, the batch average $\overline{Y}_i$ will tend to capture group differences as well as batch effects.
@@
 
-1) In the first experiment(24421904), cells were treated with glatiramer acetate (a medicine for multiple sclerosis) or a generic and mRNA was measured using microarrays alongside control samples.
-A batch effect correlating to the chip (Illumina WG-6\_V2, six samples per chip) was observed and adjusted for with ComBat, whereafter the data was tested for differetially expressed genes, yielding a list of 1000 genes (Table S5 24421904).
+1) In the first experiment(\citealp{Towfic2014}), cells were treated with glatiramer acetate (a medicine for multiple sclerosis) or a generic and mRNA was measured using microarrays alongside control samples.
+A batch effect correlating to the chip (Illumina WG-6\_V2, six samples per chip) was observed and adjusted for with ComBat, whereafter the data was tested for differetially expressed genes, yielding a list of 1000 genes (Table S5, \citealp{Towfic2014}).
 Unfortunately the batch/treatment design was unbalanced with several batches having only one of the main treatments of interest.
-When we re-analyzed their data without using ComBat, but instead blocked for batch effect in limma, only 9 genes were found (FDR<0.05).
+When we re-analyzed their data without using ComBat, but instead blocked for batch effect in limma, only 9 genes were found ($\text{FDR}<0.05$).
 Additional sanity checks with random numbers or perumted labels were also carried out and the distribution of p-values for differents settings are shown in Fig.
 2.
-Our conclusion is that most of the genes reported as differtially expresssed in (24421904) are false positives.
+Our conclusion is that most of the genes reported as differtially expresssed in (\citealp{Towfic2014}) are false positives.
 This example is a sort of "worst case" scenario for applying ComBat, since it both has a very unbalanced batch/group design and a a priori assumpotion of no difference.
@@
 
-%![fig. 2](../reanalysis/Fadi_Towfic_generic_medicine/figure/pvaluestowfic.svg)
-%**Figure 2**.
-%P-value distribution for the main comparison in (data set 2 24421904) glatiramer acetate vs. generic.  
-%- a) Real data adjusted by ComBat as described in the article followed by a significance test in limma. 
-%- b) Real data not adjusted by ComBat, tested for significance in limma using batch as a blocking factor. 
-%- c) Real data is substituted with random numbers from a normal distribution (mean=0, sd=1), but the batch/group design is retained, followed by batch adjustment in ComBat and a significance test in limma.
-%- d) 10 runs of real data with the glatiramer acetate("DP") and generic("N") labels permuted within batches, followed by ComBat adjustment and significance test in limma.  
-%_jeg tok med begge sanitycheckene her. Er d) slik einar foreslo?. Merk jeg har ikke permutert alle labeler, bare de mellom DP og N. Det er en del andre._
+%% FIGURE 2: Original position
 
-2) The second example is taken from the supporting information for the original ComBat article (16632515) where it is denoted "data set 2".
+2) The second example is taken from the supporting information for the original ComBat article (\citealp{Johnson2007}) where it is denoted "data set 2".
 Cells inhibited for the expression of the TAL1 gene were compared to controls on a microarray platform.
@@
 In addition the two sanity checks outline above were performed.
-In contrast to the above results obtained for(24421904), the P-value distributions for the alternative analyses (Fig.3 a,b) does not indicate a huge difference.
+In contrast to the above results obtained for(\citealp{Towfic2014}), the P-value distributions for the alternative analyses (Fig.3 a,b) does not indicate a huge difference.
 Nevertheless, we belive that  P-values are deflated for the ComBat adjusted analysis.
@@
 
-%![fig. 3](../reanalysis/ComBat_original/figure/pvaluesjohnson.svg)  
-%**Figure 3**.
-%P-value distribution for the main comparison in (data set 2 16632515) for  
-%- a) Real data adjusted by ComBat as described in the article followed by a significance test in limma. 
-%- b) Real data not adjusted by ComBat, tested for significance in limma using batch as a blocking factor. 
-%- c) Real data is substituted with random numbers from a normal distribution (mean=0, sd=1), but the batch/group design is retained, followed by batch adjustment in ComBat and a significance test in limma.
-%- d) 10 runs of real data with the group labels permuted within batches, followed by ComBat adjustment and significance test in limma.   
-%_problemet med permutatsjons sanitychechen slik jeg har utfoert og plottet den her er at den neppe kan sies aa feile.
-Dessuten kan de 3 foerst plottene bli plottet i ett som vist under i en alternativ version jeg synes er finere, men dette gaar ikke for permutasjonstesten tror jeg_
+%% FIGURE 3: Original position
 
-%![fig. 4](../reanalysis/ComBat_original/figure/pvaluesjohnson2.svg)   
-%_Alternativ illustrasjon av de tre p-value distribusjonene_
+%% FIGURE 4: Original position
 
 3) \NOTE{Her kan vi fylle paa med flere eksempler etter samme mal hvis det trengs.
-Feks en fersk artikkel fra Nature Genetic([24584070](http://www.nature.com/ng/journal/vaop/ncurrent/full/ng.2916.html)).
+Feks en fersk artikkel fra Nature Genetic (\citealp{Yoo2014}).
 Der har de kombinert sekvenseringsdata  (deres data) fra 9 lymfom pasienter med et (eksternt) microarray data set som har 6 lymfom proever og 5 kontroller, dvs kontroll gruppen er helt fravaerende hos dem.
@@
 
-In the original ComBat article(2007) it is clear that the primary motivation behind ComBat was to employ an Emprical Bayes method for batch-effect removal.
+In the original ComBat article (\citealp{Johnson2007}) it is clear that the primary motivation behind ComBat was to employ an Emprical Bayes method for batch-effect removal.
 The feature of retaining group differences for unbalanced designs is optional and seems to be subordinate, only exemplified in the supplementary information.
@@
 ). \NOTE{Aa linke direkte er kanskje litt ufint, alternativt kan det linkes til forumet uten konkrete innlegg.}  
-And when ComBat was incorporated in the sva package ( [22257669](\href{http://bioinformatics.oxfordjournals.org/cgi/pmidlookup?view=long&pmid=22257669}), 2012), inclusion of group labels was made almost mandatory (an undocumented option of passing a NULL value exists).
+And when ComBat was incorporated in the sva package (\citealp{Leek2012}), inclusion of group labels was made almost mandatory (an undocumented option of passing a NULL value exists).
 Thus, this problematic use has likely been common.
@@
 Upon realizing that the confidence on our group differences were exaggerated, the literature was searched for a better understanding of correct use and potential overseen limitations of ComBat.
-But the authors of ComBat and the sva package recommended our usage (16632515, 22257669).
-In addition, other works looking into the problem of batch effects were mostly recommending ComBat without much concern (22682473, 22133085).
-A brief inquiry into  some of the articles citing ComBat (574 Google Scholar) revealed few problems, and their method descriptions regarding ComBat were mostly sparse, limited to one or two sentences (24391845, 18414638, 21731603,23630272, 24584070).
-A further indication of their carefree use of this potentially devastating procedure was the frequent neglect to state the program parameters, i.e batch labels (18414638, 21731603) or group labels (24391845, 18414638, 21731603, 23630272, 23482648, 24584070).
-Often no effort was done in order to substantiate the existence of batch effects in their data, except for stating the presence of batches (24391845?, 18414638, 21731603, 23630272).
+But the authors of ComBat and the sva package recommended our usage (\citealp{Johnson2007}, \citealp{Leek2012}).
+In addition, other works looking into the problem of batch effects were mostly recommending ComBat without much concern (\citealp{Kupfer2012}, \citealp{Kitchen2011}).
+A brief inquiry into  some of the articles citing ComBat (574 Google Scholar) revealed few problems, and their method descriptions regarding ComBat were mostly sparse, limited to one or two sentences (\CITE{24391845}, \CITE{18414638}, \CITE{21731603}, \CITE{23630272}, \CITE{24584070}).
+A further indication of their carefree use of this potentially devastating procedure was the frequent neglect to state the program parameters, i.e batch labels (\CITE{18414638}, \CITE{21731603}) or group labels (\CITE{24391845}, \CITE{18414638}, \CITE{21731603}, \CITE{23630272}, \CITE{23482648}, \CITE{24584070}).
+Often no effort was done in order to substantiate the existence of batch effects in their data, except for stating the presence of batches (\CITE{24391845}?, \CITE{18414638}, \CITE{21731603}, \CITE{23630272}).
 In one instance ComBat was even applied on data were effects due to batch were investigated, but not found \NOTE{husker ikke ref}.
-The incorporation of the method into analysis pipelines (16642009,TCGA) and other packages (23452776, 21937664) could make its usage even more trivial and parameters setting harder to perceive.
+The incorporation of the method into analysis pipelines (\CITE{16642009}, TCGA) and other packages (\CITE{23452776}, \CITE{21937664}) could make its usage even more trivial and parameters setting harder to perceive.
 Taken together we fear that many published results from data adjusted by ComBat are completely or partially false.
@@
 For such situations, a superficial assessment can be performed, taking special note of batches were groups of interest are near missing and how likely a group difference is.
-In essence asking if the balanced parts (effective sample size?) of the data has enough power to detect the presumed effects and if this is the case (16632515, 24584070), treat the results more like an ordered list with the most likely true positives on top while de-emphasizing the somewhat deflated p-values .
-In contrast, if biological knowledge suggest that a group effect is unlikely  (24391845, 18414638?, 21731603, 23630272), an intermediate lack of batch/group balance  could lead to a mostly false result.
-
-\subsection{Reproducible research}
-\NOTE{vi boer klare aa tilfredstille kravene i http://biostatistics.oxfordjournals.org/content/10/3/405.full}
+In essence asking if the balanced parts (effective sample size?) of the data has enough power to detect the presumed effects and if this is the case (\citealp{Johnson2007}, \CITE{24584070}), treat the results more like an ordered list with the most likely true positives on top while de-emphasizing the somewhat deflated p-values .
+In contrast, if biological knowledge suggest that a group effect is unlikely  (\CITE{24391845}, \CITE{18414638}?, \CITE{21731603}, \CITE{23630272}), an intermediate lack of batch/group balance  could lead to a mostly false result.
 
@@
 
+\section*{Reproducible research}
+\NOTE{vi boer klare aa tilfredstille kravene i http://biostatistics.oxfordjournals.org/content/10/3/405.full}
+
 \section*{Acknowledgments}
@@
 \bibliographystyle{biorefs}
-\bibliography{refs}
+\bibliography{D:/Projects/_Appl_/_TeX_/library}
+%\bibliography{refs}
 
@@
 \begin{figure}[!p]
-\centering\includegraphics[width=15cm]{Fig/pvaluesleeksva.png}
+\centering\includegraphics[width=13cm]{Fig/boxplots.png}
+\caption{Simulated data from four study groups, sampled in three different batches, was generated where groups 1 and 2 have lower means than groups 3 and 4.
+Values and boxes showing mean and two standard errors of the mean are displayed for data without batch effects, after adding batch effects, after batch centering, and after ANOVA based batch centering.
+The last frame shows the least squares estimates of the group means from a two-way ANOVA analysis with 2 standard errors.}
+\label{fig:boxplots}
+\end{figure}
+
+\begin{figure}[!p]
+\centering\includegraphics[width=13cm]{Fig/pvaluesleeksva.png}
 \caption{
@@
 }
-\label{Fig1}
+\label{fig:p-sva}
 \end{figure}
 
-%![fig. 1](../reanalysis/leek_sva/figure/pvaluesleeksva.svg)  
-%**Figure 1**
+\begin{figure}[!p]
+\centering\includegraphics[width=13cm]{Fig/pvaluestowfic.png}
+\caption{
+P-value distribution for the main comparison in (data set 2 24421904) glatiramer acetate vs. generic.  
+a) Real data adjusted by ComBat as described in the article followed by a significance test in limma. 
+b) Real data not adjusted by ComBat, tested for significance in limma using batch as a blocking factor. 
+c) Real data is substituted with random numbers from a normal distribution (mean=0, sd=1), but the batch/group design is retained, followed by batch adjustment in ComBat and a significance test in limma.
+d) 10 runs of real data with the glatiramer acetate("DP") and generic("N") labels permuted within batches, followed by ComBat adjustment and significance test in limma.  
+\NOTE{jeg tok med begge sanitycheckene her. Er d) slik einar foreslo?. Merk jeg har ikke permutert alle labeler, bare de mellom DP og N. Det er en del andre.}
+}
+\label{p-towfic}
+\end{figure}
+
+\begin{figure}[!p]
+\centering\includegraphics[width=13cm]{Fig/pvaluesjohnson.png}
+\caption{
+P-value distribution for the main comparison in (data set 2, \citealp{Johnson2007}) for  
+a) Real data adjusted by ComBat as described in the article followed by a significance test in limma. 
+b) Real data not adjusted by ComBat, tested for significance in limma using batch as a blocking factor. 
+c) Real data is substituted with random numbers from a normal distribution (mean=0, sd=1), but the batch/group design is retained, followed by batch adjustment in ComBat and a significance test in limma.
+d) 10 runs of real data with the group labels permuted within batches, followed by ComBat adjustment and significance test in limma.   
+\NOTE{Problemet med permutatsjons sanitychechen slik jeg har utfoert og plottet den her er at den neppe kan sies aa feile.
+Dessuten kan de 3 foerst plottene bli plottet i ett som vist under i en alternativ version jeg synes er finere, men dette gaar ikke for permutasjonstesten tror jeg}
+}
+\label{fig:p-johnson}
+\end{figure}
+
+\begin{figure}[!p]
+\centering\includegraphics[width=13cm]{Fig/pvaluesjohnson2.png}
+\caption{
+\NOTE{Alternativ illustrasjon av de tre p-value distribusjonene}
+}
+\label{fig:pdist-johnson}
+\end{figure}
 

### 2014-04-16 tex/manuscript.tex
update to clear cache

@@
 % Add history information for the article if required
-\history{}
+%\history{}
+
+%% NOTES TO SELF:
+% Cite using:
+%    \citet{...}  in text
+%    \citealp{...}   within parantheses
+%
+% Temporary notifications and reminders:
+\newcommand\NB[1]{{\color{red}#1}}
+\newcommand\NOTE[1]{{\color{red}[#1]}}
+\newcommand\CITE[1]{{\color{magenta}[#1]}}
 
@@
 Many methods and tools exist for this purpose.
-One method, ComBat which is part of the R package sva, is particularly popular due to its ability to remove batch differences even when batches are small and heterogeneous, while retaining the difference between study groups, e.g.
-treatments.
-Unfortunately, this recommended and frequently used approach to batch correction will systematically induce incorrect group differences in downstream analyses when groups are unevenly distributed between the batches.
+One method, ComBat which is part of the R package sva, is particularly popular due to its ability to remove batch differences even when batches are small and heterogeneous.
+It also has the option, recommended by the authors \NOTE{Anbefalt av dem alle eller kun av Johnson?}, of preserving the difference between study groups, estimated from a two-way ANOVA model, to avoid conflating batch effects and group differences during batch adjustments.
+Unfortunately, this recommended and frequently used approach may systematically induce incorrect group differences in downstream analyses when groups are distributed between the batches in an unbalanced manner.
 The scientific community seems to be largely unaware of this problem, which most likely has contributed to false discoveries being presented in the published literature.
@@
 \end{abstract}
-
-%% NOTES TO SELF:
-% Cite using:
-%    \citet{...}  in text
-%    \citealp{...}   within parantheses
-%
-% Temporary notifications and reminders:
-\newcommand\NB[1]{{\color{red}#1}}
-\newcommand\NOTE[1]{{\color{red}[#1]}}
-\newcommand\CITE[1]{{\color{magenta}[#1]}}
 
@@
 Therefore, an alternative two step procedure has emerged.
-First the batch effects are estimated and removed, creating a "batch effect free" data set.
+First the batch effects are estimated and removed, creating a
+``batch effect free'' data set.
 In the next step, the statistical analyses are performed on the adjusted data without further consideration of batch effects.
@@
 
-We will base out discussion of data with batch effects on a simple model:
+We will base our discussion of data with batch effects on a simple model:
 \begin{equation}
@@
 \centering\includegraphics[width=13cm]{Fig/boxplots.png}
-\caption{Simulated data from four study groups, sampled in three different batches, was generated where groups 1 and 2 have lower means than groups 3 and 4.
+\caption{Simulated data from four study groups was generated where groups 1 and 2 have lower means than groups 3 and 4.
+These were placed in three different batches with batch effect added.
 Values and boxes showing mean and two standard errors of the mean are displayed for data without batch effects, after adding batch effects, after batch centering, and after ANOVA based batch centering.
-The last frame shows the least squares estimates of the group means from a two-way ANOVA analysis with 2 standard errors.}
+The last frame shows the least squares estimates of the group means from a two-way ANOVA analysis with 2 standard errors.
+This case, design and effects, was selected to illustrate the spurious effects that may arise.}
 \label{fig:boxplots}

### 2014-04-22 tex/manuscript.tex
update again
update
Added subsection on two-group comparison

@@
 Therefore, an alternative two step procedure has emerged.
-First the batch effects are estimated and removed, creating a
-``batch effect free'' data set.
+First the batch effects are estimated and removed, creating a ``batch effect free'' data set.
 In the next step, the statistical analyses are performed on the adjusted data without further consideration of batch effects.
@@
 \begin{equation}
-\Delta\widetilde Y_{ijr}=Y_{ijr}-\overline{Y}_{i}
+\Delta\widetilde Y_{ijr}=Y_{ijr}-\bar{Y}_{i}
 \quad\text{where}\quad
-\overline{Y}_i=\frac{\sum_{j=1}^M\sum_{r=1}^{n_{ij}} Y_{ijr}}{\sum_{j=1}^M n_{ij}}.
+\bar{Y}_i=\frac{\sum_{j=1}^M\sum_{r=1}^{n_{ij}} Y_{ijr}}{\sum_{j=1}^M n_{ij}}.
 \end{equation}
-An alternative is to centre each batch to the common average by adding the average value $\overline{Y}$ across the entire data set: i.e.
-$\widetilde Y_{ijr}=\Delta\widetilde Y_{ijr}+\overline{Y}$.
-When comparing groups, the common value $\overline{Y}$ has no effect, and so this is equilvalent to zero-centring each batch.
-However, if the different groups are unevenly represented in the different batches, the batch average $\overline{Y}_i$ will tend to capture group differences as well as batch effects.
+An alternative is to centre each batch to the common average by adding the average value $\bar{Y}$ across the entire data set: i.e.
+$\widetilde Y_{ijr}=\Delta\widetilde Y_{ijr}+\bar{Y}$.
+When comparing groups, the common value $\bar{Y}$ has no effect, and so this is equilvalent to zero-centring each batch.
+However, if the different groups are unevenly represented in the different batches, the batch average $\bar{Y}_i$ will tend to capture group differences as well as batch effects.
 Thus, batch centering may reduce group differences, and thus reduce the power of downstream analyses.
@@
 
-\NB{Formler og ord:
-Forklare hva som gaar galt med formler og ord. Overlater til Einar aa skrive dette.}
+\subsection{Explanation for the simple two-group comparison}
+
+To explain more clearly what is happening, and quantify the size of the problem, we may consider the simple case of estimating the difference $\Delta\beta=\beta_A-\beta_B$ between two groups, $A$ and $B$, when there are $m$ batches with batch $i$ containing $n_{iA}$ and $n_{iB}$ samples from each of the two groups for $i=1,\ldots,m$.
+
+If we estimate the group difference within batch $i$, we get
+\begin{equation}
+\widehat{\Delta\beta}_i=\bar{Y}_{iA}-\bar{Y}_{iB}\sim\text{N}\left(\Delta\beta,\frac{\sigma^2}{\nu_i}\right)
+\quad\text{where}\quad
+\bar{Y}_{ij}=\frac{\sum_{r=1}^{n_{ij}}Y_{ijr}}{n_{ij}},\,
+\nu_i=\frac{1}{\frac{1}{n_{iA}}+\frac{1}{n_{iB}}},
+\end{equation}
+from which we may express the overall estimate of $\Delta\beta$
+\begin{equation}
+\widehat{\Delta\beta}=\frac{\sum_{i=1}^m \nu_i\,\widehat{\Delta\beta}_i}{\nu}
+\sim\text{N}\left(\Delta\beta,\frac{\sigma^2}{\nu}\right)
+\quad\text{where}\quad
+\nu=\sum_{i=1}^m\nu_i.
+\end{equation}
+
+If batch and group effects are estimated using a two-way ANOVA, the estimate $\widehat{\Delta\beta}$ will be as stated above, and so the estimated group difference is unaffected.
+The batch effects are then removed, and the estimated group differences retained, leaving the estimated $\widehat{\Delta\beta}$ unchanged by the batch adjustment.
+However, if this batch adjusted data set is analysed without considering batch effects, the variance of $\widehat{\Delta\beta}$ will be computed under the assumption that it is derived from a comparison of $n_A=\sum_{i=1}^m n_{iA}$ versus $n_B=\sum_{i=1}^m n_{iB}$ samples, and thus satisfy
+\begin{equation}
+\widehat{\Delta\beta}
+\sim\text{N}\left(\Delta\beta,\sigma^2\cdot\left(\frac{1}{n_A}+\frac{1}{n_B}\right)\right).
+\end{equation}
+Using Jensen's inequality, we may then derive
+\begin{equation}
+\frac{1}{\frac{1}{n_A}+\frac{1}{n_B}}=n\cdot\frac{n_A}{n}\cdot\frac{n_B}{n}
+\le
+\nu
+=\sum_{i=1}^m \frac{1}{\frac{1}{n_{iA}}+\frac{1}{n_{iB}}}
+=\sum_{i=1}^m n_i\cdot\frac{n_{iA}}{n_i}\cdot\frac{n_{iB}}{n_i}
+\end{equation}
+with equality if and only if the ratios $n_{iA}:n_{iB}=n_A:n_B$ for all batches $i=1,\ldots,m$.
 

### 2014-05-26 tex/manuscript.tex
Updating version from May 2

@@
 % Temporary notifications and reminders:
-\newcommand\NB[1]{{\color{red}#1}}
-\newcommand\NOTE[1]{{\color{red}[#1]}}
-\newcommand\CITE[1]{{\color{magenta}[#1]}}
+\usepackage{color}
+\definecolor{NBcol}{rgb}{0.7,0,1}
+\definecolor{CITEcol}{rgb}{1,0,0.7}
+\definecolor{COMMENTcol}{rgb}{0.7,0.2,0}
+\newcommand\COMMENT[2]{\textcolor{COMMENTcol}{#1}\NOTE{#2}}
+\newcommand\NB[1]{\textcolor{NBcol}{\textit{#1}}}
+\newcommand\NOTE[1]{\NB{[#1]}}
+\newcommand\CITE[1]{\textcolor{CITEcol}{[#1]}}
 
 \begin{document}
+
+\footnotetext{\color{blue}Manuscript version \today}
 
@@
 This appealing compartmentalization is also convenient for practical purposes, for example when data-processing and statistical analyses are performed by different personnel.
+Unfortunately, as we demonstrate in this paper, when the batch--group design is unbalanced, this approach may be unreliable.
 
-The first step can be achieved by subtracting the mean of the measurements in one batch from all measurements in that batch, i.e mean adjustment or one-way ANOVA adjustment as implemented in the method \texttt{pamr.batchadjust} from the \texttt{pamr} package in R.
+A simple removal of batch effects can be achieved by subtracting the mean of the measurements in one batch from all measurements in that batch, i.e mean adjustment or one-way ANOVA adjustment as implemented in the method \texttt{pamr.batchadjust} from the \texttt{pamr} package in R.
 When the batch--group design is balanced, mean-adjustment will remove most, but not necessarily all, variance attributed to batch and leave the between group variance, thus increasing the statistical power.
-However, when the batch--group design is unbalanced, batch differences will in part be influences by group differences, and thus batch correction will reduce group differences and thereby reduce the statistical power.
+However, when the batch--group design is unbalanced, batch differences will in part be influenced by group differences, and thus batch correction will reduce group differences and thereby reduce the statistical power.
 In very uneven group--batch designs with multiple groups, spurious group differences may even be induced in this way.
@@
 using a two-way ANOVA, and only remove the batch differences from the data.
-Effectively, this means group differences are estimated based on within batch comparison, and applied to the batch adjusted data.
+Effectively, group differences are estimated based on within batch comparisons, and applied to the batch adjusted data.
 In a balanced group--batch design, group differences and batch effects are independent, and this approach becomes identical to the above described zero-centering per batch.
 If the group--batch design is heavily unbalanced, estimation of group differences and batch effects are interdependent.
-Unfortunately, when group differences estimated from an unbalanced group--batch design are applied to the entire data set, if this batch adjusted data set is later analysed for group differences, the confidence will be exaggerated as the estimation of the group differences in the unbalanced design is less accurate than in the corresponding balanced design.
+However, in applying the estimated group differences across the entire data set, the uncertainties of these estimates are ignored.
+If the group--batch design is unbalanced and batch compositions ignored after batch adjustments have been made, later analyses will systematically underestimate the statistical uncertainties and exaggerate the confidence of group differences.
 Figure~\ref{fig:boxplots} illustrates how statistical uncertainties are deflated by this batch adjustment method by comparing them to the uncertainties from the original ANOVA.
@@
 The ComBat method described in \citet{Johnson2007}, and included in the sva package (\citealp{Leek2012}), can use either of the two above described approaches to estimate batch differences, but uses an empirical Bayes approach to avoid over-correction for batch effects for small batches.
-It has popularised the two-way ANOVA procedure for retaining group differences when adjusting for batch effects, and the inclusion of group difference as a covariate when removing batch effects has been recommended \CITE{reference}.
-Based on actual use of ComBat by the authors and others, we suspect thus adjusted data are commonly treated as "batch effect free" in subsequent analysis.
+It has thus improved and popularised the two-way ANOVA procedure for retaining group differences when adjusting for batch effects, and the inclusion of group difference as a covariate when removing batch effects has been recommended \CITE{reference}.
+Based on actual use of ComBat by the authors and others, we suspect thus adjusted data are commonly treated as      ``batch effect free'' in subsequent analysis.
 And as a consequence, confidence in group effects has been overestimated and false results reported.
 
-\NB{Kort oppsummere behovet for batch-korrigering: generelt og i microarraydata.}
-
-\NB{Kort beskrive metoder som gjoer dette (batch-sentrering, ANOVA, ComBat).}
-
-\NB{Forklare tilsiktet effekt disse paa batch+gruppe-forskjeller (balansert versus ikke-balansert).}
-
-\NB{Forklare kort hva faktisk effekt er: ANOVA gir redusert gruppe-forskjeller, mens for ComBat overestimeres den statistiske styrken/sikkerheten av forskjellen.}
+Statisticians would most likely take extra precautions should the sample or batch sizes be very small.
+However, the effects of batch adjustment using two-way ANOVA or ComBat on unbalanced data sets remain even as sample and batch sizes increase.
+For example, group comparisons using one-way ANOVA on the batch adjusted data will essentially result in $F$ statistics that are inflated by a fixed factor which depends on the unevenness of the design rather than the size of the sample or batches.
+The effect of this may be further exacerbated by running these analyses a large number of times, e.g.
+on thousands of genes, and use false discovery rate to determine significant cases: an approach that is particularly sensitive to inflated false positive rates.
 
@@
 
-We will base our discussion of data with batch effects on a simple model:
+We will base our discussion on a simple model for data with batch effects:
 \begin{equation}
@@
 $\delta_{ig}=1$.
-We will discuss the effect of empirical Bayes estimation of $\gamma_{ig}$ later, but out main argument is more easily made in the simpler context.
+We will discuss the effect of empirical Bayes estimation of $\gamma_{ig}$ later, but our main argument is more easily made in the simpler context.
+Empirical Bayes may reduce the amount of batch adjustment in cases where batch effects are small or cannot be accurately estimated, e.g.
+for small batches.
+However, in cases with large batches or substantial batch effects, it should differ little from the two-way ANOVA approach.
 
@@
 
-The main ambition of batch effect adjustments is to be able to remove batch differences in such a way that downstream analyses of the adjusted data may be done without further batch adjustments.
+A common ambition of batch effect adjustments is to remove batch differences in such a way that downstream analyses of the adjusted data may be done without further corrections for batches.
 
@@
 When comparing groups, the common value $\bar{Y}$ has no effect, and so this is equilvalent to zero-centring each batch.
-However, if the different groups are unevenly represented in the different batches, the batch average $\bar{Y}_i$ will tend to capture group differences as well as batch effects.
-Thus, batch centering may reduce group differences, and thus reduce the power of downstream analyses.
+If the groups are unevenly represented in the different batches, the batch average $\bar{Y}_i$ will tend to capture group differences as well as batch effects:
+\begin{equation}
+\Delta\widetilde{Y}_{ijr}=\beta_j-\bar{\beta}_i+\epsilon_{ijr}-\bar{\epsilon}_i
+\quad\text{where}\quad
+\bar{\beta}_i=\frac{\sum_{j=1}^M n_{ij}\beta_j}{\sum_{j=1}^M n_{ij}},\;
+\bar{\epsilon}_i=\frac{\sum_{j=1}^M\sum_{r=1}^{n_{ij}} \epsilon_{ijr}}{\sum_{j=1}^M n_{ij}}.
+\end{equation}
+Thus, batch centering will tend to reduce group differences in an unbalanced design, and thus reduce the power of downstream analyses.
+By reducing the differences between some groups, i.e.
+those found together in the same batches, it may also induce false differences between other groups.
 
-Removing batch effects while retaining group differences can be done through an ANOVA analysis in which the group effects, $\beta_j$, and the batch effects, $\gamma_i$, are estimated simultaneously.
-Batch adjusted values may then be obtained by subtracting the estimated batch effects, $\hat\beta_j$:
+Removing batch effects while retaining group differences can be done through a two-way ANOVA in which group effects, $\beta_j$, and batch effects, $\gamma_i$, are estimated simultaneously.
+Batch adjusted values may then be obtained by subtracting the estimated batch effects, $\hat\gamma_i$:
 \begin{equation}
-\widetilde Y_{ijr}=Y_{ijr}-\hat\beta_j=\alpha+(\beta_j-\hat\beta_j)+\gamma_i+\epsilon_{ijr}.
+\widetilde Y_{ijr}=Y_{ijr}-\hat\beta_j=\alpha+\beta_j+(\gamma_i-\hat\gamma_i)+\epsilon_{ijr}.
 \end{equation}
@@
 
-Unfortunately, the estimation error $\hat\beta_j-\beta_j$ affects all values within the same batch in the same manner, inducing a dependency between the values.
-If the study groups are evenly represented in all batches, this will not influence estimated group differences as all groups are equally affected.
-However, if groups are unevenly represented, this induced dependency can have severe impact on downstream analyses.
+The estimation error $\hat\gamma_i-\gamma_i$ affects all values within the same batch in the same manner.
+Thus, while the aim is to remove spurious dependencies within batches, it may also induce new dependencies.
+In a balanced group--batch design, this has the same effect on all groups and thus does not influence group comparisons.
+In an unbalanced design, however, the estimated batch effects $\hat\gamma_i$ will correlate with the estimated group effects $\hat\beta_j$.
+\COMMENT{Assuming the batch adjusted values are without further batch differences corresponds to assuming that $\gamma_i=\hat\gamma_i$, and ignoring the estimation uncertainty and its impact on the correlated $\hat\beta_j$.
+It results in over-confidence in the accuracy of the group differences, which can have severe impact on downstream analyses.}{Denne forklaringen må jeg gjøre mye klarere.}
 
@@
 The documentation accompanying the sva library has a runnable example demonstrating how to adjust a data set with ComBat followed by a F-test.
-Swapping the real data with random numbers from a normal distribution (mean=0, sd=1), but otherwise following the instructions, will generate the p-value distribution shown in Figure 1.
-The skewed distribution is a indication that this approach may have a unintentional adverse effect.
-\NOTE{Jeg har ikke beskrevet permutatsjon checken fordi jeg er usikker paa om vi skal bruke den.}
+Swapping the real data with random numbers from a normal distribution (mean=0, sd=1), but otherwise following the instructions, will generate the p-value distribution shown in Figure~\ref{fig:p-sva}.
+\NOTE{Perhaps show Q-Q plot of $F$-statistics.} The skewed distribution is a indication that this approach may have a unintentional adverse effect.
+\NOTE{Tror vi dropper permutasjonstesten.}
 
-%% FIGURE 1: Original position
+If the number of samples is increased, the empirical Bayes estimates used by ComBat will shrink the batch effect estimates and thus moderate the batch adjustments.
+However, if random data with different means in each batch, but no difference betwene groups, are drawn, the problem remains.
+\NOTE{Bør dette komme her eller senere?
+Evt.
+understøttes?}
+
+%% FIGURE 1=fig:p-sva: Original position
 
@@
 
-1) In the first experiment(\citealp{Towfic2014}), cells were treated with glatiramer acetate (a medicine for multiple sclerosis) or a generic and mRNA was measured using microarrays alongside control samples.
+\subsubsection{Experiment 1}
+
+In the first experiment(\citealp{Towfic2014}), cells were treated with glatiramer acetate (a medicine for multiple sclerosis) or a generic and mRNA was measured using microarrays alongside control samples.
 A batch effect correlating to the chip (Illumina WG-6\_V2, six samples per chip) was observed and adjusted for with ComBat, whereafter the data was tested for differetially expressed genes, yielding a list of 1000 genes (Table S5, \citealp{Towfic2014}).
@@
 When we re-analyzed their data without using ComBat, but instead blocked for batch effect in limma, only 9 genes were found ($\text{FDR}<0.05$).
-Additional sanity checks with random numbers or perumted labels were also carried out and the distribution of p-values for differents settings are shown in Fig.
-2.
+\NOTE{Also run analysis using a mixed random effects model which assumes random batch effects much like in the empirical Bayes model.} Additional sanity checks with random numbers or perumted labels were also carried out and the distribution of p-values for differents settings are shown in Figure~\ref{fig:p-towfic}.
 Our conclusion is that most of the genes reported as differtially expresssed in (\citealp{Towfic2014}) are false positives.
@@
 
-%% FIGURE 2: Original position
+%% FIGURE 2=fig:p-towfic: Original position
 
-2) The second example is taken from the supporting information for the original ComBat article (\citealp{Johnson2007}) where it is denoted "data set 2".
+\subsubsection{Experiment 2}
+
+The second example is taken from the supporting information for the original ComBat article (\citealp{Johnson2007}) where it is denoted "data set 2".
 Cells inhibited for the expression of the TAL1 gene were compared to controls on a microarray platform.
@@
 In addition the two sanity checks outline above were performed.
-In contrast to the above results obtained for(\citealp{Towfic2014}), the P-value distributions for the alternative analyses (Fig.3 a,b) does not indicate a huge difference.
+In contrast to the above results obtained for(\citealp{Towfic2014}), the P-value distributions for the alternative analyses (Figure~\ref{fig:p-johnson} a,b) does not indicate a huge difference.
 Nevertheless, we belive that  P-values are deflated for the ComBat adjusted analysis.
@@
 
-%% FIGURE 3: Original position
+%% FIGURE 3=fig:p-johnson: Original position
 
-%% FIGURE 4: Original position
+%% FIGURE 4=fig:pdist-johnson: Original position
 
-3) \NOTE{Her kan vi fylle paa med flere eksempler etter samme mal hvis det trengs.
+\subsubsection{More?}
+
+\NOTE{Her kan vi fylle paa med flere eksempler etter samme mal hvis det trengs.
 Feks en fersk artikkel fra Nature Genetic (\citealp{Yoo2014}).
@@
 
+\NB{Systematic underestimation of $P$-values combined with FDR is a particularly bad combination as FDR lends strength across tests.}
+
 \subsection{Increased emphasis on preserving group difference}
 
-In the original ComBat article (\citealp{Johnson2007}) it is clear that the primary motivation behind ComBat was to employ an Emprical Bayes method for batch-effect removal.
+In the original ComBat article (\citealp{Johnson2007}) it is clear that the primary motivation behind ComBat was to employ an Emprical Bayes method for batch-effect removal to allow better handling of small batches.
 The feature of retaining group differences for unbalanced designs is optional and seems to be subordinate, only exemplified in the supplementary information.
@@
 ). \NOTE{Aa linke direkte er kanskje litt ufint, alternativt kan det linkes til forumet uten konkrete innlegg.}  
-And when ComBat was incorporated in the sva package (\citealp{Leek2012}), inclusion of group labels was made almost mandatory (an undocumented option of passing a NULL value exists).
-Thus, this problematic use has likely been common.
+When ComBat was later incorporated in the sva package (\citealp{Leek2012}), \COMMENT{batch adjustment without covariates was no longer a default as specification of covariates became a requirement, although specification of a null model was still possible although not clearly documented in the sva documentation}{Is this correct?}.
+Thus, we expect this usage to have grown more common.
 
@@
 Taken together we fear that many published results from data adjusted by ComBat are completely or partially false.
-And knowing that scientists don't give up where one analysis fails, a method that almost ensures a result given a sufficiently unbalanced design will continue to be used.
+\COMMENT{And knowing that scientists don't give up where one analysis fails, a method that almost ensures a result given a sufficiently unbalanced design will continue to be used.}{Dette bør nok formuleres anderledes.} \NOTE{Her bør vi nok fokusere tydeligere på bruk der gruppe er brukt som kovariat.
+Om dette er uklart, bør vi sjekke det før vi lister det opp.}
 

### 2014-05-26 tex/manuscript.tex
Lagt inn rettelser

@@
 % Second field is the short title of the paper
-{Exagerated effects after batch correction}
+{Exaggerated effects after batch correction}
 
@@
 One method, ComBat which is part of the R package sva, is particularly popular due to its ability to remove batch differences even when batches are small and heterogeneous.
-It also has the option, recommended by the authors \NOTE{Anbefalt av dem alle eller kun av Johnson?}, of preserving the difference between study groups, estimated from a two-way ANOVA model, to avoid conflating batch effects and group differences during batch adjustments.
+It also has the option of preserving the difference between study groups, estimated from a two-way ANOVA model, to avoid conflating batch effects and group differences during batch adjustments.
 Unfortunately, this recommended and frequently used approach may systematically induce incorrect group differences in downstream analyses when groups are distributed between the batches in an unbalanced manner.
@@
 The ComBat method described in \citet{Johnson2007}, and included in the sva package (\citealp{Leek2012}), can use either of the two above described approaches to estimate batch differences, but uses an empirical Bayes approach to avoid over-correction for batch effects for small batches.
-It has thus improved and popularised the two-way ANOVA procedure for retaining group differences when adjusting for batch effects, and the inclusion of group difference as a covariate when removing batch effects has been recommended \CITE{reference}.
-Based on actual use of ComBat by the authors and others, we suspect thus adjusted data are commonly treated as      ``batch effect free'' in subsequent analysis.
+It has thus improved and popularised the two-way ANOVA procedure for retaining group differences when adjusting for batch effects, and the inclusion of group difference as a covariate when removing batch effects has been recommended  in the sva tutorial and user fora.
+Based on actual use of ComBat by the authors and others, ourselves included, we suspect thus adjusted data are commonly treated as ``batch effect free'' in subsequent analysis.
 And as a consequence, confidence in group effects has been overestimated and false results reported.
@@
 ComBat was applied followed by a t-test in order to identify differentially expressed genes.
-First, we reproduced their analysis including the adjustment by ComBat, but using limma instead of the t-test, resulting in 1003 probes (q<0.05).
-Then, we analysed their data without batch adjustment in ComBat, but blocking for batch in limma, resulting in 377 probes (q<0.05).
-In addition the two sanity checks outline above were performed.
+First, we reproduced their analysis including the adjustment by ComBat, but using limma instead of the t-test, resulting in 1003 probes ($q<0.05$).
+Then, we analysed their data without batch adjustment in ComBat, but blocking for batch in limma, resulting in 377 probes ($q<0.05$).
+In addition, the two sanity checks outlined above were performed.
 In contrast to the above results obtained for(\citealp{Towfic2014}), the P-value distributions for the alternative analyses (Figure~\ref{fig:p-johnson} a,b) does not indicate a huge difference.

### 2014-06-18 tex/manuscript.tex
Cleanup

@@
 
+\NOTE{Er det denne emailen du vil gi fra deg? Var det ikke snakk om aa pensjonere radium emailer for noen aar siden? Vil denne virke om to år?}
+
 % Add a footnote for the corresponding author if one has been
@@
 It also has the option of preserving the difference between study groups, estimated from a two-way ANOVA model, to avoid conflating batch effects and group differences during batch adjustments.
-Unfortunately, this recommended and frequently used approach may systematically induce incorrect group differences in downstream analyses when groups are distributed between the batches in an unbalanced manner.
+Unfortunately, this frequently used and recommended approach may systematically induce incorrect group differences in downstream analyses when groups are distributed between the batches in an unbalanced manner.
 The scientific community seems to be largely unaware of this problem, which most likely has contributed to false discoveries being presented in the published literature.
@@
 It has thus improved and popularised the two-way ANOVA procedure for retaining group differences when adjusting for batch effects, and the inclusion of group difference as a covariate when removing batch effects has been recommended  in the sva tutorial and user fora.
-Based on actual use of ComBat by the authors and others, ourselves included, we suspect thus adjusted data are commonly treated as ``batch effect free'' in subsequent analysis.
+Based on actual use of ComBat by the authors and others, ourselves included, we suspect thus adjusted data are commonly treated as ``batch effect free'' in subsequent analyses.
 And as a consequence, confidence in group effects has been overestimated and false results reported.
@@
 
-As the amount of false positive results when trying to retain group differences depends on the batch/group balance, we will show two examples with varying degree of unbalancedness.
+As the amount of false positive results when trying to retain group differences depends on the batch--group balance, we will show two examples with varying degree of unbalancedness.
 
@@
 In the first experiment(\citealp{Towfic2014}), cells were treated with glatiramer acetate (a medicine for multiple sclerosis) or a generic and mRNA was measured using microarrays alongside control samples.
-A batch effect correlating to the chip (Illumina WG-6\_V2, six samples per chip) was observed and adjusted for with ComBat, whereafter the data was tested for differetially expressed genes, yielding a list of 1000 genes (Table S5, \citealp{Towfic2014}).
-Unfortunately the batch/treatment design was unbalanced with several batches having only one of the main treatments of interest.
+A batch effect correlating to the chip (Illumina WG-6\_V2, six samples per chip) was observed and adjusted for with ComBat, whereafter the data was tested for differentially expressed genes, yielding a list of 1000 genes (Table S5, \citealp{Towfic2014}).
+Unfortunately the batch--treatment design was unbalanced with several batches having only one of the main treatments of interest.
 When we re-analyzed their data without using ComBat, but instead blocked for batch effect in limma, only 9 genes were found ($\text{FDR}<0.05$).
-\NOTE{Also run analysis using a mixed random effects model which assumes random batch effects much like in the empirical Bayes model.} Additional sanity checks with random numbers or perumted labels were also carried out and the distribution of p-values for differents settings are shown in Figure~\ref{fig:p-towfic}.
-Our conclusion is that most of the genes reported as differtially expresssed in (\citealp{Towfic2014}) are false positives.
-This example is a sort of "worst case" scenario for applying ComBat, since it both has a very unbalanced batch/group design and a a priori assumpotion of no difference.
-The R-code for our analysis and a more extensive report can be downloaded from github (\href{https://github.com/vegardny/combat_tests.git}).
+\NOTE{Also run analysis using a mixed random effects model which assumes random batch effects much like in the empirical Bayes model.} The above outlined sanity check with random numbers was also carried out.
+The distribution of p-values for different settings are shown in Figure~\ref{fig:p-towfic}.
+Our conclusion is that most of the genes reported as differentially expressed in (\citealp{Towfic2014}) are false positives.
+This example is a sort of ``worst case'' scenario for applying ComBat, since it both has a very unbalanced batch--group design and a a priori assumption of no difference.
+The R-code for our analysis and a more extensive report can be downloaded from \href{https://github.com/vegardny/combat_tests.git}{Github} \NOTE{Update reference}.
 
@@
 
-The second example is taken from the supporting information for the original ComBat article (\citealp{Johnson2007}) where it is denoted "data set 2".
+The second example is taken from the supporting information for the original ComBat article (\citealp{Johnson2007}) where it is denoted ``Data set 2''.
 Cells inhibited for the expression of the TAL1 gene were compared to controls on a microarray platform.
-The experiment was conducted on three different time points (used as batches) with a total of 30 samples and a fairly balanced batch/treatment set up (6-2, 3-4 and 9-6).
+The experiment was conducted on three different time points (used as batches) with a total of 30 samples and a fairly balanced batch--treatment set up (6-2, 3-4 and 9-6).
 ComBat was applied followed by a t-test in order to identify differentially expressed genes.
@@
 Then, we analysed their data without batch adjustment in ComBat, but blocking for batch in limma, resulting in 377 probes ($q<0.05$).
-In addition, the two sanity checks outlined above were performed.
-In contrast to the above results obtained for(\citealp{Towfic2014}), the P-value distributions for the alternative analyses (Figure~\ref{fig:p-johnson} a,b) does not indicate a huge difference.
+In addition, the sanity checks outlined above were performed.
+The distribution of P-values for different settings are shown in Figure~\ref{fig:p-values}b.
+In contrast to the results obtained for\citealp{Towfic2014} (Figure~\ref{fig:p-values}a), the P-value distributions for the alternative analysis does not indicate a huge difference.
 Nevertheless, we belive that  P-values are deflated for the ComBat adjusted analysis.
-The R-code for our analysis and a more extensive report can be downloaded from github (\href{https://github.com/vegardny/combat_tests.git}).
+The R-code for our analysis can be downloaded from \href{https://github.com/vegardny/combat_tests.git}{Github} \NOTE{Update reference}.
 
@@
 The feature of retaining group differences for unbalanced designs is optional and seems to be subordinate, only exemplified in the supplementary information.
-However, over the years this feature became more important judging from advice given by the author to users
-(
-[1](\href{https://groups.google.com/d/msg/combat-user-forum/eSVSKwGtuyE/ZIWV2juYmmAJ})
-,[2](\href{https://groups.google.com/d/msg/combat-user-forum/Vkb9p7wekd4/h5Etie7FTVQJ})
-,[3](\href{https://groups.google.com/d/msg/combat-user-forum/fpBTcgDjiR8/uo4QIZL4sZgJ})
-,[4](\href{https://groups.google.com/d/msg/combat-user-forum/26FZlgU2LFQ/W6U_Lhh_64EJ})
-). \NOTE{Aa linke direkte er kanskje litt ufint, alternativt kan det linkes til forumet uten konkrete innlegg.}  
+However, over the years this feature became more important judging from advice given on user fora.
 When ComBat was later incorporated in the sva package (\citealp{Leek2012}), \COMMENT{batch adjustment without covariates was no longer a default as specification of covariates became a requirement, although specification of a null model was still possible although not clearly documented in the sva documentation}{Is this correct?}.
@@
 
-Our knowledge of ComBat came through a typical use case when trying to salvage unbalanced data which had batch effects.
+Our knowledge of the problem discussed in this article came through a typical use case when trying to adjust for batch effects in an unbalanced data set using ComBat.
 Upon realizing that the confidence on our group differences were exaggerated, the literature was searched for a better understanding of correct use and potential overseen limitations of ComBat.
@@
 In addition, other works looking into the problem of batch effects were mostly recommending ComBat without much concern (\citealp{Kupfer2012}, \citealp{Kitchen2011}).
-A brief inquiry into  some of the articles citing ComBat (574 Google Scholar) revealed few problems, and their method descriptions regarding ComBat were mostly sparse, limited to one or two sentences (\CITE{24391845}, \CITE{18414638}, \CITE{21731603}, \CITE{23630272}, \CITE{24584070}).
-A further indication of their carefree use of this potentially devastating procedure was the frequent neglect to state the program parameters, i.e batch labels (\CITE{18414638}, \CITE{21731603}) or group labels (\CITE{24391845}, \CITE{18414638}, \CITE{21731603}, \CITE{23630272}, \CITE{23482648}, \CITE{24584070}).
-Often no effort was done in order to substantiate the existence of batch effects in their data, except for stating the presence of batches (\CITE{24391845}?, \CITE{18414638}, \CITE{21731603}, \CITE{23630272}).
-In one instance ComBat was even applied on data were effects due to batch were investigated, but not found \NOTE{husker ikke ref}.
+A brief inquiry into  some of the articles citing ComBat (574 Google Scholar) revealed few problems, and their method descriptions regarding ComBat were mostly sparse, limited to one or two sentences.
+A further indication of their carefree use of this potentially devastating procedure was the frequent omittance of program parameters that were used, i.e batch labels or whether group labels were supplied as covariates.
+Often no effort was done in order to substantiate the existence of batch effects in their data, except for stating the presence of batches.
 The incorporation of the method into analysis pipelines (\CITE{16642009}, TCGA) and other packages (\CITE{23452776}, \CITE{21937664}) could make its usage even more trivial and parameters setting harder to perceive.
 Taken together we fear that many published results from data adjusted by ComBat are completely or partially false.
-\COMMENT{And knowing that scientists don't give up where one analysis fails, a method that almost ensures a result given a sufficiently unbalanced design will continue to be used.}{Dette bør nok formuleres anderledes.} \NOTE{Her bør vi nok fokusere tydeligere på bruk der gruppe er brukt som kovariat.
-Om dette er uklart, bør vi sjekke det før vi lister det opp.}
+We hope that our warning will enlighten the community and  reduce this unfortunate combination of methods.
 
@@
 
-We have shown that adjusting for batch effects while preserve the group difference may lead to varying degree of false results.
-Knowing this, to what degree can an investigator trust a result from a work applying such a method?
-Essentially, when the batch/group configuration is balanced, or group difference is ignored  (i.e.
-no group labels not given as parameters to ComBat), problems related to preserving group differences will not occur.
+For an investigator facing an unbalanced data set with batch effects, our advice is to adjust for batch inside the statistical test and avoid the two step procedure outlined above.
+If this is not possible, then applying the method described in this article should only be done with great caution and downstream confidence estimates must be treated with suspicion.
+
+Knowing that adjusting for batch effects while preserving the group difference may lead to varying degree of false results, to what extent can an investigator trust a result from a work applying such a method?
+Essentially, when the batch--group configuration is balanced, or group difference is ignored  (i.e.
+group labels not given as parameters to ComBat), problems related to preserving group differences will not occur.
 For other cases, a re-analysis without using this approach is the most rigorous path.
@@
 In essence asking if the balanced parts (effective sample size?) of the data has enough power to detect the presumed effects and if this is the case (\citealp{Johnson2007}, \CITE{24584070}), treat the results more like an ordered list with the most likely true positives on top while de-emphasizing the somewhat deflated p-values .
-In contrast, if biological knowledge suggest that a group effect is unlikely  (\CITE{24391845}, \CITE{18414638}?, \CITE{21731603}, \CITE{23630272}), an intermediate lack of batch/group balance  could lead to a mostly false result.
+In contrast, if biological knowledge suggest that a group effect is unlikely  (\CITE{24391845}, \CITE{18414638}?, \CITE{21731603}, \CITE{23630272}), an intermediate lack of batch-group balance  could lead to a mostly false result.
 
